### 2015-04-06 manuscript/main.tex
Manuscript skeleton added

\documentclass[fleqn,10pt]{wlpeerj}
\title{Designing surveys for unmarked animal populations when estimation is model-based}

\author[1]{Paul B. Conn}
\author[1]{Erin E. Moreland}
\author[1]{Michael C. Cameron}
\author[1]{Erin Richmond}
\author[1]{Peter L. Boveng}

\affil[1]{National Marine Mammal Laboratory, NOAA-NMFS, Alaska Fisheries Science Center, 7600 Sand Point Way NE, Seattle, WA 98115 USA}

\keywords{animal abundance, bearded seal, line transect survey, model-based estimation, ringed seal, spatial regression, species distribution models, survey design}

\begin{abstract}
1.  Count and occupancy surveys are often used to estimate the density, abundance, and distribution of animal populations. Recently, model-based approaches to analyzing survey data have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures. Model-based analysis often makes use of species-covariate relationships and/or spatially autocorrelated random effects to help predict density or occurrence in unsampled locations.

2.  There is little information currently available to guide ecologists on appropriate survey design protocols for implementing efficient surveys when estimation is model-based.

3. We investigate the interplay between the design of animal surveys and estimator performance (bias, precision) when statistical models are used to predict animal abundance and occurrence. In particular, we use simulation and a case study (designing an aerial survey for seals in the Chukchi Sea) to illustrate the relationship between design and estimator performance.

4. We show that disproportionately  targeting areas of high animal density or occupancy (preferential sampling) can lead to positive bias unless  there are covariates available to help control for variation in sampling intensity.  We also show that performance of a survey design can differ depending on the underlying estimation model.   Models which rely heavily on spatially autocorrelated random effects (e.g. geostatistical models) often perform well when sampling is spatially balanced.  For models that rely heavily on predictive covariates, investigators should disproportionately sample locations with high predictive variance.  For count studies, this will often mean greater sampling in locations where covariates suggest high abundance; for occupancy (presence-absence) surveys, this means greater effort in locations predicted to have occupancy near 0.5.

5. Incorporating some level of spatial balance in survey designs can help minimize integrated prediction error in spatial models and can help produce estimators that are model-robust. Pilot data and/or two-stage sampling can help investigators identify and target additional locations with high prediction variance to further enhance estimator performance.

\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Introduction}

Surveys of unmarked animal populations are often used to estimate abundance and occurrence of animal populations and to predict species distributions, enterprises central to conservation, ecology, and management. For studies of abundance, researchers have historically relied on design-based statistical inference \citep[e.g.][]{Cochran1977}, which requires adoption of a pre-defined sampling frame (e.g. using systematic random sampling, stratified random sampling, or some variant thereof).  Designing animal surveys is relatively straightforward in such applications, and unbiased point and variance estimators are available.  Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture models for repeated point counts \citep{Royle2004a}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,MillerEtAl2013,JohnsonEtAl2010}.  In such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g. via trend surfaces or spatial random effects) to predict density or distributions across the landscape.

Given that model-based inference is now routinely employed in ecology, it is perhaps surprising that there has been relatively little research directing ecologists how to allocate survey effort across space to improve predictive performance.  Notable exceptions include \citet{PeelEtAl2013}, who applied generalized additive models \citep[GAMs;][]{HastieTibshirani1999,Wood2006} to existing trawl survey data to compare alternative designs for reducing the variance of fish relative abundance indices, and \citet{FosterEtAl2014}, who considered alternative designs for autonomous underwater vehicle habitat mapping surveys.  These studies emphasize the importance of spreading sampling out over space (spatial balance) while maintaining a representative range of predictive covariates.
In the context of occupancy studies, \citet{PacificiEtAl2012} described a two-stage sampling design for increasing detection of rare species, showing that increased targeting of locations with high predicted occurrence  leads to occupancy estimators with better precision than simple random sampling.  However, this strategy did not perform as favorably when species were more abundant.

Although applications in ecology are scarce, there has been considerable research on optimal model-based spatial survey design in the statistical literature, for instance, to establish efficient pollution or air quality monitoring networks \citep[e.g.][]{CaseltonZidek1984,Muller2001}. 
Several options exist for optimizing designs, depending upon the type of model employed and whether one is interested in minimizing prediction variance across a landscape or minimizing uncertainty about parameter estimates \citep{Muller2001,DiggleLophaven2006,XiaEtAl2006}. For Gaussian, geostatistical applications, optimal (or near-optimal) designs often appear quite regular, perhaps with additional sample locations chosen close together to help estimate parameters describing spatial autocorrelation \citep{Lark2002,DiggleLophaven2006}.  However, as \citet{EvangelouZhu2012} argue, these criterion for optimality do not necessarily apply to spatial generalized linear mixed models (SGLMMs).  For instance, if animal abundance or occurrence is strongly linked to the value environmental covariates, it may be inefficient to survey as frequently in areas where values of these covariates strongly suggest presence (predicted occupancy near 1.0) or absence (predicted occupancy near 0.0).

If the ultimate goal of animal surveys is to  predict abundance or occurrence across a landscape, the notion of minimizing integrated prediction variance is an appealing design crierion.  However, the optimality of a given survey design is ultimately conditional on the assumed model for the data. An optimal survey design for one model could potentially be quite poor for another model \citep{Muller2001}.  For instance, the optimal design for a given sampling scenario may change depending upon the predictive covariates or functional forms for these covariates (e.g. linear, quadratic) that are included in the model.  The type of spatial basis used to account for spatial autocorrelation may also affect the optimality of the design \citep[e.g. via placement of knots across a landscape,][]{FinleyEtAl2009,GelfandEtAl2012}.  Since ecologists often have considerable uncertainty about the ultimate effect of covariates on animal abundance and distribution, the responsible survey planner would do well to choose a survey design that achieves high precision and low bias across a range of statistical models (i.e. a design that is model-robust).  For this reason, we focus here on developing reasonable strategies for making use of covariate data in designing surveys rather than invoking the advanced optimization (e.g. simulated annealing, exchange algorithms) needed to produce near-optimal designs conditional on a given model.

In addition to choosing designs that result in low integrated prediction variance (high precision), investigators should demand study designs and estimation models that result in unbiased estimators for the state variable of interest.  In a recent paper, \citet{DiggleEtAl2010} examined bias that can arise when preferential sampling is employed but not controlled for with relevant covariates during estimation.  This situation can arise when surveys target areas known to have high abundance or occupancy, but do not include this a priori criterion for sample inclusion in estimation.

In this article, we explore the problem of designing surveys for unmarked animal populations for the purpose of estimating abundance and occurrence under model-based statistical inference.  We start by describing a common currency for notation and basic model structures considered in this paper.  Next, we review preferential sampling bias and how to avoid it, illustrated with a SGLMM applied to simulated count data. Next, we conduct a detailed simulation study to examine application of alternative survey designs on the performance of spatial N-mixture and occupancy models.  Our primary focus here is to determine reasonable allocations between locations selected for spatial balance versus locations selected to reduce predictive variance under two-stage sampling (equivalently, pilot data could also be used in conjunction with one stage sampling).  Finally, we consider the problem of developing an aerial transect survey design for ringed and bearded seals in the Chukchi Sea. The design in this case is complicated by logistical factors such as a low number of airports (2) and effective range of the aircraft.  However, data from historical surveys in the area are available to help simulate the likely effects of different design choices on estimator performance.

\section*{Models and notation}

We focus here exclusively on discrete space models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys.  According to this approach, an investigator who wishes to make inference to a large study area breaks the  landscape up into $\mathcal{S}$ survey units (label these $U_1, U_2, \hdots, U_\mathcal{S}$), of which $S$ are selected for sampling (call the set of sampled locations ${\bf S}$. Each survey unit is assigned a vector of covariates, ${\bf x}_i$, and an indicator $H_i$ that takes on the value 1.0 if survey unit $i$ is sampled, and is 0 otherwise.  Animal abundance or occupancy is then written as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  Z_i & \sim & f(g^{-1}(\beta_0 + {\bf x}\boldsymbol{\beta}+\eta_i+\epsilon_i)). 
\end{eqnarray}
Here, $Z_i$ denotes the state variable of interest (e.g., occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), $\beta_0$ is an intercept parameter, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_K \}$ is a vector of regression parameters, $\eta_i$ is a spatially autocorrelated random effect, and $\epsilon_i$ is Gaussian error.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\eta_i$ include geostatistical models \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected an 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}
\begin{eqnarray}
  Y_i & \sim & \text{Bernoulli}(Z_i p_i),  
\end{eqnarray}
where $p_i$ is possibly a function of survey and observer specific covariates. Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is 
\begin{eqnarray}
  \label{eq:obs_pois}
  Y_i & \sim & \text{Poisson}(Z_i A_i p_i),  
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use standard bracket notation to denote probability mass and density functions.  For instance $[{\bf Z}]$ denotes the marginal probability mass function for ${\bf Z}$, and  $[{\bf Z}|{\bf Y}]$ represents the conditional distribution of ${\bf Z}$ given ${\bf Y}$.

\section*{Preferential sampling}

One of the appealing aspects of model-based estimation is that there is no requirement that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame.  For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location.  This can be a crucial advantage in surveys covering large areas with frequent inclement weather.

However, the manner in which effort is ultimately allocated can potentially have profound influence on estimator performance.  With respect to nonrandom sampling, two possible problems seem particularly likely: coarse scale preferential sampling (CSPS), and fine scale preferential sampling (FSCS) (Fig. \ref{fig:pref}).  FSPS arises when the observations taken at a  particular sampling unit are non-random with respect to the density of animals within that sampling unit.  For instance, when allocating line transect survey effort to a particular sampling unit, it may be tempting to place the transect in a manner that targets habitat or landscape features that maximize the number of animals that will be encountered.  Depending upon the interpretation of occupancy, this may or may not be reasonable.  However, if trying to estimate density or abundance, this strategy will clearly often lead to positive bias.

By contrast, CSPS arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, CSPS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of preferential sampling can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, CSPS arises when we consider the set of sampled locations as stochastic and when $[{\bf S},\boldsymbol{\eta},\boldsymbol{\epsilon}] \neq [{\bf S}][\boldsymbol{\eta},\boldsymbol{\epsilon}]$ \citep{DiggleEtAl2010}.  Bias can then result because spatial regression models require that $E(\boldsymbol{\theta})=E(\boldsymbol{\epsilon})=0$ but cells are targeted specifically because density or occupancy is higher than average.

Bias attributed to CSPS may seem counterintuitive, especially given the maxim in survey sampling to allocate more effort to strata for which animal density is high. For instance, in large scale line transect surveys under stratified sampling, the optimal amount of effort that should be allocated to stratum $s$ is $A_s D_s^{0.5}$, where $A_s$ is the area of $s$ and $D_s$ is the anticipated density \citep[][eqn 7.7]{BucklandEtAl2001}.  Thus, there are theoretical reasons to sample more in high density areas than in low density areas. The obvious solution in this instance is to compensate for CSPS in model-based inferences by accounting for variation in sampling intensity with explanatory covariates or post hoc stratification. However, this does not always work when effort is allocated in a subjective manner.

To illustrate CSPS, we conducted a small count survey simulation experiment.  For each of 100 simulations, we generated abundance of a hypothetical species over a $30 \times 30$ grid as a function of two spatially autocorrelated covariates (with log-linear effects of covariates and an interaction term; Appendix S1).   For each simulated landscape we generated count data using Eqns \ref{eq:process} \& \ref{eq:obs_pois}, setting $A_i = p_i = 1.0$ for demonstration purposes.  We conducted three types of surveys of $K=50$ survey units: (i) a spatially balanced survey, (ii) an unequal probability design where the probabilities of sample inclusion was a function of covariate values (see Appendix S1), and an unequal probability design where the probability of sample inclusion was set to $(Z_i+1)/\sum(Z_i+1)$ (representing CSPS).  We provided a count-based process model with the first of the covariates as an explanatory variable, treating the second as a variable conspiring to influence distribution and abundance and imparting spatial autocorrelation but which remains unknown to the investigator. We used Markov chain Monte Carlo to conduct statistical inference, generating posterior predictions as $N = \sum_i Z_i$ and compared the performance (bias, precision, 90\% credible interval coverage) of these predictions relative to known, true values.

However, the best guard against CSPS is to make decisions on sampling based on the values of explanatory covariates rather than

Chakraborty et al. acknowledge it's an issue for SDMs but don't attempt to model it.

\section*{Example: simulated occupancy and repeated point count surveys}

\section*{Example: planning an aerial Arctic seal survey}

Ecologists at NOAA's National Marine Mammal Laboratory plan to conduct aerial surveys of ringed and bearded seals over the eastern Chukchi Sea in the spring of 2016.  A previous survey \citep{Bengtson2005} indicated that the density of both species was higher in southern portions of the Chukchi Sea and in areas closest to mainland Alaska. Thus, in addition to

These surveys will complement earlier (2012, 2013) surveys of the Bering Sea, and will largely use the same sampling methodology.  In particular, thermal video cameras mounted through the belly port of the aircraft will provide initial detection of seals that are basking on sea ice, and high resolution digital cameras will provide information on species identity \citep{ConnEtAl2014}.  We anticipate using a model-based approach to estimation, following the hierarchical model structure outlined by \citet{ConnEtAl2014} for Bering Sea surveys.  This approach includes a spatial regression model for the distribution of seals among survey units, but the observation model is considerably complicated by imperfect detection (thermal detection misses $\approx 6\%$ of seals), incomplete availability (satellite telemetry records indicate that 40-70\% of seals are basking at the time of surveys), and incomplete coverage of the digital photograph records (20-25\% of thermally detected seals are not photographed).  \citet{ConnEtAl2014} also included species misclassification in their observation model, but updated estimates of species misclassification probabilities \citep{McClintockEtAlInReview} indicate that the probability of misclassifying a bearded seal as a ringed seal and vice versa is negligible.  However, there remains a small probability ($<2\%$) of seal species being recorded as ``unknown'', even if a thermal detection has an associated photograph.

\citet{Bengtson2005} estimated abundance in 12 spatial strata, using a combination of long and short flight tracks to sample more in areas of higher anticipated seal density.  With the goal of treating these estimates as pilot data, we have attempted to reproduce density maps from \citet{Bengtson2005}, in some cases extending strata with low density northward and eastward from their original study area to the U.S. Exclusive Economic Zone boundary (Fig. \ref{fig:Chukchi}). We have

\section*{Discussion}

In the context of polynomial response surface design, \citet{BoxDraper1959} showed that when one wishes to reduce predictive variance, but also has uncertainty about the functional form of the response surface to be estimated, that the ``optimal" strategy in terms of bias and variance combined is often to concentrate on minimizing bias alone.  For this purpose, a first order orthogonal design is required; in spatial applications easting and northing covariates, this would imply some type of systematic sample.  ``. . . conclusions obtained by minimizing variance error only and assuming the postulated model to be correct are likely to be wrong in many practical design situations." (Draper and Smith 1966, pp 18-19).

For instance, one could augment a spatially balanced sampling design with a number of locations which are known or thought to have high leverage (as a function of available covariates, for instance).

A number of authors have explored optimal knot placement in spatial models.  In the context of predictive process modeling \citep[where a covariance function is specified over a group of knots; see][]{BanerjeeEtAl2008}, \citet{FinleyEtAl2009} and \citet{GelfandEtAl2013} considered near-optimal knot placement by minimizing spatially averaged prediction variance.
Gelfand et al. knot selection

Using GAs to select knot placement using

Contrast with posterior loss (e.g., Jay's linex loss function estimator)

Contrast with cross validation

David Miller (the evil one)/Simon Wood stuff on edge effects

Contrast with other approaches-
Gelfand et al. Bayesian analysis - intrinsic CAR in SDM
Chakraborty et al. '10 - spatial abundance modeling - ordinal
Latimer et al. 2009 - spatial predictive process modelling in SPDs

Can't be complacent... still possible to get poor/biased results, e.g. if $\tau_\epsilon \rightarrow 0$.  Can't resolve pathological problems.

Presence-absence data, other link functions (e.g. probit)

extensions to models w/ secondary observation process, measurement error

much attention has been given to collinearity in multiple linear regression - suggest researchers give as much attention to predictive extrapolation bias in predictive models

For predictions with spatial , our experience is that predictions outside the minimum convex polygon where data are obtained can sometimes be more problematic than predictions within the polygon.  Spatial prediction surfaces may have a tendency to bend up or down in these areas, resulting in ``edge effects" that can lead to positive prediction bias when a log link function is employed \citep{VerHoefJansen2007}.

Judicious choices of priors can at times limit the influence of priors on calculation of the gIVH.  For example, if we specify a Zellner's g-prior \citep{Zellner1986} on the vector of regression coefficients, $[\boldsymbol{\beta}] = \textrm{MVN}({\bf 0},(\tau_\beta X^\prime X)^{-1})$, then priors are automatically scaled to the level of the gathered covariates, and the only remaining task is to investigate sensitivity to $\tau_\beta$.
Some care must be taken with prior distributions if Eq. \ref{eq:E_var_lambda} is used for calculation of prediction variance.  For instance, standard ``non-informative" or ``flat" priors may place substantial mass on implausible values.  When using the gIVH or prediction variance to help guide sampling design \citep[see][for an example using prediction variance]{DiggleLophaven2006}, we suggest using informative prior distributions.

Thus, it may prove useful when formulating survey designs since one can examine whether or not prediction points lie within the gIVH without ever collecting response data there.

Results from automated simulation routines must be taken within an appropriate context.  In real life situations, researchers often are privy to a wealth of ecological intuition and knowledge, and would likely disbelieve implausibly high or low estimates of abundance in certain areas and to adjust estimation models accordingly.  Nevertheless, it is important to

\subsection*{Figures and Tables}

Use the table and tabular commands for basic tables --- see Table~\ref{tab:widgets}, for example. You can upload a figure (JPEG, PNG or PDF) using the project menu. To include it in your document, use the includegraphics command as in the code for Figure~\ref{fig:view} below.

\begin{figure}[ht]
\centering
\caption{A depiction of two types of preferential sampling.  In (A), an investigator preferentially places point transects (red squares) within regions of high known density (blue polygons).  This can cause bias in abundance or occupancy estimators unless this a priori knowledge about density is explicitly modeled.  In (B), a fine scale version of preferential sampling occurs when a line transect (red line) is intentionally placed across a region of high quality habitat.  If a landscape is discretized into homogeneous survey units (as in a grid), it is essential that the habitat surveyed within each survey unit be randomly determined.  If not, bias (usually positive) can be expected.}
\includegraphics[width=\linewidth]{Pref_sampling}
\label{fig:pref}
\end{figure}

\begin{figure}[ht]
\centering
\caption{Planned study area for 2016 arctic seal surveys in the eastern Chukchi Sea off of Alaska. The study area is discretized into $625km^2$ grid cells on a Polar Stareophonic projection (gray lines), commensurate with the resolution of available sea ice covariate data.  The boundaries of the study area are determined by the Bering Strait to the south, the U.S. Exclusive Economic Zone to the west and north (yellow and black dashed line), and the $156^{\circ}W$ longitude line to the west (red line).  Also shown are the 10 near shore (yellow shading) and 2 off shore (green shading) employed in previous seal surveys in this area \citep{Bengtson2005}.  Water depths >500m are indicated in purple.}
\includegraphics[width=\linewidth]{LOSI_Planning_2016}
\label{fig:Chukchi}
\end{figure}

\begin{figure}[ht]
\centering
\caption{A collection of 9 hypothetical aerial transect designs for seals over the eastern Chukchi Sea, differing by (i) number of flights (4, 8, or 12; displayed on columns.), and (ii) the spatial distribution of tracks.  The first row includes scenarios where only ``long" tracks are flown, trying to achieve more or less even coverage across the study area, while subsequent rows have an increasingly greater number of short tracks flown near the coast to target areas of higher seal densities.}
\includegraphics[width=\linewidth]{sim_flights_Chukchi}
\label{fig:flights}
\end{figure}

%\begin{table}[ht]
%\centering
%%\begin{tabular}{l|r}
%Item & Quantity \\\hline
%Widgets & 42 \\
%Gadgets & 13
%\end{tabular}
%\caption{\label{tab:widgets}An example %table.}
%\end{table}

\begin{table}[ht]
\label{tab:PS}
\caption{Performance of count-based abundance estimators as a function of survey design and estimation model.  Performance measures include proportional relative bias (``Bias"), root mean squared error (``RMSE"), coefficient of variation (``CV"), and 90\% credible interval coverage (``Cov90''; the proportion of simulations where true abundance was between the 5th and 95th percentiles of posterior samples).  Mean values over 400 simulation replicates are presented for spatially balanced sampling (``Balanced"), inverse probability sampling based on covariate values (``Covariate"), and inverse probability sampling based on a priori knowledge of areas of high abundance (``Preferential"). In addition, two different estimation models were applied to each dataset, including a generalized linear model (``GLM''), and a spatial regression model (``RSR'').} 
\centering
\begin{tabular}{rrlrrrr}
  \hline
 Design & Model & Bias & RMSE & CV & Cov90 \\ 
  \hline
Balanced & GLM & 0.00 & 1038 & 0.069 & 0.92 \\ 
Balanced & RSR & 0.00 & 1034 & 0.057 & 0.90 \\ 
Covariate & GLM & 0.00 & 1392 & 0.067 & 0.88 \\ 
Covariate & RSR & 0.00 & 1152 & 0.061 & 0.88 \\ 
Preferential & GLM & 0.21 & 5807 & 0.060 & 0.24 \\ 
Preferential & RSR & 0.15 & 3040 & 0.054 & 0.29 \\ 
   \hline
\end{tabular}
\end{table}

\section*{Results and Discussion}

\section*{Acknowledgments}

So long and thanks for all the fish.

\bibliography{master_bib}

\end{document}

### 2015-04-06 manuscript/pref_sampling.tex
Manuscript skeleton added

% besdoc.tex V1.0, 17 March 2011

\documentclass[times,mee,doublespace,]{besauth2}
%%\documentclass[times,mee,]{besauth}

\newcommand{\journalnamelc}{British Ecological Society}
\newcommand{\journalabb}{British Ecological Society}
\newcommand{\journalclassshort}{BES}
%%\newcommand{\journalname}{British Ecological Society}
\usepackage{epstopdf}

\usepackage{moreverb}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\usepackage{lineno}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\bibpunct[; ]{(}{)}{;}{}{}{,}

\def\volumeyear{2015}
\def\VOC{VO}

\begin{document}
\runningheads{P.~B.~Conn, D.~S.~Johnson, \& J.~T.~Thorson}{Preferential sampling in species distribution models}

\papertype{Article}

\title{Confronting preferential sampling in count and occupancy surveys: diagnosis and triage \footnotemark[2]}

\author{Paul B. Conn\affil{1}\corrauth, Devin S. Johnson\affil{1}, James T. Thorson\affil{2}}

\address{\affilnum{1}National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; \affilnum{2} Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, National Marine Fisheries Service (NMFS),
NOAA, 2725 Montlake Boulevard E, Seattle, WA 98112, USA}

\corraddr{paul.conn@noaa.gov}

\begin{abstract}
 %\noindent {\emph{Summary}}
\small
\begin{enumerate}
\item  Count and occupancy surveys are often used to estimate the density, abundance, and distribution of animal populations. Recently, model-based approaches to analyzing survey data (i.e. species distribution models) have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures. Model-based analysis often makes use of species-covariate relationships and/or spatially autocorrelated random effects to help predict density or occurrence in unsampled locations.

\item Species distribution models often make the implicit assumption that locations chosen for sampling and animal abundance are conditionally independent given modeled covariates.  However, this assumption is likely violated in many cases when survey effort is non-randomized, leading to preferential sampling.

\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models.  The approach works by jointly modeling animal abundance/occurrence and the locations selected for sampling, and specifying a dependent correlation structure between the two models.

\item Using simulation, we show that our approach reduces bias resulting from non-random, preferential sampling relative to a traditional species distribution model.  Under strong preferential sampling, biases using traditional species distributions models can be considerable (e.g. 20\%).

\item Species case study

\item When animal populations are surveyed using a non-randomized design, we argue that ecologists routinely test and correct for preferential sampling when fitting species distribution models to animal encounter data.

\end{enumerate}

 Word count: XXXX
\end{abstract}

\keywords{count data, preferential sampling, spatial autocorrelation, species distribution model}

\maketitle \linenumbers

\def\VAR{{\rm Var}\,}
\def\COV{{\rm Cov}\,}
\def\Prob{{\rm P}\,}

\section{Introduction}

Surveys of unmarked animal populations are often used to estimate abundance and occurrence of animal populations and to predict species distributions, enterprises central to conservation, ecology, and management. For studies of abundance, researchers historically relied on design-based statistical inference \citep[e.g.][]{Cochran1977}, which requires adoption of a pre-defined sampling frame (e.g. using systematic random sampling, stratified random sampling, or some variant thereof).  Designing animal surveys is relatively straightforward in such applications, and unbiased point and variance estimators are available.  Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture models for repeated point counts \citep{Royle2004a}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,MillerEtAl2013,JohnsonEtAl2010}.  In such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g. via trend surfaces or spatial random effects) to predict density or distributions across the landscape.  We shall refer to the amalgam of model-based approaches for making spatially explicit inference about animal populations as ``Species distribution models" \citep[SDMs; {\it sensu}][]{ElithLeathwick2009}, even though this term is more often used to refer to animal occurrence than it is to density or abundance.

One of the main advantages of using SDMs advanced in the literature is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.  However,
in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling is nonrandom. The potential for biased estimates arises when sampling locations disproportionately target locations where the response of interest is higher (or lower) than the mean response than would be predicted from explanatory covariates with complete knowledge of the system.  In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates. For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.  Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.

In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS), and describe several model-based approaches for detecting and correcting for such biases.  We start by describing a common currency for notation and basic model structures considered in this paper.  Second, we review preferential sampling bias in a mathematical light, and describe prior approaches to coping with its effects. Third, we introduce a novel generalization of previously proposed PS models, allowing the investigator to jointly model animal encounter data and the locations chosen for sampling, including possible dependence structure between these two types of observations.  Fourth, we conduct a simulation study to examine the performance of traditional SDMs and our newly developed PS model when data are gathered preferentially.  Finally, we demonstrate utility of our proposed modeling approach by analyzing a data set of MYSTERY SPECIES X.

\section{Materials and methods}

\subsection{Notation and basic model structures}

We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that preferential sampling is likely to affect analyses similarly regardless of the choice of spatial domain.  We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if survey unit $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
\begin{eqnarray}
  \label{eq:process}
  Z_i & \sim & f(g^{-1}(\mu_i)).
\end{eqnarray}
In this example, $Z_i$ denotes the state variable of interest (e.g., occupancy or abundance), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a linear predictor.  In applications described in this paper, we write the linear predictor as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\eta_i+\epsilon_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, $\eta_i$ is a spatially autocorrelated random effect, and $\epsilon_i$ is Gaussian error.  For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\eta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g. conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.

The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.  However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.  For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}
\begin{eqnarray}
  Y_i & \sim & \text{Bernoulli}(Z_i p_i),
\end{eqnarray}
where $p_i$ is possibly a function of survey and observer specific covariates. Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.   For count surveys, a possible model is
\begin{eqnarray}
  \label{eq:obs_pois}
  Y_i & \sim & \text{Poisson}(Z_i A_i p_i),
\end{eqnarray}
where the $Y_i$ now represents the count of animals obtained while surveying unit $i$, $A_i$ denotes the proportion of sample unit $i$ that is surveyed, and $p_i$ gives detection probability.  Additional information will often be needed to estimate $p_i$ in this context, such as data from double observers, distance observations, or double sampling \citep[see e.g.][]{BucklandEtAl2001,Royle2004,Borchers2006,ConnEtAl2014}.

For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.  We also use standard bracket notation to denote probability mass and density functions.  For instance $[{\bf Z}]$ denotes the marginal probability mass function for ${\bf Z}$, and  $[{\bf Z}|{\bf Y}]$ represents the conditional distribution of ${\bf Z}$ given ${\bf Y}$.

\subsection{Preferential sampling: A primer}

One of the appealing aspects of model-based estimation is that there is no requirement that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame.  For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location.  This can be a crucial advantage in surveys covering large areas with frequent inclement weather.  It also opens the door for using platforms of opportunity, presence only, and citizen science data for estimation.

However, the manner in which effort is ultimately allocated can potentially have profound influence on SDM estimator performance.  With respect to nonrandom sampling, two possible problems seem particularly likely: coarse scale preferential sampling (CSPS), and fine scale preferential sampling (FSPS) (Fig. \ref{fig:pref}).  FSPS arises when the observations taken at a  particular sampling unit are non-random with respect to the density of animals within that sampling unit.  For instance, when allocating line transect survey effort, it may be tempting to place the transect in a manner that targets habitat or landscape features that maximize the number of animals that will be encountered.  Depending upon the interpretation of occupancy, this may or may not be reasonable.  However, if trying to estimate density or abundance, this strategy will clearly lead to positive bias.

By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g. density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.  For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high. \citet{DiggleEtAl2010} showed that this type of preferential sampling can lead to bias when this extra information is not included in models for the state variable of interest.  Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}.  We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.  For instance, \citet{MerckxEtAl2011} use the term ``preferential sampling" to refer to the process of visiting some sites more often than others, while \citet{ManceurKuhn2014} define it as occurring when the locations selected for sampling are a function of an environmental covariate.  Neither of these latter conditions are problematic outside of the specialized field of presence-only modelling.

\citet{DiggleEtAl2010} demonstrated PS with an environmental monitoring problem, whereby pollutant monitoring stations were more highly clustered around urban areas with high concentrations of pollutants than in rural areas with comparably low levels of pollutants.  Fitting simple geostatistical models without fixed effects led to positively biased estimates of landscape-level pollutant concentrations.  Presumably (and as noted by discussants of the article) including a fixed effect associated with a relevant covariate (e.g., an ``urbanity" index) would likely reduce or eliminate bias.  However, the primary point of \citet{DiggleEtAl2010} is well taken: inclusion of spatially autocorrelated random effects in a statistical model is insufficient to remove the potentially biasing effects of PS.

As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS.  However, in many ecological applications, predictive covariates are only able to explain a portion of variation present in the data.  If the locations selected for sampling are related (intentionally or unintentionally) to some unmodelled factor related to abundance, bias may still occur.  Despite the clear potential for bias in SDMs, we have been unable to find many cases where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.  For instance,
\citet{ChakrabortyEtAl2010} acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs, but did not attempt to model it.

Several authors have attempted model-based corrections for PS in the statistical literature. For Gaussian models in a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the locations that are chosen for sampling and the underlying random field of interest.  In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity depended linearly on spatially-referenced random field values.  For instance, writing observations of the spatial random field at a location $i$ as
\begin{eqnarray}
   Z_i & = & \mu_i + \epsilon,
   \label{eq:Z_cont}
\end{eqnarray}
the relative density of sampling locations at $i$ would be written as
\begin{eqnarray}
   p_i & \propto & \exp(\xi_i + b \mu_i).
   \label{eq:p_cont}
\end{eqnarray}
Here, the parameter $b$ describes the level of preferential sampling; $b=0$ implies no preferential sampling, $b>0$ implies a greater level of sampling in locations where the state variable is anomalously high, and $b<0$ implies greater sampling where the state variable is anomalously low.  Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``. . . accounting for informative sampling is only necessary when there
is an association between the spatial surface of interest and the sampling density that cannot be
explained by the shared spatial covariates."  \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in Eq. \ref{eq:Z_cont}, finding that this approach helped reduce bias associated with preferential sampling, but did not perform as well as the full joint model.

\subsection{A generalized preferential sampling model}

The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling preferential sampling.  However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.  Also, they require the linear predictor of the preferential sampling model to be written as a simple linear function of the expected value of the spatial process model.  In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.  For instance, if sampling intensity is high in areas close to bases of operations (e.g., landing strips in the case of aerial surveys), we might expect there to be a greater propensity for PS close to these bases (at least, if animal density is anomolously high or low in these areas).

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the process model, we start with Eq. \ref{eq:process} as a general formulation for non-Gaussian data.  We then expand the link-scale expectation from this model
(i.e. Eq. \ref{eq:mu}) as follows:
\begin{eqnarray}
  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\eta_i+\delta_i+\epsilon_i,
\label{eq:mu_aug}
\end{eqnarray}
where $\delta_i$ is an additional, spatially referenced random effect.  Next, we model the binary indicator for sample inclusion using a Bernoulli distribution:
\begin{eqnarray}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the linear predictor for this model as
\begin{eqnarray}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\gamma_i+\delta_i+b \mu_i + \varepsilon_i.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\gamma_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu_aug} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} need not be the same (although they can be).  Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs. \ref{eq:mu_aug} and \ref{eq:nu}, allowing for dependency in the two models.  Following \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we also include the potential for a linear effect of $\mu_i$ on sampling intensity in Eq. \ref{eq:nu} (with parameter $b$).  However, including both $b \mu_i$ and $\delta_i$ in Eqs. \ref{eq:mu_aug} and \ref{eq:nu} will likely lead to an overparameterized model.  Useful constraints to place on these models are (a) $b=0$ (i.e., spatially varying PS), (b) $\delta_i = 0$ (i.e., a linear effect of the link-scale state process on sampling intensity), and (c) $b=\delta_i=0$ (i.e., no preferential sampling).  A comparison of the performance of models with different sets of constraints can serve as a test of PS.  In particular, if one can demonstrate that models with constraint (c) perform similarly or better than models with (a) and (b) constraints, then PS is likely not worth modeling and inference can proceed using standard SDMs (i.e. not modeling sampling intensity).

\subsection{Simulation study}

To illustrate PS, we conducted a count survey simulation experiment.  For each of 100 simulations, we generated abundance of a hypothetical species over a $30 \times 30$ grid as a function of two spatially autocorrelated covariates (with log-linear effects of covariates and an interaction term; Appendix S1).   For each simulated landscape we generated count data using Eqns \ref{eq:process} \& \ref{eq:obs_pois}, setting $A_i = p_i = 1.0$ for demonstration purposes.  We conducted three types of surveys of $K=50$ survey units: (i) a spatially balanced survey, (ii) an unequal probability design where the probabilities of sample inclusion was a function of covariate values (see Appendix S1), and an unequal probability design where the probability of sample inclusion was set to $(Z_i+1)/\sum(Z_i+1)$ (representing PS).  We provided a count-based process model with the first of the covariates as an explanatory variable, treating the second as a variable conspiring to influence distribution and abundance and imparting spatial autocorrelation but which remains unknown to the investigator. We used Markov chain Monte Carlo to conduct statistical inference, generating posterior predictions as $N = \sum_i Z_i$ and compared the performance (bias, precision, 90\% credible interval coverage) of these predictions relative to known, true values.

\subsection{MYSTERY SPECIES X}

\section{Results}

\section{Discussion}

Bias attributed to PS may seem counterintuitive, especially given the maxim in survey sampling to allocate more effort to strata for which animal density is high. For instance, in large scale line transect surveys under stratified sampling, the optimal amount of effort that should be allocated to stratum $s$ is $A_s D_s^{0.5}$, where $A_s$ is the area of $s$ and $D_s$ is the anticipated density \citep[][eqn 7.7]{BucklandEtAl2001}.  Thus, there are theoretical reasons to sample more in high density areas than in low density areas. The obvious solution in this instance is to compensate for PS in model-based inferences by accounting for variation in sampling intensity with explanatory covariates or post hoc stratification. However, this does not always work when effort is allocated in a subjective manner.

Extension to continuous space - sampling locations as point process similar to Warton and Shepherd (2010)

 \acks{Later.}

\bibliographystyle{bes}
\bibliography{master_bib}

\pagebreak

\begin{table}[ht]
\label{tab:PS}
\caption{Performance of count-based abundance estimators as a function of survey design and estimation model.  Performance measures include proportional relative bias (``Bias"), root mean squared error (``RMSE"), coefficient of variation (``CV"), and 90\% credible interval coverage (``Cov90''; the proportion of simulations where true abundance was between the 5th and 95th percentiles of posterior samples).  Mean values over 400 simulation replicates are presented for spatially balanced sampling (``Balanced"), inverse probability sampling based on covariate values (``Covariate"), and inverse probability sampling based on a priori knowledge of areas of high abundance (``Preferential"). In addition, two different estimation models were applied to each dataset, including a generalized linear model (``GLM''), and a spatial regression model (``RSR'').}
\centering
\begin{tabular}{rrlrrrr}
  \hline
 Design & Model & Bias & RMSE & CV & Cov90 \\
  \hline
Balanced & GLM & 0.00 & 1038 & 0.069 & 0.92 \\
Balanced & RSR & 0.00 & 1034 & 0.057 & 0.90 \\
Covariate & GLM & 0.00 & 1392 & 0.067 & 0.88 \\
Covariate & RSR & 0.00 & 1152 & 0.061 & 0.88 \\
Preferential & GLM & 0.21 & 5807 & 0.060 & 0.24 \\
Preferential & RSR & 0.15 & 3040 & 0.054 & 0.29 \\
   \hline
\end{tabular}
\end{table}

%\begin{figure*}
%\begin{center}
%\includegraphics[width=170mm]{../figures/state_diagram.eps}
%\caption{A depiction of potential inferential obstacles when data from marked animals are %used to address mixed migration strategies. The migratory strategy of an animal will often %be unknown at the time of marking, but observational data may be used to help infer the %proportion of animals subscribing to each strategy, as well as the probability of %migrating for conditional migrants ($\theta_t$).
%behavior} \label{fig:state_diagram}
%\end{center}
%\end{figure*}

\begin{figure*} %[ht]
\begin{center}
\includegraphics[width=170mm]{Pref_sampling_diag.pdf}
\caption{A depiction of two types of preferential sampling.  In (A), an investigator preferentially places point transects (red squares) within regions of high known animal density (blue polygons).  This can cause bias in abundance or occupancy estimators unless this a priori knowledge about density is explicitly modeled.  In (B), a fine scale version of preferential sampling occurs when a line transect (red line) is intentionally placed across a region of high quality habitat.  If a landscape is discretized into homogeneous survey units (as in a grid), it is essential that the habitat surveyed within each survey unit be randomly determined when estimating abundance.  If not, bias (usually positive) can be expected.}
\label{fig:pref}
\end{center}
\end{figure*}

\begin{figure*} %[ht]
\begin{center}
\includegraphics[width=170mm]{PS_maps.pdf}
\caption{An example of a single simulation replicate examining the effect of preferential sampling on estimates of abundance from a species distribution model.  First, two spatially autocorrelated covariates are generated (panels A-B).  Second, true abundance is generated conditional on these covariates (C).  Next, three hypothetical sampling designs are employed to generate animal counts, including a spatially balanced sample (D), a design where sampling is proportional to values of a modeled covariate (E), and where sampling is preferential - i.e. more likely to occur where abundance is high (F).  Finally, spatially explicit estimates of abundance are generated using a traditional SDM to each of the count datasets (G-I).  Although the map produced in (I) does not look particularly bad in this instance, preferential sampling led to estimates of abundance that were biased 20\% high.}
\label{fig:PS_sims}
\end{center}
\end{figure*}

\end{document}

%http://www.plosone.org/article/info:doi/10.1371/journal.pone.0036527

### 2015-06-22 manuscript/pref_sampling.tex
Basic simulation structure in place but getting positive bias from TMB - some code included for contrasting different modeling approaches with simplified structure (e.g. no pref sampling via "spatial_only")

@@
 \begin{document}
-\runningheads{P.~B.~Conn, D.~S.~Johnson, \& J.~T.~Thorson}{Preferential sampling in species distribution models}
+\runningheads{P.~B.~Conn, J.~T.~Thorson, \& D.~S.~Johnson}{Preferential sampling in species distribution models}
 
@@
 
-\title{Confronting preferential sampling in count and occupancy surveys: diagnosis and triage \footnotemark[2]}
+\title{Confronting preferential sampling in count and occupancy surveys: diagnosis and model-based triage \footnotemark[2]}
 

### 2015-06-22 TMB_version/inst/executables/pref_samp_linear.cpp

@@ -1,0 +1,113 @@
+// Space time 
+#include <TMB.hpp>
+template<class Type>
+bool isNA(Type x){
+  return R_IsNA(asDouble(x));
+}
+
+template<class Type>
+Type plogis(Type x){
+  return 1.0 / (1.0 + exp(-x));
+}
+
+template<class Type>
+Type objective_function<Type>::operator() ()
+{
+  // Indices
+  DATA_INTEGER(n_cells);       // Total number of areal survey units in study area
+  DATA_INTEGER(ncol_Xr);       // Total number of observations (i)
+  DATA_INTEGER(ncol_Xy);       // Total number of observations (i)
+
+  // Data
+  DATA_FACTOR( R_i );         // Count for observation
+  DATA_VECTOR( Y_i );       	// Species for observation
+  DATA_VECTOR( Log_area_i );        // A vector giving the (standardized) area of suitable habitat associated with each cell (length = n_cells)
+  DATA_VECTOR( Prop_sampled_i);  // Proportion of suitable habitat in each cell that is sampled (length = n_cells) 
+
+  // Covariates
+  DATA_MATRIX( Xr );  //design matrix for Bernoulli sampling model
+  DATA_MATRIX( Xy );  //design matrix for Poisson count model
+  
+  // Spatial precision matrix (Q where Sigma^{-1} = \tau Q)  Note: not all of these are used
+  DATA_SPARSE_MATRIX(Q1);
+  DATA_SPARSE_MATRIX(Q2);
+  DATA_SPARSE_MATRIX(Q3);
+
+ 
+  // Fixed effects
+  PARAMETER(logtau_Eta);           // log-inverse SD of Eta
+  PARAMETER(logtau_Gamma);        // log-inverse SD of Gamma
+  PARAMETER(beta_pref);    //preferential sampling effect
+  //PARAMETER(logsigma_Epsilon);   
+  PARAMETER_VECTOR(betar);
+  PARAMETER_VECTOR(betay);
+
+  // Random effects
+  PARAMETER_VECTOR(Eta);  // Poisson count spatial random effect
+  PARAMETER_VECTOR(Gamma);   // Bernoulli site selection spatial random effect
[…]

### 2015-06-22 TMB_version/inst/executables/spatial_only.cpp

@@ -1,0 +1,102 @@
+// Space time 
+#include <TMB.hpp>
+template<class Type>
+bool isNA(Type x){
+  return R_IsNA(asDouble(x));
+}
+
+template<class Type>
+Type plogis(Type x){
+  return 1.0 / (1.0 + exp(-x));
+}
+
+template<class Type>
+Type objective_function<Type>::operator() ()
+{
+  // Indices
+  DATA_INTEGER(n_cells);       // Total number of areal survey units in study area
+  DATA_INTEGER(ncol_X);       // Number of covariates
+
+  // Data
+  DATA_VECTOR( Y_i );       	// Counts
+  DATA_VECTOR( Log_area_i );        // A vector giving the (standardized) area of suitable habitat associated with each cell (length = n_cells)
+  DATA_VECTOR( Prop_sampled_i);  // Proportion of suitable habitat in each cell that is sampled (length = n_cells) 
+
+  // Covariates
+  DATA_MATRIX(X);  //design matrix for Poisson count model
+  
+  // Spatial precision matrix (Q where Sigma^{-1} = \tau Q)
+  DATA_SPARSE_MATRIX(Q);
+ 
+  // Fixed effects
+  PARAMETER(logtau_Eta);           // log-inverse SD of Eta
+  PARAMETER(log_rho);
+
+  PARAMETER_VECTOR(Beta);
+
+  // Random effects
+  PARAMETER_VECTOR(Eta);  // Poisson count spatial random effect  
+
+  // global stuff
+  using namespace density; //gives access to multivariate distributions in density.cpp
+  Type jnll = 0;
+  
+  // Derived quantities related to GMRF.  NOTE: No residual white noise assumed
+  Eigen::SparseMatrix<Type> Prec;   //precision matrix for GMRF
+  //matrix<Type> Prec(n_cells,n_cells);
+  //Eigen::SparseMatrix<Type> Q_rho;
+  
+  //Q_rho=Q;
+  Type rho=exp(log_rho);
+  Type tau=exp(logtau_Eta);
+  //
+  //Probability of random fields
+  Prec=Q;
[…]

### 2015-06-22 TMB_version/inst/executables/spatial_only_skaug.cpp

@@ -1,0 +1,76 @@
+// Space time 
+#include <TMB.hpp>
+template<class Type>
+bool isNA(Type x){
+  return R_IsNA(asDouble(x));
+}
+
+template<class Type>
+Type plogis(Type x){
+  return 1.0 / (1.0 + exp(-x));
+}
+
+template<class Type>
+Type objective_function<Type>::operator() ()
+{
+  // Indices
+  DATA_INTEGER(n_cells);       // Total number of areal survey units in study area
+  DATA_INTEGER(ncol_X);       // Number of covariates
+
+  // Data
+  DATA_VECTOR( Y_i );       	// Counts
+  DATA_VECTOR( Log_area_i );        // A vector giving the (standardized) area of suitable habitat associated with each cell (length = n_cells)
+  DATA_VECTOR( Prop_sampled_i);  // Proportion of suitable habitat in each cell that is sampled (length = n_cells) 
+
+  // Covariates
+  DATA_MATRIX(X);  //design matrix for Poisson count model
+  
+  // ICAR structure
+  DATA_SPARSE_MATRIX(Q);
+ 
+  // Fixed effects
+  PARAMETER(logtau_Eta);           // log-inverse SD of Eta
+
+  PARAMETER_VECTOR(Beta);
+
+  // Random effects
+  PARAMETER_VECTOR(Eta);  // random effects  
+
+  // global stuff
+  using namespace density; //gives access to multivariate distributions in density.cpp
+  Type jnll = 0;
+  
+   //
+  //Probability of random fields
+  // ICAR prior
+  Type tau = exp(logtau_Eta);
+  vector<Type> Tmp = Q*Eta;
+  jnll -= Type(0.5)*(n_cells*logtau_Eta-tau*(Eta*Tmp).sum());
+
+  //vector<Type> Eta = Tmp/tau;       // Doing the scaling as a post-step. 100% OK!!
+
+  //prior on tau
+  jnll -= dgamma(tau,Type(1.0),Type(0.01), true ); 
+
+  // Derived quantities
+  vector<Type> Ypred_i(n_cells);
+  Type tot_abund=0;
+  
+  // Probability of observations
+  for(int i=0; i<n_cells; i++){
+    // probability of counts
+    Ypred_i(i) = Log_area_i(i)+Eta(i);
[…]

### 2015-06-22 pref_sampling/R/sim_data_abundance.R

@@ -1,0 +1,192 @@
+#' function to simulate abundance over a simulated landscape
+#' @param S Number of cells (must be a square number)
+#' @param n.covs Number of habitat covariates (default 4)
+#' @param tau.epsilon Precision of exchangeable errors in log space
+#' @return A SpatialPolygonsDataFrame holding covariates, abundance, and lambda (expected abundance)
+#' @export
+#' @keywords spatio-temporal, simulation
+#' @author Paul B. Conn
+sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
+  DEBUG=FALSE   
+  PLOT=FALSE
+  if(PLOT)set.seed(11111)
+  tau.epsilon=100
+  #if(DEBUG)tau.epsilon=100
+  
+  #source('./SpatPred/R/util_funcs.R')
+   
+  if(sqrt(S)%%1>0)cat("error in sim_data_generic; S must be a square number \n")
+  
+  x.len=sqrt(S)
+ 
+  n.basis=10
+  n.zero=6
+  Omega=matrix(0,n.basis,S)
+  Cov=matrix(0,n.covs,S)
+  flag=1
+  while(flag==1){  #reject if any |cor|>0.75
+    for(i in 1:n.basis){
+      my.mod=RMexp(var=1,scale=runif(1,5,100))
+      Omega[i,]=RFsimulate(model=my.mod,x=c(1:x.len),y=c(1:x.len))$variable1
+    }
+    for(i in 1:n.covs){
+      Wts=matrix(rnorm(n.basis),1,n.basis)
+      Wts[sample(c(1:n.basis),n.zero)]=0
+      #cat(Wts)
+      #cat('\n')
+      Cov[i,]=Wts%*%Omega
+    }
+    Covs=t((Cov-rowMeans(Cov))/sqrt(apply(Cov,1,'var')))
+    cur.cor=max(abs(c(cor(Covs[,1],Covs[,2]),cor(Covs[,1],Covs[,3]),cor(Covs[,2],Covs[,3]))))
+    if(cur.cor<0.75)flag=0
+  }
+    
+  
+  if(PLOT==TRUE){
+    #XY=data.frame(x=Dat.matern$x,y=Dat.matern$y)
+    Grid1=data.frame(y=rep(1:30,each=30),x=rep(c(1:30),30),cov=as.vector(Covs[,1]))
+    Grid2=data.frame(y=rep(1:30,each=30),x=rep(c(1:30),30),cov=as.vector(Covs[,2]))
[…]

### 2015-06-22 pref_sampling/R/spat_pred.R

@@ -1,0 +1,314 @@
+#' function to perform Bayesian analysis of log Gaussian Cox processes and to make spatial predictions 
+#' @param formula A formula object specifying the model for the fixed effects
+#' @param Data   A SpatialPolygonsDataFrame holding covariate values for all cells one wants to make predictions on (including sampled cells)
+#' @param Effort A list including "Counts" which gives the response variable for surveyed cells, and "Mapping" which is a vector indicating which sample unit is associated with which count
+#' @param Area.adjust   A vector allowing for differences in suitable habitat for each cell.  Can be used for different grid cell sizes or different
+#'        proportions of suitable habitat (e.g., 1.0 = 100 percent of habitat is suitable for the focal species)
+#' @param Offset	A vector giving the fraction of each grid cell in Mapping that was surveyed
+#' @param spat.mod Controls whether (and what type) of spatial autocorrelation is implemented (spatmod = 0: no spatial autocorrelation; spatmod=1: process convolution; spatmod=2: srr ICAR) 
+#'        Note that if spatmod=1, the argument Knots needs to be specified; if spatmod=2, Adj must be specified
+#' @param Assoc   Association matrix for all grid cells in Data (needed for ICAR models only)
+#' @param K   A (# cells)x(# knots) matrix - the 'K' matrix for process convolution modeling.  
+#' @param Names.gam A character vector giving the column names of Data that will be modeled as smooth effects
+#' @param Knots.gam A list vector (length is # of smooth params), where each list element is a vector providing locations of knots (one location for each knot)
+#' @param Control A list giving MCMC controls and additional model options
+#'  "iter": number of MCMC iterations;
+#'  "burnin": number of MCMC burnin iterations;
[…]

### 2015-06-22 pref_sampling/R/util_funcs.R

@@ -1,0 +1,203 @@
+#' Produce an RW2 structure matrix for a line (e.g. for time series)
+#' @param x length of vector
+#' @return RW2 precision matrix
+#' @export 
+#' @keywords adjacency
+#' @author Paul Conn
+linear_adj_RW2 <- function(x){
+  Q=Matrix(0,x,x)
+  for(i in 1:(x-2)){
+    Q[i,i+2]=1
+    Q[i+2,i]=1
+  }
+  for(i in 1:(x-1)){
+    Q[i,i+1]=-4
+    Q[i+1,i]=-4
+  }
+  Q[1,2]=-2
+  Q[2,1]=-2
+  Q[x,x-1]=-2
+  Q[x-1,x]=-2
+  diag(Q)=-rowSums(Q)
+  Q
+}  
+
+#' Produce an RW1 adjacency matrix for a rectangular grid for use with areal spatial models (queens move)
+#' @param x number of cells on horizontal side of grid
+#' @param y number of cells on vertical side of grid
+#' @param byrow If TRUE, cell indices are filled along rows (default is FALSE)
+#' @return adjacency matrix
+#' @export 
+#' @keywords adjacency
+#' @author Paul Conn \email{paul.conn@@noaa.gov}
+rect_adj <- function(x,y,byrow=FALSE){
+  Ind=matrix(c(1:(x*y)),y,x,byrow=byrow)
+  if(byrow==TRUE)Ind=t(Ind)
+  n.row=nrow(Ind)
+  n.col=ncol(Ind)
+  Adj=matrix(0,x*y,x*y)
+  for(i in 1:n.row){
+    for(j in 1:n.col){
+      if(i==1 & j==1){
+        Adj[Ind[i,j],Ind[i,j]+1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row+1]=1
+      }
+      if(i==1 & j>1 & j<n.col){
+        Adj[Ind[i,j],Ind[i,j]+1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row]=1
+        Adj[Ind[i,j],Ind[i,j]-n.row]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row+1]=1
+        Adj[Ind[i,j],Ind[i,j]-n.row+1]=1
+      }
+      if(i==1 & j==n.col){
[…]

### 2015-07-22 manuscript/pref_sampling.tex
Bias correction for TMB now implemented correctly for random effects, but simple spatial regression models still point to a bias in posterior predictions (likely associated with fixed effect).  Plan now is to switch to MCMC via Gibbs sampling.
Updated MS to make model stucture for preferential sampling similar to Royle and Berliner (also now better reflects model in Pati et al.).  This will hopefully resolve some identifiability issues I ran into in TMB simulations.

@@
 However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.
-Also, they require the linear predictor of the preferential sampling model to be written as a simple linear function of the expected value of the spatial process model.
+Also, they require the linear predictor of the preferential sampling model to be written as a simple linear function of the the spatial process model for density.
 In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.
-For instance, if sampling intensity is high in areas close to bases of operations (e.g., landing strips in the case of aerial surveys), we might expect there to be a greater propensity for PS close to these bases (at least, if animal density is anomolously high or low in these areas).
+For instance, sampling may be more strongly preferential close to bases of operations, (e.g., landing strips in the case of aerial surveys), but less so in areas that are harder to get to.
 
@@
 \begin{eqnarray}
-  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\eta_i+\delta_i+\epsilon_i,
+  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i+\epsilon_i,
 \label{eq:mu_aug}
 \end{eqnarray}
-where $\delta_i$ is an additional, spatially referenced random effect.  Next, we model the binary indicator for sample inclusion using a Bernoulli distribution:
+where $\delta_i$ is a spatially referenced random effect.  Next, we model the binary indicator for sample inclusion using a Bernoulli distribution:
 \begin{eqnarray}
@@
 \begin{eqnarray}
-  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\gamma_i+\delta_i+b \mu_i + \varepsilon_i.
+  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i + \varepsilon_i.
 \label{eq:nu}
 \end{eqnarray}
- In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\gamma_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.
+ In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\eta_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.
 The predictive covariates ${\bf x}_i$ from Eq.
@@
 Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs.
-\ref{eq:mu_aug} and \ref{eq:nu}, allowing for dependency in the two models.
-Following \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we also include the potential for a linear effect of $\mu_i$ on sampling intensity in Eq.
-\ref{eq:nu} (with parameter $b$).
-However, including both $b \mu_i$ and $\delta_i$ in Eqs.
-\ref{eq:mu_aug} and \ref{eq:nu} will likely lead to an overparameterized model.
-Useful constraints to place on these models are (a) $b=0$ (i.e., spatially varying PS), (b) $\delta_i = 0$ (i.e., a linear effect of the link-scale state process on sampling intensity), and (c) $b=\delta_i=0$ (i.e., no preferential sampling).
-A comparison of the performance of models with different sets of constraints can serve as a test of PS.
-In particular, if one can demonstrate that models with constraint (c) perform similarly or better than models with (a) and (b) constraints, then PS is likely not worth modeling and inference can proceed using standard SDMs (i.e.
+\ref{eq:mu_aug} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.
+
+ The formulation in Eq.
+\ref{eq:nu} is the same as previously proposed by other authors for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.
+There are multiple ways of structuring ${\bf B}$ depending on the complexity with which one wants spatial dependence that is desired \citep{RoyleBerliner1999}.
+For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no preferential sampling).
+Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear preferential sampling model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.
+Alternatively, we could allow the degree of PS to vary across the landscape.
+For instance, one can contemplate a trend surface model for preferential sampling by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
+Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surfaces or a two dimensional spline \citep{RoyleBerliner1999,UnpublishedManuscript}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.
+
+A comparison of the performance of models with different sets of constraints on ${\bf B}$ can serve as a test of PS.
+In particular, if one can demonstrate that models with ${\bf B}={\bf 0}$ perform similarly or better than models with ${\bf B} \ne {\bf 0}$, then PS is likely not worth modeling and inference can proceed using standard SDMs (i.e.
 not modeling sampling intensity).

### 2015-07-22 TMB_version/inst/executables/pref_samp_v3.cpp

@@ -30,13 +30,11 @@ Type objective_function<Type>::operator() ()
   
   // ICAR structure
   DATA_SPARSE_MATRIX(Q);
-
-
  
   // Fixed effects
   PARAMETER(logtau_Eta);           // log-inverse SD of Eta
   PARAMETER(logtau_Gamma);        // log-inverse SD of Gamma
-  PARAMETER(beta_pref);    //preferential sampling effect
+  PARAMETER(logtau_Delta);        // log-inverse SD of Delta
   //PARAMETER(logsigma_Epsilon);   
   PARAMETER_VECTOR(betar);
   PARAMETER_VECTOR(betay);
@@ -44,6 +42,7 @@ Type objective_function<Type>::operator() ()
   // Random effects
   PARAMETER_VECTOR(Eta);  // Poisson count spatial random effect
   PARAMETER_VECTOR(Gamma);   // Bernoulli site selection spatial random effect
+  PARAMETER_VECTOR(Delta);  // shared spatial random effect
   //PARAMETER_VECTOR(Epsilon);  //residual error random effect on linear predictor for count model
   
   
@@ -52,8 +51,8 @@ Type objective_function<Type>::operator() ()
   using namespace density; //gives access to multivariate distributions in density.cpp
   Type jnll = 0;
   
-  // Derived quantities related to GMRF.  NOTE: No residual white noise assumed
-
+  
+  //random effects priors
   Type tau_eta = exp(logtau_Eta);
   vector<Type> Tmp = Q*Eta;
   jnll -= Type(0.5)*(n_cells*logtau_Eta-tau_eta*(Eta*Tmp).sum());
@@ -62,27 +61,14 @@ Type objective_function<Type>::operator() ()
   Tmp = Q*Gamma;
   jnll -= Type(0.5)*(n_cells*logtau_Gamma-tau_gamma*(Gamma*Tmp).sum());
   
+  Type tau_delta = exp(logtau_Delta);
+  Tmp = Q*Delta;
+  jnll -= Type(0.5)*(n_cells*logtau_Delta-tau_delta*(Delta*Tmp).sum());
   
   //prior on tau
[…]

### 2015-07-22 pref_sampling/inst/run_TMB_sims.R

@@ -141,7 +141,7 @@ for(isamp in 1:2){  #sampling intensity
           Xr = model.matrix(~1,data=Grid@data)
           Xy = model.matrix(~cov1+cov2+cov1.quad+cov2.quad+cov12,data=Grid@data) 
         }
-        Data=list("n_cells"=S,"n_transects"=length(Effort$Counts),"ncol_Xr"=ncol(Xr), "ncol_Xy"=ncol(Xy), "R_i"=R, "Y_i"=Y, "Log_area_i"=log(Area.adjust),"Prop_sampled_i"=Offset,"Xr"=Xr, "Xy"=Xy, "Q1"=Q, "Q2"=Q,"Q3"=Q)
+        Data=list("n_cells"=S,"n_transects"=length(Effort$Counts),"ncol_Xr"=ncol(Xr), "ncol_Xy"=ncol(Xy), "R_i"=R, "Y_i"=Y, "Log_area_i"=log(Area.adjust),"Prop_sampled_i"=Offset,"Xr"=Xr, "Xy"=Xy, "Q"=Q)
         
         for(iest in c(1,2,4,5)){
           counter=counter+1
@@ -212,23 +212,24 @@ for(isamp in 1:2){  #sampling intensity
             opt[["AIC"]] = 2*opt$objective + 2*length(opt$par)
             #opt[["BIC"]] = 2*opt$objective + length(opt$par) * log(n.transects)
             
-            # Diagnostics
+             # Diagnostics
             na=0
             if(length(Random)>0){
               Report=sdreport(obj,bias.correct=TRUE)
               Random=summary(Report,"random")
               if(sum(is.na(Random[,2]))>0)na=1
+              n.hat=Report$unbiased$value
             }
             else{
               Report=sdreport(obj)
+              n.hat=Report$value
             }
             Fixed=summary(Report,"fixed")
             
-             if(sum(is.na(Fixed[,2]))>0)na=1
+            if(sum(is.na(Fixed[,2]))>0)na=1
             
             Report2 = obj$report()
             
-            n.hat=Report$value
             n.se=Report$sd
             n.true=sum(Grid[["N"]])
             

### 2015-07-22 pref_sampling/inst/run_simple_spatial.R

@@ -19,7 +19,10 @@ library( TMB )
 library( INLA )
 
 COMPILE=FALSE
-SIM=TRUE
+SIM=T
+prop.sampled=0.05
+n.transects=90
+
 
 if(COMPILE){
   Tmb.dir="./TMB_version/inst/executables/"
@@ -29,8 +32,8 @@ if(COMPILE){
   compile(TmbFile2 )
 }
 
-set.seed(123456)
-n.sims=1 #number of simulations at each design point
+set.seed(223456)
+n.sims=100 #number of simulations at each design point
 S=900
 Adj=rect_adj(sqrt(S),sqrt(S))
 Q=-Adj
@@ -38,8 +41,8 @@ diag(Q)=apply(Adj,2,'sum')+0.01  #the +1 is just for debugging to make it proper
 Q=Matrix(Q)
 
 #initialize Results data frame
-Result=matrix(0,n.sims*8*10,10)  #number of simulations * number of generating models * number of estimation models
-colnames(Result)=c("Gen.mod","Samp.intens","Covs","Est.mod","Sim.no","N.est","N.true","SE","MARE","NA.fixed")
+Result=matrix(0,n.sims,11)  #number of simulations * number of generating models * number of estimation models
+colnames(Result)=c("Gen.mod","Samp.intens","Covs","Est.mod","Sim.no","N.est","N.true","SE","MARE","NA.fixed","N.est2")
 
 
 if(SIM){   #true abundance is independent of data generating model
@@ -57,9 +60,6 @@ if(SIM){   #true abundance is independent of data generating model
 }
 
 if(SIM){
-  prop.sampled=1.0
-  n.transects=250
-     
   for(isim in 1:n.sims){
       cat('isim ',isim,'\n')
       Cur.file=paste("./Sim_data/Cov_abundance",isim,".Rda",sep='')
@@ -71,151 +71,160 @@ if(SIM){
 }
 
 set.seed(423456)
-counter=0
-Cur.file=paste("./Sim_data/Effort_bal_high",isim,".Rda",sep='')
-load(Cur.file) #load Effort 
- 
-      
-#reformat data to consist of R (binary sampled/not sampled) and Y (count)
-Y=Offset=rep(NA,S)
-Y[Effort$Mapping]=Effort$Counts
[…]

### 2015-09-30 TMB_version/inst/executables/spatial_only_skaug.cpp
Changes to simple spatial model in TMB (got rid of intercept), a few ms changes

@@ -31,7 +31,7 @@ Type objective_function<Type>::operator() ()
   // Fixed effects
   PARAMETER(logtau_Eta);           // log-inverse SD of Eta
 
-  PARAMETER_VECTOR(Beta);
+  //PARAMETER_VECTOR(Beta); //(run without since ICAR rank deficient by 1)
 
   // Random effects
   PARAMETER_VECTOR(Eta);  // random effects  
@@ -45,7 +45,7 @@ Type objective_function<Type>::operator() ()
   // ICAR prior
   Type tau = exp(logtau_Eta);
   vector<Type> Tmp = Q*Eta;
-  jnll -= Type(0.5)*(n_cells*logtau_Eta-tau*(Eta*Tmp).sum());
+  jnll -= Type(0.5)*((n_cells-1)*logtau_Eta-tau*(Eta*Tmp).sum());
 
   //vector<Type> Eta = Tmp/tau;       // Doing the scaling as a post-step. 100% OK!!
 
@@ -60,7 +60,7 @@ Type objective_function<Type>::operator() ()
   for(int i=0; i<n_cells; i++){
     // probability of counts
     Ypred_i(i) = Log_area_i(i)+Eta(i);
-    for(int c=0; c<ncol_X; c++) Ypred_i(i) += X(i,c) * Beta(c) ;
+    //for(int c=0; c<ncol_X; c++) Ypred_i(i) += X(i,c) * Beta(c) ; //run without intercept temporarily
     Ypred_i(i) = exp(Ypred_i(i));
     if( !isNA(Y_i(i)) ) jnll -= dpois( Y_i(i), Prop_sampled_i(i)*Ypred_i(i), true );
     tot_abund += Ypred_i(i);

### 2015-09-30 pref_sampling/inst/run_simple_spatial.R

@@ -20,7 +20,7 @@ library( INLA )
 
 COMPILE=FALSE
 SIM=T
-prop.sampled=0.05
+prop.sampled=0.1
 n.transects=90
 
 
@@ -72,6 +72,7 @@ if(SIM){
 
 set.seed(423456)
 for(isim in 1:n.sims){
+  cat(paste("isim",isim,"\n"))
   counter=0
   Cur.file=paste("./Sim_data/Effort_bal_high",isim,".Rda",sep='')
   load(Cur.file) #load Effort 
@@ -90,7 +91,6 @@ for(isim in 1:n.sims){
   
   
   #### Unconditional spec
-  
   TmbFile="./TMB_version/inst/executables/spatial_only"
   dyn.load(dynlib(TmbFile))
   
@@ -142,13 +142,15 @@ for(isim in 1:n.sims){
   
   
   ###########  alternative formulation using Hodges & Reich Eqn 3
+  diag(Data$Q)=diag(Data$Q)-0.01
   
   TmbFile="./TMB_version/inst/executables/spatial_only_skaug"
   dyn.load(dynlib(TmbFile))
   
   Map = list()  #specify fixed parameter values
   
-  Params = list("logtau_Eta"=log(1),"Beta"=0,"Eta"=rep(0,S))
+  #Params = list("logtau_Eta"=log(1),"Beta"=0,"Eta"=rep(0,S))
+  Params=list("logtau_Eta"=log(1),"Eta"=rep(0,S))
   Random=c("Eta")  #specify which params are random effects
   #Map[["beta_pref"]] = factor( NA )
   
@@ -210,7 +212,7 @@ for(isim in 1:n.sims){
   n.inla.mean=sum(Out.inla$summary.fitted[,"mean"])
   n.inla.median=sum(Out.inla$summary.fitted[,"0.5quant"])
   
-  Result[isim,]=c(imod,isamp,icov,iest,isim,n.hat2,n.true,n.se,sum(abs(Report2$Ypred_i-Grid[["N"]]))/n.true,na,n.inla.mean)
+  Result[isim,]=c(1,1,1,1,isim,n.hat2,n.true,n.se2,sum(abs(Report2$Ypred_i-Grid[["N"]]))/n.true,na,n.inla.mean)
   
 }
 prop.sampled

### 2015-09-30 TMB_version/inst/executables/pref_samp_v1.cpp
added back in a few .cpp files I accidentally untracked; removed all .o and .dll's

@@ -1,0 +1,105 @@
+// Space time 
+#include <TMB.hpp>
+template<class Type>
+bool isNA(Type x){
+  return R_IsNA(asDouble(x));
+}
+
+template<class Type>
+Type plogis(Type x){
+  return 1.0 / (1.0 + exp(-x));
+}
+
+template<class Type>
+Type objective_function<Type>::operator() ()
+{
+  // Indices
+  DATA_INTEGER(n_stations);       // Total number of observations (i)
+  DATA_INTEGER(n_knots);       // Total number of observations (i)
+  DATA_INTEGER(ncol_Xr);       // Total number of observations (i)
+  DATA_INTEGER(ncol_Xy);       // Total number of observations (i)
+
+  // Data
+  DATA_FACTOR( R_i );         // Count for observation
+  DATA_VECTOR( Y_i );       	// Species for observation
+
+  // Covariates
+  DATA_MATRIX( Xr );
+  DATA_MATRIX( Xy );
+  
+  // SPDE objects
+  DATA_SPARSE_MATRIX(G0);
+  DATA_SPARSE_MATRIX(G1);
+  DATA_SPARSE_MATRIX(G2);
+
+  // Fixed effects
+  PARAMETER(log_kappa);         // Controls range of spatial variation
+  PARAMETER(logtau_Nu);        // log-inverse SD of Epsilon
+  PARAMETER(logtau_Gamma);        // log-inverse SD of Epsilon
+  PARAMETER(logtau_Delta);        // log-inverse SD of Epsilon
+  PARAMETER_VECTOR(betar);
+  PARAMETER_VECTOR(betay);
+
+  // Random effects
+  PARAMETER_VECTOR(Nu_input);  // Spatial process variation
+  PARAMETER_VECTOR(Gamma_input);   // Spatial variation in carrying capacity
+  PARAMETER_VECTOR(Delta_input);
+
+  // global stuff
+  using namespace density;
+  Type jnll = 0;
+  
+  // Derived quantities related to GMRF
+  Type kappa2 = exp(2.0*log_kappa);
+  Type kappa4 = kappa2*kappa2;
+  Type pi = 3.141592;
+  Type Range = sqrt(8) / exp( log_kappa );
[…]

### 2015-09-30 TMB_version/examples/rect_adj.R
adding my initial unbiased-demonstration code

@@ -1,0 +1,75 @@
+
+#' Produce an RW1 adjacency matrix for a rectangular grid for use with areal spatial models (queens move)
+#' @param x number of cells on horizontal side of grid
+#' @param y number of cells on vertical side of grid
+#' @param byrow If TRUE, cell indices are filled along rows (default is FALSE)
+#' @return adjacency matrix
+#' @export
+#' @keywords adjacency
+#' @author Paul Conn \email{paul.conn@@noaa.gov}
+rect_adj <- function(x,y,byrow=FALSE){
+  Ind=matrix(c(1:(x*y)),y,x,byrow=byrow)
+  if(byrow==TRUE)Ind=t(Ind)
+  n.row=nrow(Ind)
+  n.col=ncol(Ind)
+  Adj=matrix(0,x*y,x*y)
+  for(i in 1:n.row){
+    for(j in 1:n.col){
+      if(i==1 & j==1){
+        Adj[Ind[i,j],Ind[i,j]+1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row+1]=1
+      }
+      if(i==1 & j>1 & j<n.col){
+        Adj[Ind[i,j],Ind[i,j]+1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row]=1
+        Adj[Ind[i,j],Ind[i,j]-n.row]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row+1]=1
+        Adj[Ind[i,j],Ind[i,j]-n.row+1]=1
+      }
+      if(i==1 & j==n.col){
+        Adj[Ind[i,j],Ind[i,j]+1]=1
+        Adj[Ind[i,j],Ind[i,j]-n.row]=1
+        Adj[Ind[i,j],Ind[i,j]-n.row+1]=1
+      }
+      if(i>1 & i<n.row & j==1){
+        Adj[Ind[i,j],Ind[i,j]+1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row]=1
+        Adj[Ind[i,j],Ind[i,j]-1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row-1]=1
+        Adj[Ind[i,j],Ind[i,j]+n.row+1]=1
[…]

### 2015-09-30 TMB_version/examples/simple_example_v2.R

@@ -1,0 +1,134 @@
+
+# Z -- true abundance
+# Y -- sampled abundance
+
+library( RandomFields )
+library( TMB )
+library( INLA )
+
+Version = "simple_v2"
+# v2 -- added option for changing distribution for random effects, and implemented ICAR distribution
+
+# Compile
+TmbFile = "C:/Users/James.Thorson/Desktop/Project_git/pref_sampling/TMB_version/inst/executables"
+setwd( TmbFile )
+compile( paste0(Version,".cpp") )
+
+# SOurce
+source( paste0(TmbFile,"/../../examples/rect_adj.R") )
+source( paste0(TmbFile,"/../../examples/rrw.R") )
+
+# Settings
+grid_dim = c("x"=30, "y"=30)
+n_samp = 50
+SpatialScale = sqrt(prod(grid_dim))/5  # Range ~ 2*Scale
+SD_Nu = 1
+betay = 2
+Use_REML = TRUE
+Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
+Spatial_model_set = c("SPDE_GMRF", "ICAR")[1:2]
+
+# Source ICAR precision matrix function
+Q = rect_adj( x=grid_dim['x'], y=grid_dim['y'] )
+Q = -1*Q
+diag(Q) = -1 * colSums(Q)
+# Sigma = solve(Q)   # Appears to be singular
+
+# Calculate C (pg. 185 Cressie and Wikle)
+#C = rect_adj( x=grid_dim['x'], y=grid_dim['y'] )
+
+# Results
+Results = array(NA, dim=c(length(Spatial_model_set),500,5), dimnames=list(Spatial_model_set,NULL,c("betay","Sigma_Nu","Range_Nu","sum_Ztrue","sum_Zpred")) )
+
+# Loop through replicates
+for(i in 1:dim(Results)[2]){
+  # Spatial model
+  loc_s = expand.grid( "x"=1:grid_dim['x'], "y"=1:grid_dim['y'])
+  model_Nu <- RMgauss(var=SD_Nu^2, scale=SpatialScale)
+
+  # Realization from GRF
+  if( Spatial_sim_model=="GP_gaussian") Nu = RFsimulate(model=model_Nu, x=loc_s[,'x'], y=loc_s[,'y'])@data[,1]
+  if( Spatial_sim_model=="ICAR") Nu = rrw( Q )[,1]
+
+  # Linear predictors
[…]

### 2015-09-30 TMB_version/inst/executables/simple_v1.cpp

@@ -1,0 +1,75 @@
+// Space time 
+#include <TMB.hpp>
+template<class Type>
+bool isNA(Type x){
+  return R_IsNA(asDouble(x));
+}
+
+template<class Type>
+Type plogis(Type x){
+  return 1.0 / (1.0 + exp(-x));
+}
+
+template<class Type>
+Type objective_function<Type>::operator() ()
+{
+  using namespace R_inla;
+  using namespace Eigen;
+  using namespace density;
+  
+  // Data
+  DATA_VECTOR( c_i );       	// Response (count) for each observation
+  DATA_FACTOR( s_i ); // Site for each sample
+  DATA_MATRIX( X_sj );  //design matrix for Bernoulli sampling model
+  
+  // Aniso objects
+  DATA_STRUCT(spde,spde_t);
+ 
+  // Parameters 
+  PARAMETER_VECTOR(beta_j);  
+  PARAMETER(logtau);      
+  PARAMETER(logkappa)
+  
+  // Random effects
+  PARAMETER_VECTOR( nuinput_s );
+  
+  // derived sizes
+  int n_i = c_i.size();
+  int n_j = X_sj.row(0).size();
+  int n_s = X_sj.col(0).size();
+  int n_k = nuinput_s.size();
+  
+  // global stuff
+  Type jnll = 0;
+  Type MargSD = 1 / sqrt(4*M_PI) / exp(logtau) / exp(logkappa);
+  Type Range = sqrt(8) / exp( logkappa );
+    
+  //random effects priors
+  SparseMatrix<Type> Q = Q_spde(spde, exp(logkappa));
+  jnll += GMRF(Q)(nuinput_s);
+  
+  // Predicted densities
+  vector<Type> Z_s(n_s);
+  for(int s=0; s<n_s; s++){
+    Z_s(s) = exp( ( X_sj.row(s).array() * beta_j ).sum() + nuinput_s(s)/exp(logtau) );
+  }
+  Type total_abundance = Z_s.sum();
+  
+  // Probability of counts
+  for(int i=0; i<n_i; i++){
+    if(!isNA(c_i(i))) jnll -= dpois( c_i(i), Z_s(s_i(i)), true );
+  }
+
+  // Reporting
+  REPORT( Z_s );
+  REPORT( total_abundance );
+  REPORT( Range );
+  REPORT( MargSD );
+  REPORT( beta_j );
+  
+  // Bias correction output
+  ADREPORT( beta_j );
+  ADREPORT( total_abundance);
+
+  return jnll;
+}

### 2015-10-01 TMB_version/examples/simple_example_v3.R
added V3 with preferential sampling option

@@ -1,0 +1,164 @@
+
+# Z -- true abundance
+# Y -- sampled abundance
+# R -- predicted probability of sampling
+# delta -- affects density
+# eta -- affects sampling probability
+
+library( RandomFields )
+library( TMB )
+library( INLA )
+
+Version = "simple_v3"
+# v2 -- added option for changing distribution for random effects, and implemented ICAR distribution
+# v3 -- added Bernoulli model for sampling locations, and removed the ICAR functionality (by commenting it out, to save speed during building the object)
+
+# Compile
+TmbFile = "C:/Users/James.Thorson/Desktop/Project_git/pref_sampling/TMB_version/inst/executables"
+setwd( TmbFile )
+compile( paste0(Version,".cpp") )
+
+# SOurce
+source( paste0(TmbFile,"/../../examples/rect_adj.R") )
+source( paste0(TmbFile,"/../../examples/rrw.R") )
+
+# Settings
+grid_dim = c("x"=30, "y"=30)
+n_samp = 50
+SpatialScale = sqrt(prod(grid_dim))/5  # Range ~ 2*Scale
+SD_eta = SD_x = SD_delta = 1
+beta0 = 2
+betax = 1        # Impact of x on density
+betax_prob = 1   # Impact of x on sampling intensity
+b = 0            # Impact of delta on sampling intensity
+Use_REML = TRUE
+Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
+Spatial_model = c("SPDE_GMRF", "ICAR")[1]
+
+# Source ICAR precision matrix function
+Q = rect_adj( x=grid_dim['x'], y=grid_dim['y'] )
+Q = -1*Q
+diag(Q) = -1 * colSums(Q)
+# Sigma = solve(Q)   # Appears to be singular
+
+# Calculate C (pg. 185 Cressie and Wikle)
+#C = rect_adj( x=grid_dim['x'], y=grid_dim['y'] )
+
+# Results
+Results = array(NA, dim=c(500,3), dimnames=list(NULL,c("b","sum_Ztrue","sum_Zpred")) )
+
+# Loop through replicates
+for(i in 1:dim(Results)[1]){
+ 
+  # Spatial model
[…]

### 2015-11-20 manuscript/pref_sampling.tex
Changes to simple spatial model in TMB (got rid of intercept), a few ms changes
Draft simulations conducted and written up in ms

@@
 
-\author{Paul B. Conn\affil{1}\corrauth, Devin S. Johnson\affil{1}, James T. Thorson\affil{2}}
+\author{Paul B. Conn\affil{1}\corrauth, James T. Thorson\affil{2}, Devin S. Johnson\affil{1}}
 
@@
 in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling is nonrandom.
-The potential for biased estimates arises when sampling locations disproportionately target locations where the response of interest is higher (or lower) than the mean response than would be predicted from explanatory covariates with complete knowledge of the system.
+The potential for biased estimates arises when sampling locations disproportionately target locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.
 In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates.
@@
 Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.
+
+Need to acknowledge that "sample selection bias" has been addressed extensively in SDM modeling with presence-only data (e.g.
+adjusting "background" or "availability" data to be similar to the locations chosen for sampling).
 
@@
 \citet{DiggleEtAl2010} showed that this type of preferential sampling can lead to bias when this extra information is not included in models for the state variable of interest.
-Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}.
+Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}, where \textbf{R} is an indicator vector taking on a value of $R_i = 1.0$ if sampling unit $i$ is sampled and is zero otherwise.
 We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.
@@
 Despite the clear potential for bias in SDMs, we have been unable to find many cases where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.
-For instance,
-\citet{ChakrabortyEtAl2010} acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs, but did not attempt to model it.
+One such example is
+\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs, but did not attempt to model it.
 
@@
 \begin{eqnarray}
-  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i+\epsilon_i,
+  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
 \label{eq:mu_aug}
@@
 \begin{eqnarray}
+ \label{eq:R}
   R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
@@
 \begin{eqnarray}
-  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i + \varepsilon_i.
+  \label{eq:nu}
+  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
 \label{eq:nu}
@@
 \ref{eq:nu} is the same as previously proposed by other authors for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.
-There are multiple ways of structuring ${\bf B}$ depending on the complexity with which one wants spatial dependence that is desired \citep{RoyleBerliner1999}.
+There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the preferential sampling process \citep{RoyleBerliner1999}.
 For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no preferential sampling).
@@
 For instance, one can contemplate a trend surface model for preferential sampling by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
-Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surfaces or a two dimensional spline \citep{RoyleBerliner1999,UnpublishedManuscript}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.
+Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surface or spline formulation \citep{RoyleBerliner1999}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.
 
@@
 
-To illustrate PS, we conducted a count survey simulation experiment.
-For each of 100 simulations, we generated abundance of a hypothetical species over a $30 \times 30$ grid as a function of two spatially autocorrelated covariates (with log-linear effects of covariates and an interaction term; Appendix S1).
-For each simulated landscape we generated count data using Eqns \ref{eq:process} \& \ref{eq:obs_pois}, setting $A_i = p_i = 1.0$ for demonstration purposes.
-We conducted three types of surveys of $K=50$ survey units: (i) a spatially balanced survey, (ii) an unequal probability design where the probabilities of sample inclusion was a function of covariate values (see Appendix S1), and an unequal probability design where the probability of sample inclusion was set to $(Z_i+1)/\sum(Z_i+1)$ (representing PS).
-We provided a count-based process model with the first of the covariates as an explanatory variable, treating the second as a variable conspiring to influence distribution and abundance and imparting spatial autocorrelation but which remains unknown to the investigator.
-We used Markov chain Monte Carlo to conduct statistical inference, generating posterior predictions as $N = \sum_i Z_i$ and compared the performance (bias, precision, 90\% credible interval coverage) of these predictions relative to known, true values.
+To illustrate PS and demonstrate that our proposed model has reasonable performance, we conducted a small simulation experiment.
+For each of 100 simulations, we generated abundance of a hypothetical species over a $25 \times 25$ grid as
+\begin{eqnarray*}
+  N_i \sim \textrm{Poisson}(\mu_i),
+\end{eqnarray*}
+where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to Eq.
+\ref{eq:mu_aug}.
+Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig.
+\ref{PS_sims}).
+Specific details of of data generation procedures are presented in Appendix S1.
 
-\subsection{MYSTERY SPECIES X}
+For each simulated landscape we generated three virtual count surveys using eqs.
+\ref{eq:R} and \ref{eq:nu}.
+Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, not covariate of spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.
+In the first, we set ${\bf B}=0$, so that the survey was a simple random sample.
+For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=5$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.
+We refer to these scenarios as moderate and pathological preferential sampling, respectively (see fig.
+\ref{fig:sim_R}).
+Simulations were configured so that $n=50$ of the 625 survey units were sampled; each survey was set to cover half of the target cell.
+
+We fitted two different models to each count dataset, both of which were provided the habitat covariate used (in part) to generate the data for which a log-linear coefficient $\beta$ was estimated.
+In the first estimation model, the elements of \textbf{B} in eq.
+\ref{eq:nu} were all set to zero.
+In this case, the abundance and sampling process submodels were independent, as is the case canonical SDMs (at lest when fitted to presence-absence or count data).
+In the second estimation model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix.
+
+We used maximum a posteriori (MAP) estimation to conduct statistical inference.
+This approach is Bayesian; prior parameter distributions are specified and inference is conducted relative to marginal posterior distributions.
+However, in contrast to Markov chain Monte Carlo (MCMC) where the posterior is sampled via simulation, the joint posterior is maximized numerically in this case.
+We used Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}, interfaced with the R programming environment, to conduct maximization.
+The TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{Tierney1989} to obtain abundance estimates that properly account for uncertainty attributable to random effects (\textit{correct, Jim?}).
+This approach allowed for a facile implementation and fast computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with MCMC.
+In this study, we report the results of 500 simulation replicates.
+Further detail on statistical methods are provided in Appendix S1; requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.
+
+\subsection{Bearded seal aerial surveys}
+
+\subsection{Avian point counts}
 
 \section{Results}
+
+\subsection{Simulation study}
+
+Estimates of cumulative animal abundance across simulated landscapes were median unbiased for both estimation methods when the sites selected for sampling were independent of animal density, though when $b$ was estimated abundance estimates were more right skewed and had higher variance (fig.
+\ref{fig:sim_bias}).
+Under moderate preferential sampling ($b=1$), the jointly modeling the density and the sampling process (i.e., estimating $b$) led to a median bias of 5\%, while the canonical SDM model had a median bias of 40\%.
+Under pathological preferential sampling ($b=5$), both estimation methods were extremely biased, but was even more severe for the naive model ignoring preferential sampling (fig.
+\ref{sim_bias}).
 
@@
 Extension to continuous space - sampling locations as point process similar to Warton and Shepherd (2010)
+
+Presumably, trends less biased when pref sampling ignored.
 
@@
 
+\pagebreak
+
 %\begin{figure*}
@@
 \begin{center}
-\includegraphics[width=170mm]{PS_maps.pdf}
-\caption{An example of a single simulation replicate examining the effect of preferential sampling on estimates of abundance from a species distribution model.
-First, two spatially autocorrelated covariates are generated (panels A-B).
-Second, true abundance is generated conditional on these covariates (C).
-Next, three hypothetical sampling designs are employed to generate animal counts, including a spatially balanced sample (D), a design where sampling is proportional to values of a modeled covariate (E), and where sampling is preferential - i.e.
-more likely to occur where abundance is high (F).
-Finally, spatially explicit estimates of abundance are generated using a traditional SDM to each of the count datasets (G-I).
-Although the map produced in (I) does not look particularly bad in this instance, preferential sampling led to estimates of abundance that were biased 20\% high.}
+\includegraphics[width=170mm]{Pref_samp_sim_maps.pdf}
+\caption{An example of a single simulation replicate examining estimates of abundance from a naive species distribution model under preferential sampling.
+First, true abundance (C) is generated as a function of a spatially autocorrelated covariate (A) and a spatially autocorrelated random effect (B).
+Second, counts are generated for three different types of surveys, including a simple random sample ($b=0$; D) and surveys with moderate ($b=1$; E) or pathological ($b=5$; F) levels of preferential sampling.
+Finally, spatially explicit estimates of abundance are generated using a traditional SDM (with $b$ set to 0.0) to each of the count datasets (G-I).
+In this particular simulation replicate, cumulative abundance was underestimated by 18\% when $b=0$, overestimated by 17\% when $b=1$, and overestimated by 293\% when $b=5$.
+For a summary of bias over 500 simulation replicates, see fig.
+\ref{fig:sim_bias}.}
 \label{fig:PS_sims}
+\end{center}
+\end{figure*}
+
+\begin{figure*} %[ht]
+\begin{center}
+\includegraphics[width=170mm]{sim_R.pdf}
+\caption{Expected relationship between the probability of a survey unit being selected for sampling and its abundance residual in the simulation study.
+The base case $b=0$ represents simple random sampling, while $b=1$ and $b=5$ represent moderate and pathological levels of preferential sampling, respectively.
+Also shown are is the realized distribution (smoothed histogram) of abundance residuals among survey units in the simulation study, scaled to fit in the plot margins (solid black line).}
+\label{fig:sim_R}
+\end{center}
+\end{figure*}
+
+\begin{figure*} %[ht]
+\begin{center}
+\includegraphics[width=170mm]{bias.pdf}
+\caption{Relative proportional error in abundance from the simulation experiment as computed with respect the posterior mode with a bias correction.
+Each boxplot summarizes the distribution of relative proportional error as a function of the type of sampling, including simple random sampling ($b=0$), moderate preferential sampling ($b=1$), and pathological preferential sampling ($b=5$).
+Results vary by the type of estimation model; in the ``independent" model, $b$ is set to 0.0; in the ``joint" model, $b$ is estimated.
+Lower and upper limits of each box correspond to first and third quartiles, while whiskers extend to the lowest and highest observed bias within 1.5 interquartile range units from the box.
+Points denote outliers outside of this range.
+Horizontal lines within boxes denote median bias.
+}
+\label{fig:sim_bias}
 \end{center}

### 2015-11-20 TMB_version/examples/simple_example_v2.R

@@ -24,15 +24,14 @@ n_samp = 50
 SpatialScale = sqrt(prod(grid_dim))/5  # Range ~ 2*Scale
 SD_Nu = 1
 betay = 2
-<<<<<<< HEAD
+
 Use_REML = FALSE
 Spatial_model_set = c("SPDE_GMRF", "ICAR")[2]
 Spatial_sim_model = c("GP_gaussian", "ICAR")[2]
-=======
+
 Use_REML = TRUE
 Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
 Spatial_model_set = c("SPDE_GMRF", "ICAR")[1:2]
->>>>>>> 878ed302f089be797086cd504d8ff21ef2397795
 
 # Source ICAR precision matrix function
 Q = rect_adj( x=grid_dim['x'], y=grid_dim['y'] )

### 2015-11-20 TMB_version/examples/simple_example_v5.R

@@ -10,10 +10,11 @@ library( TMB )
 library( INLA )
 #library( ThorsonUtilities )
 
-Version = "simple_v4"
+Version = "simple_v5"
 # v2 -- added option for changing distribution for random effects, and implemented ICAR distribution
 # v3 -- added Bernoulli model for sampling locations, and removed the ICAR functionality (by commenting it out, to save speed during building the object)
 # v4 -- added simpler less-smooth SPDE option (may be faster!), and option to turn off delta or eta
+# v5 -- Implements Devin's finite population correction approach to posterior prediction; proportion area sampled additional input vector  
 
 # Compile
 TmbFile = "C:/Users/paul.conn/git/pref_sampling/TMB_version/inst/executables"
@@ -25,15 +26,18 @@ source( paste0(TmbFile,"/../../examples/rect_adj.R") )
 source( paste0(TmbFile,"/../../examples/rrw.R") )
 
 # Settings
-grid_dim = c("x"=30, "y"=30)
+grid_dim = c("x"=25, "y"=25)
 n_samp = 50
+n_cells = grid_dim[1]*grid_dim[2]
+prop_sampled=0.5
+Prop_sampled=rep(prop_sampled,n_samp)
 SpatialScale = sqrt(prod(grid_dim))/5  # Range ~ 2*Scale
 SD_eta = SD_x = SD_delta = 1
 beta0 = 2
 Use_REML = FALSE   # 
 Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
 Spatial_model = c("SPDE_GMRF", "ICAR")[1]
-Alpha = 1  # Smothness for GMRF, 1 or 2 (1 is faster)
+Alpha = 1  # Smoothness for GMRF, 1 or 2 (1 is faster)
 RandomSeed = ceiling(runif(1,min=1,max=1e6))
 n_sim = 500
 
@@ -61,8 +65,11 @@ for(EstI in 1:length(EM_set)){
   EM = EM_set[EstI]
   set.seed( RandomSeed + i )
   
+  #note: betax set to zero in last commit by Jim (v4)
   betax = runif(1,-.5,.5)       # Impact of x on density
[…]

### 2015-11-20 TMB_version/inst/executables/simple_v5.cpp

@@ -1,0 +1,254 @@
+// Space time 
+#include <TMB.hpp>
+//#include <atomic_math.hpp>  //for D_lgamma
+
+
+/** Precision matrix for the anisotropic case, eqn (20) in Lindgren et al. (2011) */    
+namespace R_inla_generalized {
+using namespace Eigen;
+using namespace tmbutils;
+using namespace R_inla;
+
+template<class Type>
+  SparseMatrix<Type> Q_spde_generalized(spde_t<Type> spde, Type kappa, int alpha=2){
+  Type kappa_pow2 = kappa*kappa;
+  Type kappa_pow4 = kappa_pow2*kappa_pow2;
+  	
+  if( alpha==1 ) return kappa_pow2*spde.M0 + spde.M1;
+  if( alpha==2 ) return kappa_pow4*spde.M0 + Type(2.0)*kappa_pow2*spde.M1 + spde.M2;
+}
+
+template<class Type>
+  SparseMatrix<Type> Q_spde_generalized(spde_aniso_t<Type> spde, Type kappa, matrix<Type> H, int alpha=2){
+
+  int i;
+  Type kappa_pow2 = kappa*kappa;
+  Type kappa_pow4 = kappa_pow2*kappa_pow2;
+  
+  int n_s = spde.n_s;
+  int n_tri = spde.n_tri;
+  vector<Type> Tri_Area = spde.Tri_Area;
+  matrix<Type> E0 = spde.E0;
+  matrix<Type> E1 = spde.E1;
+  matrix<Type> E2 = spde.E2;
+  matrix<int> TV = spde.TV;
+  SparseMatrix<Type> G0 = spde.G0;
+  SparseMatrix<Type> G0_inv = spde.G0_inv;
+	  	  
+  //Type H_trace = H(0,0)+H(1,1);
+  //Type H_det = H(0,0)*H(1,1)-H(0,1)*H(1,0);
+  SparseMatrix<Type> G1_aniso(n_s,n_s); 
+  SparseMatrix<Type> G2_aniso(n_s,n_s); 
+  // Calculate adjugate of H
+  matrix<Type> adj_H(2,2);
+  adj_H(0,0) = H(1,1);
+  adj_H(0,1) = -1 * H(0,1);
+  adj_H(1,0) = -1 * H(1,0);
+  adj_H(1,1) = H(0,0);
[…]

### 2015-11-20 TMB_version/inst/executables/simple_v6.cpp

@@ -1,0 +1,256 @@
+// Space time 
+#include <TMB.hpp>
+//#include <atomic_math.hpp>  //for D_lgamma
+
+
+/** Precision matrix for the anisotropic case, eqn (20) in Lindgren et al. (2011) */    
+namespace R_inla_generalized {
+using namespace Eigen;
+using namespace tmbutils;
+using namespace R_inla;
+
+template<class Type>
+  SparseMatrix<Type> Q_spde_generalized(spde_t<Type> spde, Type kappa, int alpha=2){
+  Type kappa_pow2 = kappa*kappa;
+  Type kappa_pow4 = kappa_pow2*kappa_pow2;
+  	
+  if( alpha==1 ) return kappa_pow2*spde.M0 + spde.M1;
+  if( alpha==2 ) return kappa_pow4*spde.M0 + Type(2.0)*kappa_pow2*spde.M1 + spde.M2;
+}
+
+template<class Type>
+  SparseMatrix<Type> Q_spde_generalized(spde_aniso_t<Type> spde, Type kappa, matrix<Type> H, int alpha=2){
+
+  int i;
+  Type kappa_pow2 = kappa*kappa;
+  Type kappa_pow4 = kappa_pow2*kappa_pow2;
+  
+  int n_s = spde.n_s;
+  int n_tri = spde.n_tri;
+  vector<Type> Tri_Area = spde.Tri_Area;
+  matrix<Type> E0 = spde.E0;
+  matrix<Type> E1 = spde.E1;
+  matrix<Type> E2 = spde.E2;
+  matrix<int> TV = spde.TV;
+  SparseMatrix<Type> G0 = spde.G0;
+  SparseMatrix<Type> G0_inv = spde.G0_inv;
+	  	  
+  //Type H_trace = H(0,0)+H(1,1);
+  //Type H_det = H(0,0)*H(1,1)-H(0,1)*H(1,0);
+  SparseMatrix<Type> G1_aniso(n_s,n_s); 
+  SparseMatrix<Type> G2_aniso(n_s,n_s); 
+  // Calculate adjugate of H
+  matrix<Type> adj_H(2,2);
+  adj_H(0,0) = H(1,1);
+  adj_H(0,1) = -1 * H(0,1);
+  adj_H(1,0) = -1 * H(1,0);
+  adj_H(1,1) = H(0,0);
[…]

### 2015-11-20 pref_sampling/inst/plot_sim_b.R

@@ -1,0 +1,103 @@
+#### plot expectations of probability of sampling vs. abundance residuals
+
+
+# Z -- true abundance
+# Y -- sampled abundance
+# R -- predicted probability of sampling
+# delta -- affects density
+# eta -- affects sampling probability
+
+library( RandomFields )
+library( INLA )
+
+# change directory
+TmbDir = "C:/Users/paul.conn/git/pref_sampling/TMB_version/inst/executables"
+setwd( TmbDir )
+
+# SOurce
+source( paste0(TmbDir,"/../../examples/rect_adj.R") )
+source( paste0(TmbDir,"/../../examples/rrw.R") )
+
+# Settings
+grid_dim = c("x"=25, "y"=25)
+n_samp = 50
+n_cells = grid_dim[1]*grid_dim[2]
+SpatialScale = sqrt(prod(grid_dim))/5  # Range ~ 2*Scale
+SD_eta = SD_x = SD_delta = 1
+beta0 = 2
+Alpha = 1  # Smoothness for GMRF, 1 or 2 (1 is faster)
+n_sim = 500
+
+# Configurations
+b_set = c(0,1,5)            # Impact of delta on sampling intensity
+Data=matrix(0,4,n_sim*n_cells)
+
+# Loop through replicates
+set.seed(12345)
+irow=1
+for(i in 1:n_sim){
+  #simulate abundance
+  # Settings
+
+  #note: betax set to zero in last commit by Jim (v4)
+  betax = runif(1,-.5,.5)       # Impact of x on density
+  betax_prob = runif(1,-.5,.5)   # Impact of x on sampling intensity
+  #betax=0
+  #betax_prob=0
+  
+  # Spatial model
+  loc_s = expand.grid( "x"=1:grid_dim['x'], "y"=1:grid_dim['y'])
+  model_delta <- RMgauss(var=SD_delta^2, scale=SpatialScale)
+  model_x <- RMgauss(var=SD_x^2, scale=SpatialScale)
+  model_eta <- RMgauss(var=SD_eta^2, scale=SpatialScale)
+  
+  # Realization from GRF
+  # delta_s -- spatial variation that affects both density and sampling intensity
+  # x_s -- spatial variation in 
[…]

### 2015-11-20 pref_sampling/inst/plot_sim_example.R

@@ -1,0 +1,66 @@
+### plot preferential sampling plots using data from "run_sim_example.R"
+library(sp)
+library(maptools)
+library(rgeos)
+library(ggplot2)
+library(RColorBrewer)
+library(gridExtra)
+library(grid)
+
+setwd('c:/users/paul.conn/git/pref_sampling')
+source('../SpatPred/SpatPred/R/util_funcs.R')
+load('./Sim_data/sim_plot_data.Rdata')
+
+N.true=matrix(Sim$N.true,ncol=1)
+N.1=matrix(Sim$N.1,ncol=1)
+N.2=matrix(Sim$N.2,ncol=1)
+N.3=matrix(Sim$N.3,ncol=1)
+C.1=matrix(Sim$C.1,ncol=1)
+C.2=matrix(Sim$C.2,ncol=1)
+C.3=matrix(Sim$C.3,ncol=1)
+Delta=matrix(Sim$Delta,ncol=1)
+Cov=matrix(Sim$Cov,ncol=1)
+
+greenPalette <- colorRampPalette(brewer.pal(9, "Greens"))
+bluePalette <- colorRampPalette(brewer.pal(9, "Blues"))
+YlOrBr<- colorRampPalette(brewer.pal(9, "YlOrBr"))
+txt.size=8
+
+
+#turn sim data into SpatialPolygonsDataFrame for use with plot_N_map
+Grid.topo=GridTopology(c(0,0),c(1,1),c(25,25))
+Grid.SpG=SpatialGrid(Grid.topo)
+Grid.SpP=as(as(Grid.SpG,"SpatialPixels"),"SpatialPolygons")
+laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
+                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
+proj4string(Grid.SpP)=CRS(laea_180_proj)   
+Grid=SpatialPolygonsDataFrame(Grid.SpP,data=data.frame(Cov=Cov),match.ID=FALSE)
+
+
+
+Grid.list=vector("list",1)
+Grid.list[[1]]=Grid
+max.N=100
[…]

### 2015-11-20 pref_sampling/inst/run_sim_example.R

@@ -1,0 +1,209 @@
+### run simple preferential sampling example.  Doing in this in two steps since there seems to be
+# a conflict between TMB loop and rgeos ('gCentroid') functionality
+# need to declare an ADREPORT for Z_s in TMB for this
+setwd('c:/users/paul.conn/git/pref_sampling')
+source('../SpatPred/SpatPred/R/util_funcs.R')
+library(sp)
+library(maptools)
+library( RandomFields )
+library( TMB )
+library( INLA )
+#library( ThorsonUtilities )
+
+Version = "simple_v6"
+# v2 -- added option for changing distribution for random effects, and implemented ICAR distribution
+# v3 -- added Bernoulli model for sampling locations, and removed the ICAR functionality (by commenting it out, to save speed during building the object)
+# v4 -- added simpler less-smooth SPDE option (may be faster!), and option to turn off delta or eta
+# v5 -- Implements Devin's finite population correction approach to posterior prediction; proportion area sampled additional input vector  
+# v6 -- Includes option to output Z_s in ADREPORT so unbiased map estimates can be made using bias.correct
+
+# Compile
+TmbFile = "C:/Users/paul.conn/git/pref_sampling/TMB_version/inst/executables"
+setwd( TmbFile )
+compile( paste0(Version,".cpp") )
+
+# SOurce
+source( paste0(TmbFile,"/../../examples/rect_adj.R") )
+source( paste0(TmbFile,"/../../examples/rrw.R") )
+
+# Settings
+grid_dim = c("x"=25, "y"=25)
+n_samp = 50
+n_cells = grid_dim[1]*grid_dim[2]
+prop_sampled=0.5
+Prop_sampled=rep(prop_sampled,n_samp)
+SpatialScale = sqrt(prod(grid_dim))/5  # Range ~ 2*Scale
+SD_eta = SD_x = SD_delta = 1
+beta0 = 2
+Use_REML = FALSE   # 
+Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
+Spatial_model = c("SPDE_GMRF", "ICAR")[1]
[…]

### 2015-12-01 manuscript/AppendixS1.tex
A few minor changes to ms following Thorson initial comments; Appendix S1 describing model and simulation study in more detail added

% besdoc.tex V1.0, 17 March 2011

\documentclass[times,mee,doublespace,]{besauth2}
%%\documentclass[times,mee,]{besauth}

\newcommand{\journalnamelc}{British Ecological Society}
\newcommand{\journalabb}{British Ecological Society}
\newcommand{\journalclassshort}{BES}
%%\newcommand{\journalname}{British Ecological Society}
\usepackage{epstopdf,comment}

\usepackage{moreverb}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\usepackage{lineno}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\bibpunct[; ]{(}{)}{;}{}{}{,}

\def\volumeyear{2015}
\def\VOC{VO}

\begin{document}

\begin{center} \bf {\large Confronting preferential sampling in count and occupancy surveys: diagnosis and model-based triage}

\vspace{0.7cm}
Paul B. Conn$^{1,*}$, James T. Thorson$^2$, Devin S. Johnson$^1$
\end{center}
\vspace{0.5cm}

\rm
\small

$^1$ National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; $^2$ Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, National Marine Fisheries Service (NMFS),
NOAA, 2725 Montlake Boulevard E, Seattle, WA 98112, USA

$^*$paul.conn@noaa.gov

\large
\bigskip
\centerline{Appendix S1: Further details on preferential sampling model development and simulations}
\bigskip
\small

\linenumbers

\def\VAR{{\rm Var}\,}
\def\COV{{\rm Cov}\,}
\def\Prob{{\rm P}\,}

\begin{flushleft}

\raggedbottom

In this appendix, we provide further details on model development, implementation, and simulation testing for the preferential sampling model outlined in the main article text. Our modeling approach works by constructing (a) a species distribution model (SDM) describing abundance or presence/absence that includes spatially autocorrelated random effects, and (b) a model for the probability of site selection.  Both submodels can be expressed in terms of habitat or landscape covariates.  However, site selection probability can also be written as a function of the same spatially autocorrelated random effects in (a).  Including the same set of random effects in both submodels is the mechanism for allowing dependence betwen the two models, and allows us to estimate and control for the potentially biasing effects of preferential sampling.

\subsubsection*{Preferential sampling model development}

We focus here on a discrete spatial domain; extensions to point process models in continuous space are relatively straightforward.  In particular, we suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$). 
Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if survey unit $i$ is sampled (i.e. if $U_i \in \mathcal{S}$), and is 0 otherwise.  Let $Z_i$ denote the state variable of interest in cell $i$ (e.g., occupancy, abundance), and let $Y_i$ denote observations made in cell $i$.  In some cases, observations may be equivalent to the state variable in sampled cells (i.e. when detection probability is 1.0).  In others, an observation model $[Y_i | Z_i, \boldsymbol{\theta}]$ relating $Y_i$ to the underlying process of interest will be required, where $\boldsymbol{\theta}$ are nuisance parameters describing detection.  For instance, for survey counts in our simulation study and in the bearded seal example, we set 
\begin{eqnarray*}
  [Y_i | Z_i, \boldsymbol{\theta}] & = & \textrm{Binomial}(Z_i, A_i p_i),
\end{eqnarray*}
where $A_i$ gives the proportion of survey unit $i$ that is sampled and $p_i$ is a detection probability ($p_i$ is set to 1.0 for the simulation study and an informative prior distribution is specified in the bearded seal example).

\hspace{0.5in} We express variation in $Z_i$ generically as
\begin{eqnarray}
  \label{eq:process}
  Z_i & \sim & f(g^{-1}(\mu_i)),
\end{eqnarray}
where $f()$ is a probability mass function (e.g., Poisson for count data, Bernoulli for occupancy), $g()$ is a link function (e.g. probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value.  In applications described in this paper, we specify this intensity as
\begin{eqnarray}
\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\lambda_i,
 \label{eq:mu}
\end{eqnarray}
where $\beta_0$ is an intercept, ${\bf x}_i$ is a vector of covariates used to explain variation in abundance (here indexed by the site, $i$), and $\boldsymbol{\lambda}=\{ \lambda_1,\lambda_2,\hdots,\lambda_S \}$ are spatially autocorrelated random effects.

\hspace{0.5in} We model the site selection probabilities in the same manner.  First, we write $R_i$, as
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
\label{eq:nu}
\end{eqnarray}
Here, $\beta_0^*$ is an intercept, ${\bf x}_i^*$ is a vector of covariates (in practice, these can be the same or different from ${\bf x}_i$), and $\boldsymbol{\eta}=\{ \eta_1,\eta_2,\hdots,\eta_S \}$ are spatially autocorrelated random effects.  The connection between the two models (eqs. \ref{eq:mu} and \ref{eq:nu}) is the reliance of both on $\delta_i$.  In eq. \ref{eq:nu}, the matrix ${\bf B}$ describes this relationship.  Previous preferential sampling models in the statistical literature have set $\textbf{B} = b \textbf{I}$, where $\textbf{I}$ is an identity matrix and $b$ is a parameter to be estimated.  This is the approach we take in simulation study and in several of the models fit to real data.  Another approach is to construct $\textbf{B}$ so it changes smoothly in space, as in \citep{RoyleBerliner1999}.  In this manner, the degree of preferential sampling could vary across the landscape; presumably, one would need fairly rich data to be able to model a spatially varying relationship.

\hspace{0.5in} We have yet to specify a model for spatial autocorrelation.  For analyses in this paper, we model mean-zero spatial random effects (i.e., the $\lambda_i$ and $\eta_i$) for each site $i$ as multivariate normal:
\begin{eqnarray*}
  \boldsymbol{\lambda} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\lambda}), \text{ and} \\
  \boldsymbol{\eta} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\eta}),
\end{eqnarray*}
where $\boldsymbol{\Sigma}_{\lambda}$ and $\boldsymbol{\Sigma}_{\eta}$ are covariance matrices.  As is typical of spatial models in statistics, we allow the magnitude of spatial autocorrelation to decrease as a function of the distance between survey units.  In particular, we use the well known Mat\'{e}rn covariance function to model covariance as a function of distance.  In addition to the distances between survey units (which we calculate relative to survey unit centroids), the Mat\'{e}rn covariance function uses a range (or scale) parameter $\kappa$, a precision parameter $\tau$ (inverse marginal variance), and a smoothness parameter to model covariance.  For analyses in this paper, we set the smoothness parameter to 1.0 and treat $\kappa$ and $\tau$ as parameters to be estimated.  To improve computational efficiency, we use a Gaussian Markov random field (GMRF) representation of the underlying Gaussian field \citep{LindgrenEtAl2011}, as described by \citet{ThorsonEtAl2015} (Appendix B).

To conduct statistical analysis of the preceding models, we use maximum marginal likelihood as implemented in Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}. Interfaced with the R programming environment, the TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.  Requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.

\subsubsection*{Simulation study details}

For each of 500 simulations, we performed the following steps:

\begin{enumerate}
  \item Generate a spatially autocorrelated covariate over a $25 \times 25$ grid, where the covariate at grid cell $i$, $X_i$, is generated via a multivariate normal distribution:
      \begin{eqnarray}
        \textbf{X} & \sim & \textrm{MVN} (\textbf{0},\boldsymbol{\Sigma}).
        \label{eq:cov}
      \end{eqnarray}
      An isotropic, Gaussian covariance structure is used for $\boldsymbol{\Sigma}$, where the $i$th row and $j$th column of $\boldsymbol{\Sigma}$ is set to $\Sigma_{ij}=\sigma^2 \exp{-(d_{ij}/s)^2}$, where $d_{ij}$ is the distance between the centroid of grid cell $i$ and grid cell $j$, $\sigma$ is a standard deviation (here, set to 1.0), and $s$ is a scale parameter (here, set to 5.0).
  \item Generate true abundance in each survey unit (grid cell), $N_i$, as a function of the covariate from step 1, together with additional spatially autocorrelated error.  We generate a vector of such abundances as
      \begin{eqnarray}
        N_i & \sim & \textrm{Poisson}(\exp(\beta_0 + X_i \beta_1 + \delta_i)),
      \end{eqnarray}
      where the $\boldsymbol{\delta}$ are drawn in an identical manner to $\textbf{X}$ in step 1.  We set $\beta_0 = 2.0$ and drew $\beta_1 \sim \textrm{Uniform}(-0.5,0.5)$ for each simulation.
  \item For each simulation, we selected 50 grid cells for sampling (i.e., set $R_i = 1.0$) according to 3 alternative survey configurations.  Each survey configuration, grid cells were drawn randomly without replacement, where the probability of sampling cell $i$ is proportional to
      \begin{equation*}
        \exp(X_i \beta_1^*  + \eta_i + b*\delta_i).
      \end{equation*}      
      Here, $\eta_i$ is a spatially autocorrelated random effect drawn in an identical manner to $\textbf{X}$ in step 1, and $b$ describes the level of preferential sampling.  Three levels are used: $b=0$, which makes the sampling model conditionally independent from the abundance process (i.e., no preferential sampling); $b=1$, which induces some dependence between the abundance and sampling process (i.e., moderate preferential sampling); $b=5$ which makes cells of high abundance much more likely to be sampled than low abundance cells (i.e., pathological preferential sampling).   We drew $\beta_1^* \sim \textrm{Uniform}(-0.5,0.5)$ for each simulation replicate.  An illustration of the consequences of different survey configurations is provided in Fig. 2 of the main manuscript.
  \item For each of the 3 sampling configurations, we generated a count dataset.  In each case, we generated a count $Y_i$ for each cell selected for sampling (i.e., for cells where $R_i=1.0$) according to 
      \begin{eqnarray*}
        Y_i & \sim & \textrm{Binomial}(N_i,A).
      \end{eqnarray*}
      Here, $A$ represents the portion of the grid cell that is surveyed, and was set to 0.5 for all simulations and grid cells.
  \item For each count data set, we fitted 2 different estimation models in TMB (see \textit{Preferential sampling model development}).  In the first, the preferential sampling parameter $b$ was fixed to 0.0 (thus in effect assuming no preferential sampling).  In the second, $b$ was estimated.
  \item For each combination of simulation replicate, survey configuration, and estimation model, we recorded relative bias with respect to total estimated abundance, $\hat{N} = \sum \hat{N_i}$, where a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
\end{enumerate}

\bibliographystyle{bes}
\bibliography{master_bib}

\pagebreak

\end{flushleft}
\end{document}

%http://www.plosone.org/article/info:doi/10.1371/journal.pone.0036527

### 2015-12-01 manuscript/pref_sampling.tex
A few minor changes to ms following Thorson initial comments; Appendix S1 describing model and simulation study in more detail added

@@
 %%\newcommand{\journalname}{British Ecological Society}
-\usepackage{epstopdf}
+\usepackage{epstopdf,comment}
 
@@
 
-\item Using simulation, we show that our approach reduces bias resulting from non-random, preferential sampling relative to a traditional species distribution model.
-Under strong preferential sampling, biases using traditional species distributions models can be considerable (e.g.
-20\%).
+\item Using simulation, we show that under moderate levels of preferential sampling, our modeling approach reduces bias from $\approx$40\% (under a naive species distribution model) to $approx$5\%.
 
@@
 
-\item When animal populations are surveyed using a non-randomized design, we argue that ecologists routinely test and correct for preferential sampling when fitting species distribution models to animal encounter data.
-
+\item When animal population surveys are conducted without a well defined sampling frame, ecologists should recognize the potential for biasing effects of preferential sampling.
+Joint models, such as those described in this paper, can be used to test and correct for such biases.
 \end{enumerate}
@@
 Designing animal surveys is relatively straightforward in such applications, and unbiased point and variance estimators are available.
-Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture models for repeated point counts \citep{Royle2004a}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,MillerEtAl2013,JohnsonEtAl2010}.
+Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture and Dail-Madsen models for repeated point counts \citep{Royle2004a,DailMadsen2011}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,MillerEtAl2013,JohnsonEtAl2010}.
 In such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g.
@@
 However,
-in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling is nonrandom.
-The potential for biased estimates arises when sampling locations disproportionately target locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.
+in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.
 In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates.
@@
 In this example, $Z_i$ denotes the state variable of interest (e.g., occupancy or abundance), $g()$ is a link function (e.g.
-probit or logit for occupancy, log for count data), and $\mu_i$ is a linear predictor.
-In applications described in this paper, we write the linear predictor as
+probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value.
+In applications described in this paper, we write the intensity as
 \begin{eqnarray}
-\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\eta_i+\epsilon_i,
+\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\lambda_i,
  \label{eq:mu}
 \end{eqnarray}
-where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, $\eta_i$ is a spatially autocorrelated random effect, and $\epsilon_i$ is Gaussian error.
-For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\eta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g.
+where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\lambda}=\{ \lambda_1,\lambda_2,\hdots,\lambda_S \}$ are spatially autocorrelated random effects.
+For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\lambda_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g.
 conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.
@@
 For instance $[{\bf Z}]$ denotes the marginal probability mass function for ${\bf Z}$, and  $[{\bf Z}|{\bf Y}]$ represents the conditional distribution of ${\bf Z}$ given ${\bf Y}$.
+We use $\boldsymbol{\mu}$ and $\boldsymbol{\nu}$ to denote log-scale abundance and the logit of the probability of sampling, so that $Z_i \sim f(\mu_i)$, and $Y_i \sim f(\nu_i)$.
+We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.
 
@@
 However, the manner in which effort is ultimately allocated can potentially have profound influence on SDM estimator performance.
-With respect to nonrandom sampling, two possible problems seem particularly likely: coarse scale preferential sampling (CSPS), and fine scale preferential sampling (FSPS) (Fig.
+With respect to nonrandom sampling, two possible problems seem particularly likely in discrete spatial domains: coarse scale preferential sampling (CSPS), and fine scale preferential sampling (FSPS) (Fig.
 \ref{fig:pref}).
@@
 \citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs, but did not attempt to model it.
+
+In design-based sampling, unequal sampling intensity is often accommodated via stratification or unequal probability sampling, as with Horvitz-Thompson-like estimators where the probability of inclusion varies by sampling unit \citep{Cochran1977}.
+However, in the case of PS, this inclusion probability also depends on the value of the response associated with the sampling unit.
+Evidently, any approach to account for PS should also account for the dependence between the state variable of interest and the locations chosen for sampling.
 
@@
 Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling.
-For the process model, we start with Eq.
+For the abundance process model, we start with eq.
 \ref{eq:process} as a general formulation for non-Gaussian data.
-We then expand the link-scale expectation from this model
-(i.e. Eq. \ref{eq:mu}) as follows:
-\begin{eqnarray}
-  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
-\label{eq:mu_aug}
-\end{eqnarray}
-where $\delta_i$ is a spatially referenced random effect.  Next, we model the binary indicator for sample inclusion using a Bernoulli distribution:
+We then write the link-scale expectation as in eq.
+\ref{eq:mu}.
+Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the linear predictor for this model as
+where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the intensity for this model as
 \begin{eqnarray}
@@
 The predictive covariates ${\bf x}_i$ from Eq.
-\ref{eq:mu_aug} and ${\bf x}_i^*$ from Eq.
+\ref{eq:mu} and ${\bf x}_i^*$ from Eq.
 \ref{eq:nu} need not be the same (although they can be).
 Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs.
-\ref{eq:mu_aug} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.
+\ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.
+The spatially autocorrelated random effects $\eta_i$ are assumed independent of the abundance process.
 
@@
 where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to Eq.
-\ref{eq:mu_aug}.
+\ref{eq:mu}.
 Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig.
@@
 
-We used maximum a posteriori (MAP) estimation to conduct statistical inference.
-This approach is Bayesian; prior parameter distributions are specified and inference is conducted relative to marginal posterior distributions.
-However, in contrast to Markov chain Monte Carlo (MCMC) where the posterior is sampled via simulation, the joint posterior is maximized numerically in this case.
-We used Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}, interfaced with the R programming environment, to conduct maximization.
-The TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{Tierney1989} to obtain abundance estimates that properly account for uncertainty attributable to random effects (\textit{correct, Jim?}).
-This approach allowed for a facile implementation and fast computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with MCMC.
+We used maximum marginal likelihood estimation to conduct statistical inference.
+In particular, we used Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}, interfaced with the R programming environment, to conduct maximization.
+The TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
+This approach allowed for a facile implementation and fast computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.
 In this study, we report the results of 500 simulation replicates.
@@
 Under pathological preferential sampling ($b=5$), both estimation methods were extremely biased, but was even more severe for the naive model ignoring preferential sampling (fig.
-\ref{sim_bias}).
+\ref{fig:sim_bias}).
 
@@
 
+\begin{comment}
 \begin{table}[ht]
@@
 \pagebreak
-
+\end{comment}
 %\begin{figure*}

### 2016-05-26 manuscript/AppendixS1.tex
Complete rough draft of paper; simulations and bearded seal example both worked up

@@
 
-\begin{center} \bf {\large Confronting preferential sampling in count and occupancy surveys: diagnosis and model-based triage}
+\begin{center} \bf {\large Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage}
 
@@
 
-We focus here on a discrete spatial domain; extensions to point process models in continuous space are relatively straightforward.
+We focus here on a discrete spatial domain; extensions to point process models in continuous space should be relatively straightforward.
 In particular, we suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$).
@@
 \begin{eqnarray}
-\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\lambda_i,
+\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
  \label{eq:mu}
 \end{eqnarray}
-where $\beta_0$ is an intercept, ${\bf x}_i$ is a vector of covariates used to explain variation in abundance (here indexed by the site, $i$), and $\boldsymbol{\lambda}=\{ \lambda_1,\lambda_2,\hdots,\lambda_S \}$ are spatially autocorrelated random effects.
+where $\beta_0$ is an intercept, ${\bf x}_i$ is a vector of covariates used to explain variation in abundance (here indexed by the site, $i$), and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.
 
@@
 \hspace{0.5in} We have yet to specify a model for spatial autocorrelation.
-For analyses in this paper, we model mean-zero spatial random effects (i.e., the $\lambda_i$ and $\eta_i$) for each site $i$ as multivariate normal:
+For analyses in this paper, we model mean-zero spatial random effects (i.e., the $\delta_i$ and $\eta_i$) for each site $i$ as multivariate normal:
 \begin{eqnarray*}
-  \boldsymbol{\lambda} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\lambda}), \text{ and} \\
+  \boldsymbol{\lambda} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\delta}), \text{ and} \\
   \boldsymbol{\eta} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\eta}),
@@
 
-To conduct statistical analysis of the preceding models, we use maximum marginal likelihood as implemented in Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}.
+\hspace{0.5in} To conduct statistical analysis of the preceding models, we use maximum marginal likelihood as implemented in Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}.
 Interfaced with the R programming environment, the TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
@@
       \begin{equation*}
-        \exp(X_i \beta_1^*  + \eta_i + b*\delta_i).
+        \exp(X_i \beta_1^*  + \eta_i + b \delta_i).
       \end{equation*}
@@
   \item For each combination of simulation replicate, survey configuration, and estimation model, we recorded relative bias with respect to total estimated abundance, $\hat{N} = \sum \hat{N_i}$, where a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
+This bias correction is also used to obtain abundance estimates in the bearded seal example
 \end{enumerate}

### 2016-05-26 manuscript/pref_sampling.tex
Complete rough draft of paper; simulations and bearded seal example both worked up

@@
 
-\title{Confronting preferential sampling in count and occupancy surveys: diagnosis and model-based triage \footnotemark[2]}
+\title{Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]}
 
@@
 \begin{enumerate}
-\item  Count and occupancy surveys are often used to estimate the density, abundance, and distribution of animal populations.
-Recently, model-based approaches to analyzing survey data (i.e.
-species distribution models) have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures.
-Model-based analysis often makes use of species-covariate relationships and/or spatially autocorrelated random effects to help predict density or occurrence in unsampled locations.
+\item  Wildlife surveys are often used to estimate the density, abundance, or distribution of animal populations.
+Recently, model-based approaches to analyzing survey data have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures.
 
-\item Species distribution models often make the implicit assumption that locations chosen for sampling and animal abundance are conditionally independent given modeled covariates.
+\item Species distribution models fitted to wildlife survey data often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.
 However, this assumption is likely violated in many cases when survey effort is non-randomized, leading to preferential sampling.
 
-\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models.
-The approach works by jointly modeling animal abundance/occurrence and the locations selected for sampling, and specifying a dependent correlation structure between the two models.
+\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models fitted to count data.
+The approach works by jointly modeling wildlife state variables and the locations selected for sampling, and specifying a dependent correlation structure between the two models.
 
-\item Using simulation, we show that under moderate levels of preferential sampling, our modeling approach reduces bias from $\approx$40\% (under a naive species distribution model) to $approx$5\%.
+\item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g.
+$40\%$) bias in estimates of animal density, and that our modeling approach can considerably reduce bias.
 
-\item Species case study
+\item We apply our approach to aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the eastern Bering Sea.
+Models that included a preferential sampling effect led to lower estimates of abundance than models without, but were not favored by an integrated Akaike Information Criterion.
+Importantly, the effect size of the preferential sampling parameter decreased in models that included explanatory environmental covariates.
 
-\item When animal population surveys are conducted without a well defined sampling frame, ecologists should recognize the potential for biasing effects of preferential sampling.
+\item When wildlife surveys are conducted without a well defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling.
 Joint models, such as those described in this paper, can be used to test and correct for such biases.
+Predictive covariates are also useful for bias reduction, but ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
 \end{enumerate}
 
- Word count: XXXX
+ Word count: 5779
 \end{abstract}
@@
 \def\Prob{{\rm P}\,}
+\def\bfX{\bf X}
+\def\bfbeta{\boldsymbol{\beta}}
 
@@
 
-One of the main advantages of using SDMs advanced in the literature is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.
+One of the main advantages of using SDMs is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.
 However,
@@
 
-Need to acknowledge that "sample selection bias" has been addressed extensively in SDM modeling with presence-only data (e.g.
-adjusting "background" or "availability" data to be similar to the locations chosen for sampling).
-
 In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS), and describe several model-based approaches for detecting and correcting for such biases.
 We start by describing a common currency for notation and basic model structures considered in this paper.
-Second, we review preferential sampling bias in a mathematical light, and describe prior approaches to coping with its effects.
+Second, we review PS bias in a mathematical light, and describe prior approaches to coping with its effects.
 Third, we introduce a novel generalization of previously proposed PS models, allowing the investigator to jointly model animal encounter data and the locations chosen for sampling, including possible dependence structure between these two types of observations.
 Fourth, we conduct a simulation study to examine the performance of traditional SDMs and our newly developed PS model when data are gathered preferentially.
-Finally, we demonstrate utility of our proposed modeling approach by analyzing a data set of MYSTERY SPECIES X.
+Finally, we demonstrate our modeling approach by analyzing aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the Bering Sea.
 
@@
 
-We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that preferential sampling is likely to affect analyses similarly regardless of the choice of spatial domain.
+We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.
 We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$).
@@
 \end{eqnarray}
-In this example, $Z_i$ denotes the state variable of interest (e.g., occupancy or abundance), $g()$ is a link function (e.g.
+In this example, $Z_i$ denotes the state variable of interest (e.g.
+occupancy or abundance), $g()$ is a link function (e.g.
 probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value.
@@
 \begin{eqnarray}
-\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\lambda_i,
+\mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i,
  \label{eq:mu}
 \end{eqnarray}
-where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\lambda}=\{ \lambda_1,\lambda_2,\hdots,\lambda_S \}$ are spatially autocorrelated random effects.
-For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\lambda_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g.
+where $\beta_0$ is an intercept parameter, ${\bf x}_i$ is a row vector of $m$ predictive covariates associated with site $i$, $\boldsymbol{\beta} = \{ \beta_1, \beta_2, \hdots, \beta_m \}$ is a column vector of $m$ regression parameters, and $\boldsymbol{\delta}=\{ \delta_1,\delta_2,\hdots,\delta_S \}$ are spatially autocorrelated random effects.
+For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g.
 conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.
@@
 However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.
-For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}
+For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
 \begin{eqnarray}
-  Y_i & \sim & \text{Bernoulli}(Z_i p_i),
+  Y_i & \sim & \text{Bernoulli}(Z_i p_i).
 \end{eqnarray}
-where $p_i$ is possibly a function of survey and observer specific covariates.
+ Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates.
 Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.
@@
 For instance $[{\bf Z}]$ denotes the marginal probability mass function for ${\bf Z}$, and  $[{\bf Z}|{\bf Y}]$ represents the conditional distribution of ${\bf Z}$ given ${\bf Y}$.
-We use $\boldsymbol{\mu}$ and $\boldsymbol{\nu}$ to denote log-scale abundance and the logit of the probability of sampling, so that $Z_i \sim f(\mu_i)$, and $Y_i \sim f(\nu_i)$.
+We use $\boldsymbol{\mu}$ and $\boldsymbol{\nu}$ to denote log-scale abundance and the logit of the probability of sampling, so that $Z_i \sim f(\mu_i)$, and $R_i \sim f(\nu_i)$.
 We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.
@@
 For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high.
-\citet{DiggleEtAl2010} showed that this type of preferential sampling can lead to bias when this extra information is not included in models for the state variable of interest.
-Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}, where \textbf{R} is an indicator vector taking on a value of $R_i = 1.0$ if sampling unit $i$ is sampled and is zero otherwise.
+\citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.
+Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise.
 We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.
@@
 Fitting simple geostatistical models without fixed effects led to positively biased estimates of landscape-level pollutant concentrations.
-Presumably (and as noted by discussants of the article) including a fixed effect associated with a relevant covariate (e.g., an ``urbanity" index) would likely reduce or eliminate bias.
+Presumably (and as noted by discussants of the article) including a fixed effect associated with a relevant covariate (e.g.
+an ``urbanity" index) would likely reduce or eliminate bias.
 However, the primary point of \citet{DiggleEtAl2010} is well taken: inclusion of spatially autocorrelated random effects in a statistical model is insufficient to remove the potentially biasing effects of PS.
@@
 As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS.
-However, in many ecological applications, predictive covariates are only able to explain a portion of variation present in the data.
-If the locations selected for sampling are related (intentionally or unintentionally) to some unmodelled factor related to abundance, bias may still occur.
-Despite the clear potential for bias in SDMs, we have been unable to find many cases where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.
-One such example is
-\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs, but did not attempt to model it.
+However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.
+If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally) , bias may still occur.
+Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.
+One exception is
+\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs.
+However, they did not attempt to account for PS in their models.
 
@@
 \end{eqnarray}
-the relative density of sampling locations at $i$ would be written as
+the spatially continuous relative intensity ($\psi_i$) of sampling locations at $i$ could be written as
 \begin{eqnarray}
-   p_i & \propto & \exp(\xi_i + b \mu_i).
+   \psi_i & \propto & \exp(\xi_i + b \mu_i).
    \label{eq:p_cont}
 \end{eqnarray}
-Here, the parameter $b$ describes the level of preferential sampling; $b=0$ implies no preferential sampling, $b>0$ implies a greater level of sampling in locations where the state variable is anomalously high, and $b<0$ implies greater sampling where the state variable is anomalously low.
+Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the state variable is anomalously high, and $b<0$ implies greater sampling where the state variable is anomalously low.
 Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``.
@@
 explained by the shared spatial covariates."  \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in Eq.
-\ref{eq:Z_cont}, finding that this approach helped reduce bias associated with preferential sampling, but did not perform as well as the full joint model.
+\ref{eq:Z_cont}, finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.
 
@@
 
-The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling preferential sampling.
+The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling PS.
 However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.
-Also, they require the linear predictor of the preferential sampling model to be written as a simple linear function of the the spatial process model for density.
+Also, they require the linear predictor of the PS model to be written as a simple linear function of the the spatial process model for density.
 In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.
-For instance, sampling may be more strongly preferential close to bases of operations, (e.g., landing strips in the case of aerial surveys), but less so in areas that are harder to get to.
+For instance, sampling may be more strongly preferential close to bases of operations, (e.g.
+landing strips in the case of aerial surveys), but less so in areas that are harder to get to.
 
@@
 The spatially autocorrelated random effects $\eta_i$ are assumed independent of the abundance process.
+In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in Eq.
+\ref{eq:nu} are estimated to permit parameter identification.
 
  The formulation in Eq.
-\ref{eq:nu} is the same as previously proposed by other authors for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.
-There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the preferential sampling process \citep{RoyleBerliner1999}.
-For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no preferential sampling).
-Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear preferential sampling model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.
+\ref{eq:nu} is similar to the one previously proposed for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.
+There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the PS process \citep{RoyleBerliner1999}.
+For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no PS).
+Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear PS model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.
 Alternatively, we could allow the degree of PS to vary across the landscape.
-For instance, one can contemplate a trend surface model for preferential sampling by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
+For instance, one can contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
 Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surface or spline formulation \citep{RoyleBerliner1999}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.
@@
 \begin{eqnarray*}
-  N_i \sim \textrm{Poisson}(\mu_i),
+  N_i \sim \textrm{Poisson}(\exp(\mu_i)),
 \end{eqnarray*}
@@
 Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig.
-\ref{PS_sims}).
+\ref{fig:sim_bias}).
 Specific details of of data generation procedures are presented in Appendix S1.
@@
 \ref{eq:R} and \ref{eq:nu}.
-Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, not covariate of spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.
+Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, no covariate or spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.
 In the first, we set ${\bf B}=0$, so that the survey was a simple random sample.
 For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=5$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.
-We refer to these scenarios as moderate and pathological preferential sampling, respectively (see fig.
+We refer to these scenarios as moderate and pathological PS, respectively (see fig.
 \ref{fig:sim_R}).
@@
 The TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
-This approach allowed for a facile implementation and fast computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.
+This approach resulted in a facile implementation and speedy computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.
 In this study, we report the results of 500 simulation replicates.
@@
 
-\subsection{Bearded seal aerial surveys}
+\subsection{Bearded seal count surveys}
 
-\subsection{Avian point counts}
+We applied our modeling technique to counts of bearded seals obtained on aerial transects flown over the eastern Bering Sea from 10-16 April, 2012 (Fig.
+\ref{fig:bearded_ests}).
+These counts were gathered as part of a larger survey designed to estimate abundance of four species of ice-associated seals; the survey is described in greater detail elsewhere \citep{ConnEtAl2014,ConnEtAl2015}.
+The survey area considered here consists of 25 by 25 km grid cells bordered to the north by the Bering straight, to the west by the international date line, to the south by maximal April ice extent, and to the east by the Alaska, USA mainland.
+Here, we limit counts to those gathered within a one week period so that relative abundance will remain relatively constant throughout the study area.
+Our primary focus in this application is to diagnose PS (rather than to estimate absolute abundance).
+As such, we do not attempt to correct for nuisance processes such as incomplete detection or species misclassification, which requires models of increased sophistication \citep{ConnEtAl2014}.
+
+Our choice to model bearded seal counts, as opposed to one of the other seal species, is based on the observation that bearded seal densities tend to be highest in the northern portion of the study area.
+This is also the location of one of the primary airports used to prosecute surveys (Nome, Alaska, USA).
+Higher survey coverage in areas of high bearded seal density could potentially lead to positive bias in apparent abundance owing to PS.
+
+To test for such an effect, we modeled bearded seal counts using the formulation
+\begin{eqnarray*}
+  Y_i & \sim & \textrm{Poisson}(P_i Z_i), \text{ where}\\
+  Z_i & = & A_i \exp(\mu_i),
+ \end{eqnarray*}
+where $P_i$ defines the proportion of grid cell $i$ that is sampled, $A_i$ gives the proportion of grid cell $i$ that is composed of salt water habitat, and $\mu_i$ is defined in Eq. \ref{eq:mu}.
+We modeled the grid cells that were chosen for sampling using Eqs. \ref{eq:R}-\ref{eq:nu}.
+
+We fitted a total of four models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1}$) to bearded seal count data.
+Models varied by (i) whether or not habitat and landscape variables were used as predictors of bearded seal density ($cov=1$ and $cov=0$, respectively), and (ii) whether or not the linear PS parameter $b$ was estimated or fixed to zero ($b=1$ and $b=0$, respectively).
+We also attempted to fit models where the PS ${\bf B}$ matrix varied over the landscape using a trend surface specification, but these models were numerically unstable and are not reported here.
+When habitat and landscape variables were included, we used the following log-linear predictors: sea ice concentration (\textit{ice}), $ice^2$, and distance from the southern ice edge (\textit{edge}).
+Remotely sensed sea ice data were obtained at a $25 \times 25$ km resolution from the National Snow and Ice Data Center, Boulder, CO, USA, as described by \citet{ConnEtAl2014}.
+Models for $\mu_i$ and $\nu_i$ both utilized spatially autocorrelated random effects with a Mat\'{e}rn covariance function between grid cell centroids (see Appendix S1 for further details).
+When covariates were included, they were included in both models (i.e.
+for $\mu_i$ and $\nu_i$).
 
@@
 \ref{fig:sim_bias}).
-Under moderate preferential sampling ($b=1$), the jointly modeling the density and the sampling process (i.e., estimating $b$) led to a median bias of 5\%, while the canonical SDM model had a median bias of 40\%.
-Under pathological preferential sampling ($b=5$), both estimation methods were extremely biased, but was even more severe for the naive model ignoring preferential sampling (fig.
+Under moderate PS ($b=1$), estimation of the PS parameter $b$ led to a median bias of 5\%, while the canonical SDM model ignoring preferential sampling had a median bias of 40\%.
+Under pathological PS ($b=5$), both estimation methods were extremely biased, but was even more severe for the naive model ignoring PS (fig.
 \ref{fig:sim_bias}).
 
+\subsection{Bearded seal analysis}
+
+Integrated AIC$_\textrm{c}$ strongly favored models with covariate effects, and gave most (90\%) support to models without a PS effect (Table \ref{tab:aic}).
+However, further intuition can be gained by gained by examining estimates of the PS parameter, $b$.
+For the model without predictive covariates,  $\hat{b}=0.27$ (SE 0.11), and for the model with predictive covariates, $\hat{b}=0.19$ (SE 0.13).
+Thus, there did appear to be some evidence for positive PS that was not picked up by examining integrated AIC$_\textrm{c}$ scores.
+However, the magnitude of PS decreased when covariate predictors were included in the model, an effect that follows theoretical results provided by \citet{PatiEtAl2011}.
+Estimates of abundance were substantially higher for models without a PS effect, with the non-PS model having a 49\% higher estimate when covariates were not modeled, and a 19\% higher estimate when covariate effects were included.
+
+Note that unlike the other models, $M_{cov=1,b=0}$ predicted anomalously high bearded seal abundance in the extreme southern portion of the study area where sea ice was absent (and thus there was no habitat for seals).
+Thus, while we present original likelihood and AIC$_\textrm{c}$ values to permit direct comparison with other models, we refitted $M_{cov=1,b=0}$ to produce an estimate of abundance without this feature.
+Specifically, we refitted the model with 20 pseudo-absences in this portion of the study area to better inform abundance-covariate relationships.
+
 \section{Discussion}
+
+In this study, we showed that course-grained preferential sampling (Fig.
+\ref{fig:pref}) can have a profound impact on the quality of estimates (e.g.
+animal abundance) when sampling is non-randomized.
+In simulations, estimators were increasingly positively biased as PS increased.
+When PS was present, we were able to substantially reduce bias by conducting estimation under a framework where the state variable of interest and the sites chosen for sampling were jointly modeled under a dependent covariance structure.
+In absence of PS, simulations suggest that this structure results in lower precision than a model without a PS effect; thus the need to account for PS reduces the quality of inference.
 
@@
 The obvious solution in this instance is to compensate for PS in model-based inferences by accounting for variation in sampling intensity with explanatory covariates or post hoc stratification.
-However, this does not always work when effort is allocated in a subjective manner.
+However, it is not always clear how to perform such stratification when effort is allocated in a subjective manner.
 
-Extension to continuous space - sampling locations as point process similar to Warton and Shepherd (2010)
+When applied to bearded seal count data, greater support was given to models without a PS effect and to those that had fixed effects associated with environmental and landscape covariates.
+Somewhat disturbingly, however, the PS effect size was estimated to be positive and to produce considerably lower abundance estimates than models without a PS effect.
+Differences between apparent abundance predictions decreased when covariates were added to model structure, confirming previous theoretical results \citep{DiggleEtAl2010,PatiEtAl2011} that covariates serve to decrease the conditional dependence between site selection and the state variable of interest.
 
-Presumably, trends less biased when pref sampling ignored.
+We attempted to fit models to bearded seal data where the degree of PS changed over the landscape, in a similar manner to multivariate spatial models \citep{RoyleBerliner1999}.
+However, such models appeared numerically unstable in our application.
+Evidently, such models may require rich data sets, where the number of locations sampled is greater than in our example ($n=394$) or has a better degree of spatial balance.
+As this seems unlikely in many ecological studies, we suggest limiting initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
 
- \acks{Later.}
+The models we have developed here are specific to spatial models with discrete support, as when data are gathered at a plot level, or aggregated prior to analysis.
+However, it should be possible to extend our approach to continuous space.
+One approach would be to model sampling locations as realizations from a spatial point process in a manner similar to \citet{WartonShepherd2010}.
+
+Our conception of PS is related, but not equivalent to ``sample selection bias" \citep[e.g.][]{PhillipsEtAl2009} in presence-only models.
+In the latter, absence of a species at a given site is never directly observed.
+To make inferences about space use, it is thus necessary to produce a background sample representing the range of locations and habitats that could have been sampled.
+Sample selection bias then results if the characteristics of sites selected for sampling (e.g.
+by a volunteer or museum collector) differ systematically from the assumed background sample.
+In our case, we use PS to refer to the case where absences are available, but where the probability of sampling is dependent on some unknown factor that is also related to abundance or presence of the target species.
+
+\section{Conclusion}
+
+Model-based approaches to estimation of abundance or occurrence have become popular in recent years.
+We (the authors) have noticed a tendency for analysts to assume that inclusion of spatial covariates or random effects into predictive models will make the underlying sampling design ignorable.
+We have shown in this paper that this is not the case, although our results do suggest that including predictive covariates can indeed decrease bias from preferential sampling.
+We have also shown that it is possible to further diagnose and adjust for preferential sampling by jointly modeling dependence between the data collection mechanism and the process of interest (e.g.
+abundance or occupancy).
+However, such models can be less precise and have greater instability than models without a preferential sampling parameter and it would be preferable not to have to go to such elaborations.
+Where possible, we suggest that survey planners incorporate design-based elements (e.g.
+random or systematic sampling) into their survey designs to reduce the need for model-based triage.
+
+ \acks{
+Funding for aerial surveys was provided
+by the U.S.
+National Oceanic and Atmospheric Administration and by the U.S.
+Bureau of Ocean Energy Management.
+Views expressed are those of the authors and do not necessarily represent findings or policy of any government agency.
+Use of trade or brand names does not indicate endorsement by the U.S.
+government.}
+
+\vspace{.3in}
+\section{Data accessibility} 
+R scripts and data necessary to recreate analyses have been collated into an R package, which is currently available at \url{https://github.com/NMML/pref_sampling}.
+We plan to publish the package to an online archive/repository upon acceptance.
+\\
 
@@
 \pagebreak
-
-\begin{comment}
 \begin{table}[ht]
-\label{tab:PS}
-\caption{Performance of count-based abundance estimators as a function of survey design and estimation model.
-Performance measures include proportional relative bias (``Bias"), root mean squared error (``RMSE"), coefficient of variation (``CV"), and 90\% credible interval coverage (``Cov90''; the proportion of simulations where true abundance was between the 5th and 95th percentiles of posterior samples).
-Mean values over 400 simulation replicates are presented for spatially balanced sampling (``Balanced"), inverse probability sampling based on covariate values (``Covariate"), and inverse probability sampling based on a priori knowledge of areas of high abundance (``Preferential").
-In addition, two different estimation models were applied to each dataset, including a generalized linear model (``GLM''), and a spatial regression model (``RSR'').}
-\centering
-\begin{tabular}{rrlrrrr}
+\caption{A summary of model selection results and estimated abundance for the four models fitted to bearded seal counts.
+The models include formulations with or without predictive covariates ($cov=1$ or 0, respectively) , and with or without the preferential sampling parameter $b$ estimated ($b=1$ or 0, respectively) .
+All models included spatially autocorrelated random effects on log-scale abundance intensity.
+Shown are the log integrated likelihood, the number of fixed effect parameters, $\Delta \textrm{AIC}_\textrm{c}$, $\textrm{AIC}_\textrm{c}$ model weights, and estimated apparent abundance over the landscape ($\hat{N}$) together with a Hessian-based standard error estimate.
+}
+\label{tab:aic}
+\raggedright
+\begin{tabular}{lccccc}
   \hline
- Design & Model & Bias & RMSE & CV & Cov90 \\
+  Model & Log likelihood & Params & $\Delta \textrm{AIC}_\textrm{c}$ & Wgt & $\hat{N}$(SE) \\
   \hline
-Balanced & GLM & 0.00 & 1038 & 0.069 & 0.92 \\
-Balanced & RSR & 0.00 & 1034 & 0.057 & 0.90 \\
-Covariate & GLM & 0.00 & 1392 & 0.067 & 0.88 \\
-Covariate & RSR & 0.00 & 1152 & 0.061 & 0.88 \\
-Preferential & GLM & 0.21 & 5807 & 0.060 & 0.24 \\
-Preferential & RSR & 0.15 & 3040 & 0.054 & 0.29 \\
+  $M_{cov=0,b=0}$ & -3326.0 & 3 & 9.9 & 0.007 & 68556 (7408)\\
+  $M_{cov=0,b=1}$ & -3327.3 & 4 & 14.6 & $<0.001$ & 45857 (5114) \\
+  $M_{cov=1,b=0}$ & -3315.1 & 9 & 0.0 & 0.891 & 59312$^\dag$ (5231)  \\
+  $M_{cov=1,b=1}$ & -3316.2 & 10 & 4.4 & 0.102 & 49826 (10369) \\
 \hline
 \end{tabular}
+\\
+$\dag$ Refitted model; see \textit{Results}.
 \end{table}
@@
 \pagebreak
-\end{comment}
 %\begin{figure*}
@@
 In (B), a fine scale version of preferential sampling occurs when a line transect (red line) is intentionally placed across a region of high quality habitat.
-If a landscape is discretized into homogeneous survey units (as in a grid), it is essential that the habitat surveyed within each survey unit be randomly determined when estimating abundance.
+If a landscape is discretized into homogeneous survey units for analysis (as in a grid), it is essential that the habitat surveyed within each survey unit be randomly determined when estimating abundance.
 If not, bias (usually positive) can be expected.}
@@
 
+\begin{figure*} %[ht]
+\begin{center}
+\includegraphics[width=170mm]{bearded_ests.pdf}
+\caption{Aerial survey counts and estimated apparent abundance of bearded seals in the eastern Bering Sea, April 10-16, 2012.
+Counts and estimates are shown relative to a survey grid that extends south from the Bering Strait and borders the Alaska, USA mainland to the east.
+In (A), tan shading denotes land, unsurveyed grid cells appear in dark gray, and counts appear in a white-blue spectrum.
+Apparent bearded seal abundance estimates (B) are presented from the model with the lowest integrated AICc score, which included covariate effects but no preferential sampling effect.
+Apparent abundance estimates are uncorrected for imperfect detection or species misclassification.}
+\label{fig:bearded_ests}
+\end{center}
+\end{figure*}
+
 \end{document}

### 2016-05-26 TMB_version/inst/executables/simple_v7.cpp

@@ -1,0 +1,261 @@
+// Space time 
+#include <TMB.hpp>
+//#include <atomic_math.hpp>  //for D_lgamma
+
+
+/** Precision matrix for the anisotropic case, eqn (20) in Lindgren et al. (2011) */    
+namespace R_inla_generalized {
+using namespace Eigen;
+using namespace tmbutils;
+using namespace R_inla;
+
+template<class Type>
+  SparseMatrix<Type> Q_spde_generalized(spde_t<Type> spde, Type kappa, int alpha=2){
+  Type kappa_pow2 = kappa*kappa;
+  Type kappa_pow4 = kappa_pow2*kappa_pow2;
+  	
+  if( alpha==1 ) return kappa_pow2*spde.M0 + spde.M1;
+  if( alpha==2 ) return kappa_pow4*spde.M0 + Type(2.0)*kappa_pow2*spde.M1 + spde.M2;
+}
+
+template<class Type>
+  SparseMatrix<Type> Q_spde_generalized(spde_aniso_t<Type> spde, Type kappa, matrix<Type> H, int alpha=2){
+
+  int i;
+  Type kappa_pow2 = kappa*kappa;
+  Type kappa_pow4 = kappa_pow2*kappa_pow2;
+  
+  int n_s = spde.n_s;
+  int n_tri = spde.n_tri;
+  vector<Type> Tri_Area = spde.Tri_Area;
+  matrix<Type> E0 = spde.E0;
+  matrix<Type> E1 = spde.E1;
+  matrix<Type> E2 = spde.E2;
+  matrix<int> TV = spde.TV;
+  SparseMatrix<Type> G0 = spde.G0;
+  SparseMatrix<Type> G0_inv = spde.G0_inv;
+	  	  
+  //Type H_trace = H(0,0)+H(1,1);
+  //Type H_det = H(0,0)*H(1,1)-H(0,1)*H(1,0);
+  SparseMatrix<Type> G1_aniso(n_s,n_s); 
+  SparseMatrix<Type> G2_aniso(n_s,n_s); 
+  // Calculate adjugate of H
+  matrix<Type> adj_H(2,2);
+  adj_H(0,0) = H(1,1);
+  adj_H(0,1) = -1 * H(0,1);
+  adj_H(1,0) = -1 * H(1,0);
+  adj_H(1,1) = H(0,0);
[…]

### 2016-05-26 pref_sampling/R/util_funcs.R

@@ -113,19 +113,23 @@ d_biv_normal<-function(Tmp.vec,XY,Sigma){
   
 #' Function to plot abundance map for Bering Sea survey grid
 #' @param cur.t  Time step to plot map for
-#' @param N A matrix holding abundance point estimates; different rows correspond to different sampling units, columns correspond to time step
-#' @param Grid A list object, each element of which holds a spatial polygons data frame for each time step
+#' @param N A matrix holding abundance point estimates; different rows correspond to different sampling units
+#' @param Grid A spatial polygons data frame holding grid entries
 #' @param highlight If provided, this vector specifies which cells to separately highlight during plotting
 #' @param cell.width if highlight is provided, cell.width must provide the width of a composite grid cell
 #' @param leg.title Title for legend of plot (if different from the default "Abundance")
 #' @param hcolor If highight provided, gives color for highlighting (default = "yellow")
 #' @param cur.max If provided, sets maximum value for plotted abundance
+#' @param myPallete A colorRampPalette; default is a spectral palette from RColorBrewer
 #' @return An abundance map
 #' @export 
 #' @keywords abundance map, plot
 #' @author Paul Conn \email{paul.conn@@noaa.gov}
-plot_N_map<-function(N,Grid,highlight=NULL,cell.width=1,leg.title="Abundance",hcolor='yellow',cur.max=NULL){
-  myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))  
+plot_N_map<-function(cur.t=1,N,Grid,highlight=NULL,cell.width=1,leg.title="Abundance",hcolor='yellow',cur.max=NULL,myPalette=NULL){
+  require(ggplot2)
+  require(rgeos)
+  require(RColorBrewer)
+  if(is.null(myPalette)==TRUE)myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))  
   Tmp=Grid
   if(is.null(highlight)==FALSE){
     midpoints=data.frame(gCentroid(Tmp[highlight,],byid=TRUE))

### 2016-05-26 pref_sampling/inst/run_bearded_models.R

@@ -1,0 +1,226 @@
+### run simple preferential sampling example.  Doing in this in two steps since there seems to be
+# a conflict between TMB loop and rgeos ('gCentroid') functionality
+# need to declare an ADREPORT for Z_s in TMB for this
+setwd('c:/users/paul.conn/git/pref_sampling')
+source('./pref_sampling/R/util_funcs.R')
+library(sp)
+library(grid)
+library(maptools)
+library( RandomFields )
+library( TMB )
+library( INLA )
+library(rgeos)
+library(gridExtra)
+#load("Effort2012_bearded_PrefSamp_21Dec2015.Rdata")  
+load('Data_for_ST_plot.Rdata') #includes spatial objects associated w landscape
+load("Bearded_effort.Rda")  #formatted with BOSSst "format_Bering2012_all.R"
+load("AlaskaBeringData2012_2013_14Dec2015.Rdat")  #from bass package on github/nmml
+Effort=Bearded_effort
+#library( ThorsonUtilities )
+
+Version = "simple_v7"
+# v2 -- added option for changing distribution for random effects, and implemented ICAR distribution
+# v3 -- added Bernoulli model for sampling locations, and removed the ICAR functionality (by commenting it out, to save speed during building the object)
+# v4 -- added simpler less-smooth SPDE option (may be faster!), and option to turn off delta or eta
+# v5 -- Implements Devin's finite population correction approach to posterior prediction; proportion area sampled additional input vector  
+# v6 -- Includes option to output Z_s in ADREPORT so unbiased map estimates can be made using bias.correct
+# v7 -- Includes option to allows b to vary as a trend surface
+
+# Compile
+TmbFile = "C:/Users/paul.conn/git/pref_sampling/TMB_version/inst/executables"
+setwd( TmbFile )
+compile( paste0(Version,".cpp") )
+
+# SOurce
+#source( paste0(TmbFile,"/../../examples/rect_adj.R") )
[…]

### 2016-06-11 manuscript/AppendixS2.tex
Changes in code to fix pulling off wrong values of likelihood for AIC calcs in bearded seal example; changes to manuscript, and new Appendix S2 detailing trend surface model failure.  Paper version sent out for PRC (for the 2nd time).

% besdoc.tex V1.0, 17 March 2011

\documentclass[times,mee,doublespace,]{besauth2}
%%\documentclass[times,mee,]{besauth}

\newcommand{\journalnamelc}{British Ecological Society}
\newcommand{\journalabb}{British Ecological Society}
\newcommand{\journalclassshort}{BES}
%%\newcommand{\journalname}{British Ecological Society}
\usepackage{epstopdf,comment}

\usepackage{moreverb}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\usepackage{lineno}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\bibpunct[; ]{(}{)}{;}{}{}{,}

\def\volumeyear{2015}
\def\VOC{VO}

\begin{document}

\begin{center} \bf {\large Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage}

\vspace{0.7cm}
Paul B. Conn$^{1,*}$, James T. Thorson$^2$, Devin S. Johnson$^1$
\end{center}
\vspace{0.5cm}

\rm
\small

$^1$ National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; $^2$ Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, National Marine Fisheries Service (NMFS),
NOAA, 2725 Montlake Boulevard E, Seattle, WA 98112, USA

$^*$paul.conn@noaa.gov

\large
\bigskip
\centerline{Appendix S2: Specific details on parameter estimation in bearded seal example}
\bigskip
\small

\linenumbers

\def\VAR{{\rm Var}\,}
\def\COV{{\rm Cov}\,}
\def\Prob{{\rm P}\,}

\begin{flushleft}

In addition to the models reported in the paper, we attempted to fit two different models with a trend surface specification for the preferential sampling process.  The models differed by whether or not predictive covariates were also modeled (the ice covariates described in the main article text), but otherwise the preferential sampling process was modeled similarly.  Specifically, when the probability $R_i$ that a location $i$ is selected for sampling is modeled as
\begin{eqnarray}
 \label{eq:R}
  R_i & \sim & \text{Bernoulli}(\textrm{expit}(\nu_i))),
\end{eqnarray}
where
\begin{eqnarray}
  \label{eq:nu}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i,
\label{eq:nu}
\end{eqnarray}
we implemented a trend surface model by writing $\bf{B}$ as
\begin{eqnarray}
  \label{eq:B}
  \textbf{B} & = & {\bf C}{\bf b}.
\end{eqnarray} 
Here, ${\bf b}$ is a column vector consisting of three estimated parameters, $b_0$, $b_1$, and $b_2$, and ${\bf C}$ is an $(S \times 3)$ design matrix constructed as follows:
\begin{eqnarray}
  \textbf{C} & = & \left[ 
    \begin{array}{ccc}
    1 & northing_1 & easting_1 \\
    1 & northing_2 & easting_2 \\
    \vdots & \vdots & \vdots \\
    1 & northing_S & easting_S 
    \end{array} 
  \right].
\end{eqnarray}
The northing and easting values are simply the $y$ and $x$ coordinates associated with each grid cell centroid on the particular geographical projection of the earth's surface that we were using (in our case, EASE Grid 2.0).

\hspace{0.5in} On the surface, there did not appear to be any numerical errors when fitting these trend surface models to bearded seal data.  The Hessian was invertible, and incorporation of the trend surface model resulted in a much lower AIC score than the other models (AIC scores were 19 and 26 points better than the highest ranked AIC model reported in the paper for the trend surface models omitting and including covariate effects, respectively).  However, when examining estimates graphically, it was clear that the trend surface models suffered from some parameter identification and/or aliasing issues (Fig. \ref{fig:bearded_RE_maps}).  Specifically, most variation in the probability of sampling was passed on from the $\boldsymbol{\eta}$ random effect to the shared random effect $\boldsymbol{\delta}$ such that it was difficult to disentangle the sampling process from the abundance process.  To our mind, this was a clear sign that the model was overparameterized, leading to poor estimation.

\raggedbottom

\bibliographystyle{bes}
\bibliography{master_bib}

\pagebreak

\begin{figure*} %[ht]
\begin{center}
\includegraphics[width=170mm]{bearded_RE_maps.pdf}
\caption{Estimates of spatial random effects for two models fitted to bearded seal count data.  Estimates from preferential sampling model $M_{cov=1,b=1}$ are provided in the first column of plots, while estimates from a second model with a trend surface specification for the preferential sampling parameter appear in the second column (this model also included covariate effects).  In the first column, the $\boldsymbol{\eta}$ random effects specific to the sampling probability model appear well estimated and to indicate a higher probability of sampling where transects were flown (as intended).  However, under the trend surface specification, these random effects are estimated to be close to zero, and the increased probability of sampling is passed on to the shared $\boldsymbol{\delta}$ random effects.} 
\label{fig:bearded_RE_maps}
\end{center}
\end{figure*}

\end{flushleft}
\end{document}

%http://www.plosone.org/article/info:doi/10.1371/journal.pone.0036527

### 2016-06-11 manuscript/pref_sampling.tex
Changes in code to fix pulling off wrong values of likelihood for AIC calcs in bearded seal example; changes to manuscript, and new Appendix S2 detailing trend surface model failure.  Paper version sent out for PRC (for the 2nd time).

@@
 \item We apply our approach to aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the eastern Bering Sea.
-Models that included a preferential sampling effect led to lower estimates of abundance than models without, but were not favored by an integrated Akaike Information Criterion.
-Importantly, the effect size of the preferential sampling parameter decreased in models that included explanatory environmental covariates.
+Models that included a preferential sampling effect led to lower estimates of abundance than models without, but the effect size of the preferential sampling parameter decreased in models that included explanatory environmental covariates.
 
@@
 
- Word count: 5779
+ Word count: 6038
 \end{abstract}
@@
 \def\bfbeta{\boldsymbol{\beta}}
+\def\bfdelta{\boldsymbol{\delta}}
+\def\bfeta{\boldsymbol{\eta}}
 
@@
 We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$).
-Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if survey unit $i$ is sampled (i.e.
+Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if location $i$ is sampled (i.e.
 if $U_i \in \mathcal{S}$), and is 0 otherwise.
@@
 Presumably (and as noted by discussants of the article) including a fixed effect associated with a relevant covariate (e.g.
-an ``urbanity" index) would likely reduce or eliminate bias.
+a development index) would likely reduce or eliminate bias.
 However, the primary point of \citet{DiggleEtAl2010} is well taken: inclusion of spatially autocorrelated random effects in a statistical model is insufficient to remove the potentially biasing effects of PS.
@@
 \begin{eqnarray}
-   Z_i & = & \mu_i + \epsilon,
+   Z_i & = & \mu_i + \epsilon_i,
    \label{eq:Z_cont}
@@
 \end{eqnarray}
-Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the state variable is anomalously high, and $b<0$ implies greater sampling where the state variable is anomalously low.
+Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g.
+animal density) is high, and $b<0$ implies greater sampling where the spatial process is low.
 Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``.
@@
 For the abundance process model, we start with eq.
-\ref{eq:process} as a general formulation for non-Gaussian data.
-We then write the link-scale expectation as in eq.
+\ref{eq:process} as a general formulation for non-Gaussian data, writing the link-scale expectation as in eq.
 \ref{eq:mu}.
@@
 \ref{eq:mu} and ${\bf x}_i^*$ from Eq.
-\ref{eq:nu} need not be the same (although they can be).
+\ref{eq:nu} may or may not be the same.
 Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs.
 \ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.
-The spatially autocorrelated random effects $\eta_i$ are assumed independent of the abundance process.
+The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.
 In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in Eq.
@@
 There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the PS process \citep{RoyleBerliner1999}.
-For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no PS).
+For instance, setting ${\bf B} = {\bf 0}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no PS).
 Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear PS model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.
@@
 To illustrate PS and demonstrate that our proposed model has reasonable performance, we conducted a small simulation experiment.
-For each of 100 simulations, we generated abundance of a hypothetical species over a $25 \times 25$ grid as
+For each of 500 simulations, we generated abundance of a hypothetical species over a $25 \times 25$ grid as
 \begin{eqnarray*}
-  N_i \sim \textrm{Poisson}(\exp(\mu_i)),
+  [N_i|\mu_i] & = & \textrm{Poisson}(\exp(\mu_i)),
 \end{eqnarray*}
@@
 Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig.
-\ref{fig:sim_bias}).
+\ref{fig:PS_sims}).
 Specific details of of data generation procedures are presented in Appendix S1.
@@
 Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, no covariate or spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.
-In the first, we set ${\bf B}=0$, so that the survey was a simple random sample.
+In the first, we set ${\bf B}=0$, so that surveyed locations were selected independently of the abundance generating process.
 For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=5$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.
@@
 We fitted two different models to each count dataset, both of which were provided the habitat covariate used (in part) to generate the data for which a log-linear coefficient $\beta$ was estimated.
-In the first estimation model, the elements of \textbf{B} in eq.
+In the first model, the elements of \textbf{B} in eq.
 \ref{eq:nu} were all set to zero.
 In this case, the abundance and sampling process submodels were independent, as is the case canonical SDMs (at lest when fitted to presence-absence or count data).
-In the second estimation model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix.
+In the second model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix.
 
-We used maximum marginal likelihood estimation to conduct statistical inference.
+We used maximum likelihood to conduct statistical inference.
 In particular, we used Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}, interfaced with the R programming environment, to conduct maximization.
-The TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
+The TMB software uses a Laplace approximation to integrate out random effects ($\bfeta$ and $\bfdelta$), and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
 This approach resulted in a facile implementation and speedy computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.
-In this study, we report the results of 500 simulation replicates.
 Further detail on statistical methods are provided in Appendix S1; requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.
@@
 These counts were gathered as part of a larger survey designed to estimate abundance of four species of ice-associated seals; the survey is described in greater detail elsewhere \citep{ConnEtAl2014,ConnEtAl2015}.
-The survey area considered here consists of 25 by 25 km grid cells bordered to the north by the Bering straight, to the west by the international date line, to the south by maximal April ice extent, and to the east by the Alaska, USA mainland.
+The survey area considered here consists of 25 by 25 km grid cells bordered to the north by the Bering strait, to the west by the international date line, to the south by maximal April ice extent, and to the east by the Alaska, USA mainland.
 Here, we limit counts to those gathered within a one week period so that relative abundance will remain relatively constant throughout the study area.
@@
 \begin{eqnarray*}
-  Y_i & \sim & \textrm{Poisson}(P_i Z_i), \text{ where}\\
+  [Y_i|Z_i] & = & \textrm{Poisson}(P_i Z_i), \text{ where}\\
   Z_i & = & A_i \exp(\mu_i),
@@
 
-We fitted a total of four models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1}$) to bearded seal count data.
-Models varied by (i) whether or not habitat and landscape variables were used as predictors of bearded seal density ($cov=1$ and $cov=0$, respectively), and (ii) whether or not the linear PS parameter $b$ was estimated or fixed to zero ($b=1$ and $b=0$, respectively).
-We also attempted to fit models where the PS ${\bf B}$ matrix varied over the landscape using a trend surface specification, but these models were numerically unstable and are not reported here.
-When habitat and landscape variables were included, we used the following log-linear predictors: sea ice concentration (\textit{ice}), $ice^2$, and distance from the southern ice edge (\textit{edge}).
+We fitted a total of six models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1}$) to bearded seal count data using the same estimation framework as in the simulation study.
+Models varied by (i) whether or not habitat and landscape variables were used as predictors of bearded seal density ($cov=1$ and $cov=0$, respectively), and (ii) the form of PS ($b=0$ indicates no PS; $b=1$ indicates ${\bf B}=b{\bf I}$, where $b$ is an estimated parameter).
+We also attempted to fit models where the PS ${\bf B}$ matrix varied over the landscape using a trend surface specification, but parameter identification was suspect in these models and are not reported here (see Appendix S2 for more information).
+When habitat and landscape variables were included, we used three log-linear predictors: linear and quadratic functions of sea ice concentration, and distance from the southern ice edge.
 Remotely sensed sea ice data were obtained at a $25 \times 25$ km resolution from the National Snow and Ice Data Center, Boulder, CO, USA, as described by \citet{ConnEtAl2014}.
@@
 
-Integrated AIC$_\textrm{c}$ strongly favored models with covariate effects, and gave most (90\%) support to models without a PS effect (Table \ref{tab:aic}).
-However, further intuition can be gained by gained by examining estimates of the PS parameter, $b$.
-For the model without predictive covariates,  $\hat{b}=0.27$ (SE 0.11), and for the model with predictive covariates, $\hat{b}=0.19$ (SE 0.13).
-Thus, there did appear to be some evidence for positive PS that was not picked up by examining integrated AIC$_\textrm{c}$ scores.
-However, the magnitude of PS decreased when covariate predictors were included in the model, an effect that follows theoretical results provided by \citet{PatiEtAl2011}.
+Integrated AIC strongly favored models with covariate effects, but for such models the presence of PS was equivocal (Table \ref{tab:aic}).
+Further intuition can be gained by examining estimates of the PS parameter, $b$.
+For the PS model without predictive covariates,  $\hat{b}=0.27$ (SE 0.11), and for the model with predictive covariates, $\hat{b}=0.19$ (SE 0.13).
+Thus, it appeared that including predictive covariates decreased the PS effect size, as suggested by \citet{PatiEtAl2011}.
 Estimates of abundance were substantially higher for models without a PS effect, with the non-PS model having a 49\% higher estimate when covariates were not modeled, and a 19\% higher estimate when covariate effects were included.
 
-Note that unlike the other models, $M_{cov=1,b=0}$ predicted anomalously high bearded seal abundance in the extreme southern portion of the study area where sea ice was absent (and thus there was no habitat for seals).
-Thus, while we present original likelihood and AIC$_\textrm{c}$ values to permit direct comparison with other models, we refitted $M_{cov=1,b=0}$ to produce an estimate of abundance without this feature.
+Note that unlike the other models, $M_{cov=1,b=0}$ predicted anomalously high bearded seal abundance in the extreme southern portion of the study area where sea ice was absent (where there was no habitat for seals).
+Thus, while we present original likelihood and AIC values to permit direct comparison with other models, we refitted $M_{cov=1,b=0}$ to produce an estimate of apparent abundance without this feature.
 Specifically, we refitted the model with 20 pseudo-absences in this portion of the study area to better inform abundance-covariate relationships.
+
+The fact that models with and without a PS effect garner approximately equal weight suggests a need to account for PS when producing abundance estimates from this data set.
+A model averaged estimate calculated using AIC machinery \citep{BurnhamAnderson2002} is 54854 (SE 9351), which is 7.5\% less than the estimate assuming no PS.
+Notably, the standard error of the model averaged estimate was 79\% higher than the model assuming no PS.
 
@@
 
-When applied to bearded seal count data, greater support was given to models without a PS effect and to those that had fixed effects associated with environmental and landscape covariates.
-Somewhat disturbingly, however, the PS effect size was estimated to be positive and to produce considerably lower abundance estimates than models without a PS effect.
-Differences between apparent abundance predictions decreased when covariates were added to model structure, confirming previous theoretical results \citep{DiggleEtAl2010,PatiEtAl2011} that covariates serve to decrease the conditional dependence between site selection and the state variable of interest.
+When applied to bearded seal count data, approximately equal support was given to models with and without a PS effect.
+The PS effect size was estimated to be positive and to produce considerably lower abundance estimates than models without a PS effect.
+Differences between apparent abundance estimates decreased when covariates were added to model structure, supporting previous theoretical results \citep{DiggleEtAl2010,PatiEtAl2011} that covariates serve to decrease the conditional dependence between site selection and the state variable of interest.
+However, in our data set, adding covariates did not eliminate evidence of PS.
+Accounting for PS in a model averaging framework led to a moderate decrease in our apparent abundance estimate for bearded seals in this region, and markedly decreased precision.
+As in our simulations, the need to account for PS thus appeared to have a real cost in terms variance inflation.
 
 We attempted to fit models to bearded seal data where the degree of PS changed over the landscape, in a similar manner to multivariate spatial models \citep{RoyleBerliner1999}.
-However, such models appeared numerically unstable in our application.
-Evidently, such models may require rich data sets, where the number of locations sampled is greater than in our example ($n=394$) or has a better degree of spatial balance.
-As this seems unlikely in many ecological studies, we suggest limiting initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
+However, such models led to difficulties with parameter identification in our bearded seal application (see Appendix S2).
+Evidently, such models may require greater spatial balance or richer data sets.
+At this time, we suggest limiting initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
+Although more complex models are clearly identifiable in some situations \citep{RoyleBerliner1999}, further research on the viability of such models as a function of data quality appears warranted.
+It would also be worthwhile to investigate whether parameter identification varies as a function of the support of the state variable being modeled (e.g.
+binary vs.
+count data).
 
@@
 One approach would be to model sampling locations as realizations from a spatial point process in a manner similar to \citet{WartonShepherd2010}.
+Another possible extension would be to consider models for the sampling process where sampling occurs without replacement for a fixed sample size.
+For instance, the Bernoulli sampling model makes the implicit assumption that sample size is random.
+If, instead, a fixed number of locations are sampled, the Bernoulli model is somewhat misspecified.
+Our simulations suggest some robustness to this misspecification, as the Bernoulli model performed reasonably well when sampling was without replacement for a fixed sample size (Fig.
+\ref{fig:sim_bias}).
+Still, a more precise treatment would need to rely on an extended hypergeometric distribution with variable inclusion probabilities when formulating the sampling model; this extension is nontrivial.
 
 Our conception of PS is related, but not equivalent to ``sample selection bias" \citep[e.g.][]{PhillipsEtAl2009} in presence-only models.
-In the latter, absence of a species at a given site is never directly observed.
-To make inferences about space use, it is thus necessary to produce a background sample representing the range of locations and habitats that could have been sampled.
+In such models, absence of a species at a given site is never directly observed.
+To draw inference about space use, it is thus necessary to produce a background sample representing the range of locations and habitats that could have been sampled.
 Sample selection bias then results if the characteristics of sites selected for sampling (e.g.
@@
 abundance or occupancy).
-However, such models can be less precise and have greater instability than models without a preferential sampling parameter and it would be preferable not to have to go to such elaborations.
+However, such models can be considerably less precise and have greater instability than models without a preferential sampling parameter and it would be preferable not to have to go to such elaborations.
 Where possible, we suggest that survey planners incorporate design-based elements (e.g.
@@
 All models included spatially autocorrelated random effects on log-scale abundance intensity.
-Shown are the log integrated likelihood, the number of fixed effect parameters, $\Delta \textrm{AIC}_\textrm{c}$, $\textrm{AIC}_\textrm{c}$ model weights, and estimated apparent abundance over the landscape ($\hat{N}$) together with a Hessian-based standard error estimate.
+Shown are the log integrated likelihood, the number of fixed effect parameters, $\Delta \textrm{AIC}$, AIC model weights, and estimated apparent abundance over the landscape ($\hat{N}$) together with a Hessian-based standard error estimate.
 }
@@
   \hline
-  Model & Log likelihood & Params & $\Delta \textrm{AIC}_\textrm{c}$ & Wgt & $\hat{N}$(SE) \\
+  Model & Log likelihood & Params & $\Delta \textrm{AIC}$ & Wgt & $\hat{N}$(SE) \\
   \hline
-  $M_{cov=0,b=0}$ & -3326.0 & 3 & 9.9 & 0.007 & 68556 (7408)\\
-  $M_{cov=0,b=1}$ & -3327.3 & 4 & 14.6 & $<0.001$ & 45857 (5114) \\
-  $M_{cov=1,b=0}$ & -3315.1 & 9 & 0.0 & 0.891 & 59312$^\dag$ (5231)  \\
-  $M_{cov=1,b=1}$ & -3316.2 & 10 & 4.4 & 0.102 & 49826 (10369) \\
+  $M_{cov=0,b=0}$ & -2667.1 & 3 & 21.7 & 0.00 & 68556 (7408)\\
+  $M_{cov=0,b=1}$ & -2665.3 & 4 & 20.1 & 0.00 & 45857 (5114) \\
+  $M_{cov=1,b=0}$ & -2650.3 & 9 &  0.0 & 0.53 & 59312$^\dag$ (5231)  \\
+  $M_{cov=1,b=1}$ & -2649.4 & 10 & 0.3 & 0.47 & 49826 (10369) \\
 \hline
@@
 In (A), tan shading denotes land, unsurveyed grid cells appear in dark gray, and counts appear in a white-blue spectrum.
-Apparent bearded seal abundance estimates (B) are presented from the model with the lowest integrated AICc score, which included covariate effects but no preferential sampling effect.
+Apparent bearded seal abundance estimates (B) are presented from the model with the lowest integrated AIC score, which included covariate effects but no preferential sampling effect.
 Apparent abundance estimates are uncorrected for imperfect detection or species misclassification.}

### 2016-06-11 pref_sampling/R/util_funcs.R

@@ -129,7 +129,8 @@ plot_N_map<-function(cur.t=1,N,Grid,highlight=NULL,cell.width=1,leg.title="Abund
   require(ggplot2)
   require(rgeos)
   require(RColorBrewer)
-  if(is.null(myPalette)==TRUE)myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))  
+  #if(is.null(myPalette)==TRUE)myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))
+  if(is.null(myPalette)==TRUE)myPalette <- colorRampPalette(brewer.pal(9, "YlOrBr"))
   Tmp=Grid
   if(is.null(highlight)==FALSE){
     midpoints=data.frame(gCentroid(Tmp[highlight,],byid=TRUE))

### 2016-06-11 pref_sampling/inst/analyze_plot_bearded.R

@@ -1,0 +1,91 @@
+### analyze and visualize bearded seal model output
+library(ggplot2)
+library(grid)
+library(gridExtra)
+setwd('c:/users/paul.conn/git/pref_sampling')
+load("AlaskaBeringData2012_2013_14Dec2015.Rdat")  #from bass package on github/nmml
+source('./pref_sampling/R/util_funcs.R')
+setwd('c:/users/paul.conn/git/pref_sampling/Output')
+n.cells=nrow(Data$Adj)
+  
+N.table = matrix(0,6,4)
+colnames(N.table)=c("cov.mod","b.mod","N.est","SE")
+AIC.table = matrix(0,6,5)
+colnames(AIC.table)=c("cov.mod","b.mod","Likelihood","No.params","AICc")
+
+N.b.par=c(0,1,3)
+N.cov.par=c(0,3)
+
+N.s=matrix(0,n.cells,6)
+
+n.counter=1
+Plot=vector("list",1)
+for(icov in 1:2){
+  for(ib in 1:3){
+    fname=paste0("OutBearded_cov",icov,"_b",ib,".RData")
+    load(file=fname)
+    N.table[n.counter,1:2]=AIC.table[n.counter,1:2]=c(icov,ib)
+    N.table[n.counter,3]=Out$SD$unbiased$value["total_abundance"]
+    cur.pl=which(names(Out$SD$unbiased$value)=="total_abundance")
+    N.table[n.counter,4]=Out$SD$sd[cur.pl]
+    AIC.table[n.counter,3]=Out$log_lik
+    AIC.table[n.counter,4]=N.b.par[ib]+N.cov.par[icov]*2+3
+    AIC.table[n.counter,5]=-2*AIC.table[n.counter,3]+2*AIC.table[n.counter,4]
+    
+    N.s[,n.counter]=Out$SD$unbiased$value[which(names(Out$SD$unbiased$value)=="Z_s")]
+    Plot[[n.counter]]=plot_N_map(cur.t=1,N=matrix(N.s[,n.counter],ncol=1),Grid=Data$Grid$y2012[[1]])
+
+    n.counter=n.counter+1
+  }
+}
+
+#Make some plots to examine response surface for preferential sampling
+Grid.new=Data$Grid$y2012[[1]]
+n_cells=length(Grid.new)
+
[…]

### 2016-06-11 pref_sampling/inst/format_effort_bearded.R

@@ -1,0 +1,238 @@
+### format_effort.R
+### function to format effort and hotspot/count data for spatio-temporal abundance analysis
+
+
+require(sp)
+require(rgeos)
+require(rgdal)
+require(Matrix)
+require(lubridate)
+source('c:/users/paul.conn/git/BOSSst/BOSSst/R/util_funcs.R')
+
+date.start=as.Date("2012-04-10")
+date.end=as.Date("2012-04-16")
+t.steps=as.numeric(date.end-date.start)
+
+#load spatio-temporal covariate & grid data
+#read in Sp lines DF for transects ("on_effort_tracks.sldf") and Sp points Df for FMC points ("fmclogs.spdf")
+load('c:/users/paul.conn/git/STabundance/BOSS_2012Effort_22Apr14.Rdata')  
+load("c:/users/paul.conn/git/bass/AlaskaBeringData2012_17April2014_shelf_fixed.Rda")  #boss grid, ice data
+#rename IDs for grid cells to be 1:1299 instead of old IDs associated with 'full' Bering grid
+# (needed for some of the intersection stuff below)
+S=nrow(Data$Grid[[1]])
+for(it in 1:length(Data$Grid)){
+  for(ipoly in 1:S){
+    Data$Grid[[it]]@polygons[[ipoly]]@ID=as.character(ipoly)
+  }
+}
+
+laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
+                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
+#reproject onto BOSS Grid
+Tracks=spTransform(on_effort_tracks.sldf, CRS(laea_180_proj))
+Points=spTransform(fmclogs.spdf, CRS(laea_180_proj))
+Hotspots=spTransform(hotspots.spdf,CRS(laea_180_proj))
+Hotspots=Hotspots[which(Hotspots@data[,"species"]=="bd"),]  #limit to bearded seals
+
+Tracks$Day=rep(0,nrow(Tracks))
+Flt.ids=unique(Tracks$flightid)
+
+Flt.table=data.frame(id=Flt.ids,min.time=rep(0,length(Flt.ids)),day=rep(0,length(Flt.ids)))
+for(irow in 1:nrow(Flt.table)){
[…]

### 2016-06-11 pref_sampling/inst/run_bearded_models.R

@@ -88,7 +88,7 @@ Options_vec = c( 'Prior'=switch(Spatial_model,"ICAR"=1,"SPDE_GMRF"=0), 'Alpha'=A
 
 
 for(icov in 1:2){
-  for(ib in 1:2){
+  for(ib in 1:3){
     # Data
     Data = list("Options_vec"=Options_vec, "c_i"=c_i, "P_i"=Prop_sampled,"A_s"=1-Grid.new@data[,"land_cover"],"s_i"=s_i-1, "y_s"=y_s, "spde"=spde)
     if(icov==1)Data$X_sj=Data$X_sk=matrix(1,nrow=n_cells)
@@ -171,7 +171,7 @@ for(icov in 1:2){
     # SD
     #Report = Obj$report()
     SD=sdreport(Obj,bias.correct=TRUE)
-    Out=list(SD=SD,Report=Report)
+    Out=list(SD=SD,Report=Report,log_lik=-Opt$objective)
     fname=paste0("c:/users/paul.conn/git/pref_sampling/Output/OutBearded_cov",icov,"_b",ib,".RData")
     save(Out,file=fname)
   } 

### 2016-06-11 pref_sampling/inst/summarize_TMB_sims.R

@@ -10,7 +10,7 @@ load('Pref_sims_result4.Rda')
 Dat=rbind(Dat,Result)
 
 #remove 24 results with NAs occurring in fixed effect REs
-Dat=Dat[-which(Dat[,"NA.fixed"]>0),]
+#Dat=Dat[-which(Dat[,"NA.fixed"]>0),]
 #remove missinig sim records (original Results rows > # sims conducted)
 Dat=Dat[-which(Dat[,1]==0),]
 Dat=data.frame(Dat)

### 2016-06-24 manuscript/pref_sampling.tex
Restructured R files for package compilation and public consumption (including putting TMB .cpp code in the src directory for automatic compilation when package is loaded). Final changes to ms. before submission.

@@
 \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.
-Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$ \citep{DiggleEtAl2010}, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise.
+Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise.
 We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.
@@
   \label{eq:nu}
-  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i.
+  \boldsymbol{\nu} & = & \beta_0^* + {\bf x}^* \boldsymbol{\beta}^*+\boldsymbol{\eta}+{\bf B} \boldsymbol{\delta}.
 \label{eq:nu}
 \end{eqnarray}
- In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\eta_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.
+ In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).
 The predictive covariates ${\bf x}_i$ from Eq.
@@
 \ref{eq:nu} may or may not be the same.
-Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs.
+Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both Eqs.
 \ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.
@@
 \ref{fig:PS_sims}).
-Specific details of of data generation procedures are presented in Appendix S1.
+Specific details of data generation procedures are provided in Appendix S1.
 
@@
 
-Estimates of cumulative animal abundance across simulated landscapes were median unbiased for both estimation methods when the sites selected for sampling were independent of animal density, though when $b$ was estimated abundance estimates were more right skewed and had higher variance (fig.
+Estimates of cumulative animal abundance across simulated landscapes were median unbiased for both estimation methods when the sites selected for sampling were independent of animal density, though when $b$ was estimated, abundance estimates were more right skewed and had higher variance (fig.
 \ref{fig:sim_bias}).
@@
 
-Integrated AIC strongly favored models with covariate effects, but for such models the presence of PS was equivocal (Table \ref{tab:aic}).
+Marginal AIC strongly favored models with covariate effects, but for such models the presence of PS was equivocal (Table \ref{tab:aic}).
 Further intuition can be gained by examining estimates of the PS parameter, $b$.
@@
 Thus, there are theoretical reasons to sample more in high density areas than in low density areas.
-The obvious solution in this instance is to compensate for PS in model-based inferences by accounting for variation in sampling intensity with explanatory covariates or post hoc stratification.
-However, it is not always clear how to perform such stratification when effort is allocated in a subjective manner.
+The obvious solution in this instance is to account for variation in sampling intensity with explanatory covariates or post hoc stratification.
+However, it is not always clear how to perform post hoc stratification when effort is allocated in a subjective manner.
 
@@
 Accounting for PS in a model averaging framework led to a moderate decrease in our apparent abundance estimate for bearded seals in this region, and markedly decreased precision.
-As in our simulations, the need to account for PS thus appeared to have a real cost in terms variance inflation.
+As in our simulations, the need to account for PS thus appeared to have a real cost in terms of variance inflation.
 

### 2016-06-24 pref_sampling/R/RcppExports.R

@@ -1,19 +1,0 @@
-# This file was generated by Rcpp::compileAttributes
-# Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
-
-armaNorm <- function(n) {
-    .Call('PCspatreg_armaNorm', PACKAGE = 'PCspatreg', n)
-}
-
-GCN <- function(V, v) {
-    .Call('PCspatreg_GCN', PACKAGE = 'PCspatreg', V, v)
-}
-
-GCN2 <- function(V, v) {
-    .Call('PCspatreg_GCN2', PACKAGE = 'PCspatreg', V, v)
-}
-
-PCspatregMCMC <- function(y, X, K, sampleIndex, betaMean, betaSig, phiScale, burn, iter) {
-    .Call('PCspatreg_PCspatregMCMC', PACKAGE = 'PCspatreg', y, X, K, sampleIndex, betaMean, betaSig, phiScale, burn, iter)
-}
-

### 2016-06-24 pref_sampling/R/sim_data_abundance.R

@@ -1,192 +1,0 @@
-#' function to simulate abundance over a simulated landscape
-#' @param S Number of cells (must be a square number)
-#' @param n.covs Number of habitat covariates (default 4)
-#' @param tau.epsilon Precision of exchangeable errors in log space
-#' @return A SpatialPolygonsDataFrame holding covariates, abundance, and lambda (expected abundance)
-#' @export
-#' @keywords spatio-temporal, simulation
-#' @author Paul B. Conn
-sim_data_generic<-function(S,n.covs=4,tau.epsilon=100){
-  DEBUG=FALSE   
-  PLOT=FALSE
-  if(PLOT)set.seed(11111)
-  tau.epsilon=100
-  #if(DEBUG)tau.epsilon=100
-  
-  #source('./SpatPred/R/util_funcs.R')
-   
-  if(sqrt(S)%%1>0)cat("error in sim_data_generic; S must be a square number \n")
-  
-  x.len=sqrt(S)
- 
-  n.basis=10
-  n.zero=6
-  Omega=matrix(0,n.basis,S)
-  Cov=matrix(0,n.covs,S)
-  flag=1
-  while(flag==1){  #reject if any |cor|>0.75
-    for(i in 1:n.basis){
-      my.mod=RMexp(var=1,scale=runif(1,5,100))
-      Omega[i,]=RFsimulate(model=my.mod,x=c(1:x.len),y=c(1:x.len))$variable1
-    }
-    for(i in 1:n.covs){
-      Wts=matrix(rnorm(n.basis),1,n.basis)
-      Wts[sample(c(1:n.basis),n.zero)]=0
-      #cat(Wts)
-      #cat('\n')
-      Cov[i,]=Wts%*%Omega
-    }
-    Covs=t((Cov-rowMeans(Cov))/sqrt(apply(Cov,1,'var')))
-    cur.cor=max(abs(c(cor(Covs[,1],Covs[,2]),cor(Covs[,1],Covs[,3]),cor(Covs[,2],Covs[,3]))))
-    if(cur.cor<0.75)flag=0
-  }
-    
-  
-  if(PLOT==TRUE){
-    #XY=data.frame(x=Dat.matern$x,y=Dat.matern$y)
-    Grid1=data.frame(y=rep(1:30,each=30),x=rep(c(1:30),30),cov=as.vector(Covs[,1]))
-    Grid2=data.frame(y=rep(1:30,each=30),x=rep(c(1:30),30),cov=as.vector(Covs[,2]))
[…]

### 2016-06-24 pref_sampling/R/spat_pred.R

@@ -1,314 +1,0 @@
-#' function to perform Bayesian analysis of log Gaussian Cox processes and to make spatial predictions 
-#' @param formula A formula object specifying the model for the fixed effects
-#' @param Data   A SpatialPolygonsDataFrame holding covariate values for all cells one wants to make predictions on (including sampled cells)
-#' @param Effort A list including "Counts" which gives the response variable for surveyed cells, and "Mapping" which is a vector indicating which sample unit is associated with which count
-#' @param Area.adjust   A vector allowing for differences in suitable habitat for each cell.  Can be used for different grid cell sizes or different
-#'        proportions of suitable habitat (e.g., 1.0 = 100 percent of habitat is suitable for the focal species)
-#' @param Offset	A vector giving the fraction of each grid cell in Mapping that was surveyed
-#' @param spat.mod Controls whether (and what type) of spatial autocorrelation is implemented (spatmod = 0: no spatial autocorrelation; spatmod=1: process convolution; spatmod=2: srr ICAR) 
-#'        Note that if spatmod=1, the argument Knots needs to be specified; if spatmod=2, Adj must be specified
-#' @param Assoc   Association matrix for all grid cells in Data (needed for ICAR models only)
-#' @param K   A (# cells)x(# knots) matrix - the 'K' matrix for process convolution modeling.  
-#' @param Names.gam A character vector giving the column names of Data that will be modeled as smooth effects
-#' @param Knots.gam A list vector (length is # of smooth params), where each list element is a vector providing locations of knots (one location for each knot)
-#' @param Control A list giving MCMC controls and additional model options
-#'  "iter": number of MCMC iterations;
-#'  "burnin": number of MCMC burnin iterations;
[…]

### 2016-06-24 pref_sampling/R/util_funcs.R

@@ -1,27 +1,4 @@
-#' Produce an RW2 structure matrix for a line (e.g. for time series)
-#' @param x length of vector
-#' @return RW2 precision matrix
-#' @export 
-#' @keywords adjacency
-#' @author Paul Conn
-linear_adj_RW2 <- function(x){
-  Q=Matrix(0,x,x)
-  for(i in 1:(x-2)){
-    Q[i,i+2]=1
-    Q[i+2,i]=1
-  }
-  for(i in 1:(x-1)){
-    Q[i,i+1]=-4
-    Q[i+1,i]=-4
-  }
-  Q[1,2]=-2
-  Q[2,1]=-2
-  Q[x,x-1]=-2
-  Q[x-1,x]=-2
-  diag(Q)=-rowSums(Q)
-  Q
-}  
-
+ 
 #' Produce an RW1 adjacency matrix for a rectangular grid for use with areal spatial models (queens move)
 #' @param x number of cells on horizontal side of grid
 #' @param y number of cells on vertical side of grid
@@ -97,19 +74,6 @@ rect_adj <- function(x,y,byrow=FALSE){
   return(Adj)
 }
 
-#' Produce weights for bivariate normal kernel for upper left triangular quadrant
-#' To save computing time, some building blocks are passed into
-#' the function
-#' @param Tmp.vec A length n vector for holding integrals of bivariate normal
-#' @param XY 2xn matrix giving x and y distances
-#' @param Sigma vector of length 2 giving sd of bivariate normal in the x and y directions 
-#' @return a filled vector that holds unnormalized redistribution kernel values
-#' @export 
-#' @keywords bivariate normal, kernel weight
-#' @author Paul Conn \email{paul.conn@@noaa.gov}
-d_biv_normal<-function(Tmp.vec,XY,Sigma){
-  return(dnorm(XY[1,],0,Sigma[1])*dnorm(XY[2,],0,Sigma[2]))
-}
   
 #' Function to plot abundance map for Bering Sea survey grid
 #' @param cur.t  Time step to plot map for
@@ -126,9 +90,9 @@ d_biv_normal<-function(Tmp.vec,XY,Sigma){
 #' @keywords abundance map, plot
 #' @author Paul Conn \email{paul.conn@@noaa.gov}
[…]

### 2016-06-24 pref_sampling/demo/PCspatregDemo.R

@@ -1,47 +1,0 @@
-set.seed(123)
-library(sp)
-library(gstat)
-library(fields)
-library(PCspatreg)
-xy <- expand.grid(1:100, 1:100)
-names(xy) <- c('x','y')
-g.dummy <- gstat(formula=z~1, locations=~x+y, dummy=T, beta=0, model=vgm(psill=1, range=20, nugget=0.01, model='Gau'), nmax=20)
-simData <- predict(g.dummy, newdata=xy, nsim=4)
-simData$sim4 = simData$sim4-mean(simData$sim4)
-
-simData$resp = 3 + 2*simData$sim1 + 0.1*simData$sim3 + (simData$sim4 + rnorm(1000,0,1))
-
-idx = sort(sample(1:10000, 100, replace=FALSE))
-sampData = simData[idx,]
-
-knots = expand.grid(x=seq(-10, 110, 15), y=seq(-10, 110, 15))
-
-plot(knots$x, knots$y, asp=1, type="n")
-image(x=1:100, y=1:100, z=t(matrix(simData$resp,100,100)[100:1,]), asp=TRUE, add=TRUE)
-points(knots$x, knots$y)
-points(sampData$x, sampData$y, pch=16)
-
-K = zapsmall(exp(-0.5*rdist(simData[,c("x","y")], knots)^2/15^2)/(15*sqrt(2*pi)))
-K = sweep(K, 2, colSums(K),"/")/81
-X = as.matrix(cbind(1, simData[,3:5]))
-
-betaMean = c(mean(simData$resp[idx]), rep(0,3))
-betaSig=10*solve(crossprod(X[idx,]))
-
-
-postSamp = PCspatregMCMC(y=sampData$resp, X=X, K=K, sampleIndex=idx, betaMean=betaMean, betaSig=betaSig, 
-                         phiScale=10, burn=1000, iter=10000)
-
-library(coda)
-plot(mcmc(postSamp$beta))
-plot(mcmc(abs(postSamp$phi)))
-plot(mcmc(postSamp$sigma))
-
-
-
-simData$pred = colMeans(postSamp$pred)
-plot(knots$x, knots$y, asp=1, type="n")
-image(x=1:100, y=1:100, z=t(matrix(simData$pred,100,100)[100:1,]), asp=TRUE, add=TRUE)
-points(knots$x, knots$y)
-points(sampData$x, sampData$y, pch=16)
-

### 2016-06-24 pref_sampling/inst/analyze_plot_bearded.R

@@ -1,91 +1,0 @@
-### analyze and visualize bearded seal model output
-library(ggplot2)
-library(grid)
-library(gridExtra)
-setwd('c:/users/paul.conn/git/pref_sampling')
-load("AlaskaBeringData2012_2013_14Dec2015.Rdat")  #from bass package on github/nmml
-source('./pref_sampling/R/util_funcs.R')
-setwd('c:/users/paul.conn/git/pref_sampling/Output')
-n.cells=nrow(Data$Adj)
-  
-N.table = matrix(0,6,4)
-colnames(N.table)=c("cov.mod","b.mod","N.est","SE")
-AIC.table = matrix(0,6,5)
-colnames(AIC.table)=c("cov.mod","b.mod","Likelihood","No.params","AICc")
-
-N.b.par=c(0,1,3)
-N.cov.par=c(0,3)
-
-N.s=matrix(0,n.cells,6)
-
-n.counter=1
-Plot=vector("list",1)
-for(icov in 1:2){
-  for(ib in 1:3){
-    fname=paste0("OutBearded_cov",icov,"_b",ib,".RData")
-    load(file=fname)
-    N.table[n.counter,1:2]=AIC.table[n.counter,1:2]=c(icov,ib)
-    N.table[n.counter,3]=Out$SD$unbiased$value["total_abundance"]
-    cur.pl=which(names(Out$SD$unbiased$value)=="total_abundance")
-    N.table[n.counter,4]=Out$SD$sd[cur.pl]
-    AIC.table[n.counter,3]=Out$log_lik
-    AIC.table[n.counter,4]=N.b.par[ib]+N.cov.par[icov]*2+3
-    AIC.table[n.counter,5]=-2*AIC.table[n.counter,3]+2*AIC.table[n.counter,4]
-    
-    N.s[,n.counter]=Out$SD$unbiased$value[which(names(Out$SD$unbiased$value)=="Z_s")]
-    Plot[[n.counter]]=plot_N_map(cur.t=1,N=matrix(N.s[,n.counter],ncol=1),Grid=Data$Grid$y2012[[1]])
-
-    n.counter=n.counter+1
-  }
-}
-
-#Make some plots to examine response surface for preferential sampling
-Grid.new=Data$Grid$y2012[[1]]
-n_cells=length(Grid.new)
-
[…]

### 2016-06-24 pref_sampling/inst/format_effort_bearded.R

@@ -1,238 +1,0 @@
-### format_effort.R
-### function to format effort and hotspot/count data for spatio-temporal abundance analysis
-
-
-require(sp)
-require(rgeos)
-require(rgdal)
-require(Matrix)
-require(lubridate)
-source('c:/users/paul.conn/git/BOSSst/BOSSst/R/util_funcs.R')
-
-date.start=as.Date("2012-04-10")
-date.end=as.Date("2012-04-16")
-t.steps=as.numeric(date.end-date.start)
-
-#load spatio-temporal covariate & grid data
-#read in Sp lines DF for transects ("on_effort_tracks.sldf") and Sp points Df for FMC points ("fmclogs.spdf")
-load('c:/users/paul.conn/git/STabundance/BOSS_2012Effort_22Apr14.Rdata')  
-load("c:/users/paul.conn/git/bass/AlaskaBeringData2012_17April2014_shelf_fixed.Rda")  #boss grid, ice data
-#rename IDs for grid cells to be 1:1299 instead of old IDs associated with 'full' Bering grid
-# (needed for some of the intersection stuff below)
-S=nrow(Data$Grid[[1]])
-for(it in 1:length(Data$Grid)){
-  for(ipoly in 1:S){
-    Data$Grid[[it]]@polygons[[ipoly]]@ID=as.character(ipoly)
-  }
-}
-
-laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
-                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
-#reproject onto BOSS Grid
-Tracks=spTransform(on_effort_tracks.sldf, CRS(laea_180_proj))
-Points=spTransform(fmclogs.spdf, CRS(laea_180_proj))
-Hotspots=spTransform(hotspots.spdf,CRS(laea_180_proj))
-Hotspots=Hotspots[which(Hotspots@data[,"species"]=="bd"),]  #limit to bearded seals
-
-Tracks$Day=rep(0,nrow(Tracks))
-Flt.ids=unique(Tracks$flightid)
-
-Flt.table=data.frame(id=Flt.ids,min.time=rep(0,length(Flt.ids)),day=rep(0,length(Flt.ids)))
-for(irow in 1:nrow(Flt.table)){
[…]

### 2016-06-24 pref_sampling/inst/plot_sim_b.R

@@ -1,103 +1,0 @@
-#### plot expectations of probability of sampling vs. abundance residuals
-
-
-# Z -- true abundance
-# Y -- sampled abundance
-# R -- predicted probability of sampling
-# delta -- affects density
-# eta -- affects sampling probability
-
-library( RandomFields )
-library( INLA )
-
-# change directory
-TmbDir = "C:/Users/paul.conn/git/pref_sampling/TMB_version/inst/executables"
-setwd( TmbDir )
-
-# SOurce
-source( paste0(TmbDir,"/../../examples/rect_adj.R") )
-source( paste0(TmbDir,"/../../examples/rrw.R") )
-
-# Settings
-grid_dim = c("x"=25, "y"=25)
-n_samp = 50
-n_cells = grid_dim[1]*grid_dim[2]
-SpatialScale = sqrt(prod(grid_dim))/5  # Range ~ 2*Scale
-SD_eta = SD_x = SD_delta = 1
-beta0 = 2
-Alpha = 1  # Smoothness for GMRF, 1 or 2 (1 is faster)
-n_sim = 500
-
-# Configurations
-b_set = c(0,1,5)            # Impact of delta on sampling intensity
-Data=matrix(0,4,n_sim*n_cells)
-
-# Loop through replicates
-set.seed(12345)
-irow=1
-for(i in 1:n_sim){
-  #simulate abundance
-  # Settings
-
-  #note: betax set to zero in last commit by Jim (v4)
-  betax = runif(1,-.5,.5)       # Impact of x on density
-  betax_prob = runif(1,-.5,.5)   # Impact of x on sampling intensity
-  #betax=0
-  #betax_prob=0
-  
-  # Spatial model
-  loc_s = expand.grid( "x"=1:grid_dim['x'], "y"=1:grid_dim['y'])
-  model_delta <- RMgauss(var=SD_delta^2, scale=SpatialScale)
-  model_x <- RMgauss(var=SD_x^2, scale=SpatialScale)
-  model_eta <- RMgauss(var=SD_eta^2, scale=SpatialScale)
-  
-  # Realization from GRF
-  # delta_s -- spatial variation that affects both density and sampling intensity
-  # x_s -- spatial variation in 
[…]

### 2016-06-24 pref_sampling/inst/plot_sim_example.R

@@ -1,15 +1,9 @@
 ### plot preferential sampling plots using data from "run_sim_example.R"
-library(sp)
-library(maptools)
-library(rgeos)
-library(ggplot2)
-library(RColorBrewer)
-library(gridExtra)
-library(grid)
+library(PrefSampling)
 
-setwd('c:/users/paul.conn/git/pref_sampling')
-source('../SpatPred/SpatPred/R/util_funcs.R')
-load('./Sim_data/sim_plot_data.Rdata')
+#requires that the dataset sim_plot_data.RData be in the current working directory
+# this file can be created with "run_sim_example.R"
+load('./sim_plot_data.Rdata')
 
 N.true=matrix(Sim$N.true,ncol=1)
 N.1=matrix(Sim$N.1,ncol=1)
@@ -37,23 +31,19 @@ proj4string(Grid.SpP)=CRS(laea_180_proj)
 Grid=SpatialPolygonsDataFrame(Grid.SpP,data=data.frame(Cov=Cov),match.ID=FALSE)
 
 
-
-Grid.list=vector("list",1)
-Grid.list[[1]]=Grid
 max.N=100
-Ntrue.plot=plot_N_map(1,N.true,Grid=Grid.list)+ggtitle(expression(paste("C. True abundance (",N[i],")")))+theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),text=element_text(size=txt.size),plot.margin=unit(c(0,0,0,0),"lines"))+scale_fill_gradientn(colours=YlOrBr(100),limits=c(0,max.N))   
-N.plot.1=plot_N_map(1,N.1,Grid=Grid.list)+ggtitle(expression(paste("G. ",hat(N[i]),', b=0')))+theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),text=element_text(size=txt.size),plot.margin=unit(c(0,0,0,0),"lines"))+scale_fill_gradientn(colours=YlOrBr(100),limits=c(0,max.N))   
-N.plot.2=plot_N_map(1,N.2,Grid=Grid.list)+ggtitle(expression(paste("H. ",hat(N[i]),', b=1')))+theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),text=element_text(size=txt.size),plot.margin=unit(c(0,0,0,0),"lines"))+scale_fill_gradientn(colours=YlOrBr(100),limits=c(0,max.N))   
[…]

### 2016-06-24 pref_sampling/inst/run_TMB_sims.R

@@ -1,299 +1,0 @@
-# run_generic_sims.R
-# script to run generic spatio-temporal count data simulations
-# using preferential sampling model 
-# Y|R=1 ~ Pois( exp(Xy*betay + Nu + Delta + Epsilon) )
-# R ~ Bernoulli( exp(Xr*betar + Gamma + Delta) )
-
-#funtions and libraries for data generating model
-source("./pref_sampling/R/sim_data_abundance.R")
-source("./pref_sampling/R/util_funcs.R")
-source("./pref_sampling/R/spat_pred.R")
-library(spsurvey)
-library(rgeos)
-library(ggplot2)
-library(RColorBrewer)
-
-#stuff for estimation model (TMB)
-library( RandomFields )
-library( TMB )
-library( INLA )
-
-COMPILE=TRUE
-SIM=FALSE
-
-if(COMPILE){
-  Tmb.dir="./TMB_version/inst/executables/"
-  TmbFile = paste0(Tmb.dir,"pref_samp_v3.cpp")  #includes shared spatial model specification
-  compile(TmbFile )
-  TmbFile = paste0(Tmb.dir,"pref_samp_linear.cpp")  #linear predictor for R_i a linear function of linear predictor for Y_i
-  compile(TmbFile )
-}
-
-set.seed(123456)
-n.sims=125 #number of simulations at each design point
-S=900
-Adj=rect_adj(sqrt(S),sqrt(S))
-Q=-Adj
-diag(Q)=apply(Adj,2,'sum')+0.01  #the +1 is just for debugging to make it proper
-Q=Matrix(Q)
-
-#initialize Results data frame
-Result=matrix(0,n.sims*8*10,10)  #number of simulations * number of generating models * number of estimation models
-colnames(Result)=c("Gen.mod","Samp.intens","Covs","Est.mod","Sim.no","N.est","N.true","SE","MARE","NA.fixed")
-
-
-if(SIM){   #true abundance is independent of data generating model
-  for(isim in 1:n.sims){
-    #simulate covariates, abundance
-    Grid=sim_data_generic(S=S,n.covs=3,tau.epsilon=10)
-    Grid$cov1.quad=Grid$cov1^2
-    Grid$cov2.quad=Grid$cov2^2
-    Grid$cov3.quad=Grid$cov3^2
-    Grid$cov12=Grid$cov1*Grid$cov2
[…]

### 2016-06-24 pref_sampling/inst/run_bearded_models.R

@@ -1,41 +1,13 @@
 ### run simple preferential sampling example.  Doing in this in two steps since there seems to be
 # a conflict between TMB loop and rgeos ('gCentroid') functionality
 # need to declare an ADREPORT for Z_s in TMB for this
-if( FALSE ) setwd('c:/users/paul.conn/git/pref_sampling')
-if( Sys.info()['user']=="James.Thorson" ) setwd('C:/Users/James.Thorson/Desktop/Project_git/pref_sampling')
-source('./pref_sampling/R/util_funcs.R')
-library(sp)
-library(grid)
-library(maptools)
-library( RandomFields )
-library( TMB )
-library( INLA )
-library(rgeos)
-library(gridExtra)
-#load("Effort2012_bearded_PrefSamp_21Dec2015.Rdata")  
-load('Data_for_ST_plot.Rdata') #includes spatial objects associated w landscape
-load("Bearded_effort.Rda")  #formatted with BOSSst "format_Bering2012_all.R"
-load("AlaskaBeringData2012_2013_14Dec2015.Rdat")  #from bass package on github/nmml
+library(PrefSampling)
+
+data(AlaskaBeringData)
+data(Bearded_effort)
+data(Data_for_ST_plot)
+
 Effort=Bearded_effort
-#library( ThorsonUtilities )
-
-Version = "simple_v7"
-# v2 -- added option for changing distribution for random effects, and implemented ICAR distribution
-# v3 -- added Bernoulli model for sampling locations, and removed the ICAR functionality (by commenting it out, to save speed during building the object)
-# v4 -- added simpler less-smooth SPDE option (may be faster!), and option to turn off delta or eta
-# v5 -- Implements Devin's finite population correction approach to posterior prediction; proportion area sampled additional input vector  
-# v6 -- Includes option to output Z_s in ADREPORT so unbiased map estimates can be made using bias.correct
-# v7 -- Includes option to allows b to vary as a trend surface
-
-# Compile
-if( FALSE ) TmbFile = "C:/Users/paul.conn/git/pref_sampling/TMB_version/inst/executables"
[…]

### 2016-06-24 pref_sampling/inst/run_bearded_models_ice0.R

@@ -1,0 +1,199 @@
+### run simple preferential sampling example.  Doing in this in two steps since there seems to be
+# a conflict between TMB loop and rgeos ('gCentroid') functionality
+# need to declare an ADREPORT for Z_s in TMB for this
+library(PrefSampling)
+data('Data_for_ST_plot') #includes spatial objects associated w landscape
+data('Bearded_effort_ice0')  #formatted with BOSSst "format_Bering2012_all.R" ; includes several extra absences in southern ice-free area to prevent large estimates there
+data("AlaskaBeringData")  #from bass package on github/nmml
+Effort=Bearded_effort
+
+#create single Grid SpatialPolygonsDataFrame, averaging over covariate values
+Grid.new=Data$Grid$y2012[[1]]
+Days = c(6:12) #Apr 10-16
+Grid.new@data=0*Grid.new@data
+for(iday in Days)Grid.new@data=Grid.new@data+Data$Grid$y2012[[iday]]@data
+Grid.new@data=Grid.new@data/length(Days)
+
+#limit effort data to April 10-16
+Which.incl = which(Effort$Mapping[,2]<8)
+Mapping=Effort$Mapping[Which.incl,1]
+Prop_sampled=Effort$Area.trans[Which.incl]
+Prop_sampled[which(is.na(Prop_sampled))]=1.0
+Count.data=Effort$Count.data[Which.incl,]
+
+# Settings
+n_cells=length(Grid.new)
+n_samp = length(Mapping)
+#SpatialScale = sqrt(n_cells)/5  # Range ~ 2*Scale
+#SD_eta = SD_x = SD_delta = 1
+#beta0 = 2
+Use_REML = FALSE   # 
+#Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
+Spatial_model = c("SPDE_GMRF", "ICAR")[1]
+Alpha = 2  # Smoothness for GMRF, 1 or 2 (1 is faster)
+
+# Model configurations
+EM_set = c("fix_b","est_b")
+
+set.seed(12345)
+
+# massage into paper notation
+s_i = Mapping
+y_s = ifelse(1:n_cells %in% s_i, 1, 0)
[…]

### 2016-06-24 pref_sampling/inst/run_sim_example.R

@@ -1,30 +1,7 @@
 ### run simple preferential sampling example.  Doing in this in two steps since there seems to be
 # a conflict between TMB loop and rgeos ('gCentroid') functionality
 # need to declare an ADREPORT for Z_s in TMB for this
-setwd('c:/users/paul.conn/git/pref_sampling')
-source('../SpatPred/SpatPred/R/util_funcs.R')
-library(sp)
-library(maptools)
-library( RandomFields )
-library( TMB )
-library( INLA )
-#library( ThorsonUtilities )
-
-Version = "simple_v6"
-# v2 -- added option for changing distribution for random effects, and implemented ICAR distribution
-# v3 -- added Bernoulli model for sampling locations, and removed the ICAR functionality (by commenting it out, to save speed during building the object)
-# v4 -- added simpler less-smooth SPDE option (may be faster!), and option to turn off delta or eta
-# v5 -- Implements Devin's finite population correction approach to posterior prediction; proportion area sampled additional input vector  
-# v6 -- Includes option to output Z_s in ADREPORT so unbiased map estimates can be made using bias.correct
-
-# Compile
-TmbFile = "C:/Users/paul.conn/git/pref_sampling/TMB_version/inst/executables"
-setwd( TmbFile )
-compile( paste0(Version,".cpp") )
-
-# SOurce
-source( paste0(TmbFile,"/../../examples/rect_adj.R") )
-source( paste0(TmbFile,"/../../examples/rrw.R") )
+library(PrefSampling)
 
 # Settings
 grid_dim = c("x"=25, "y"=25)
@@ -117,24 +94,19 @@ for(SimI in 1:length(b_set)){
     
     # Data
     spde <- (inla.spde2.matern(mesh, alpha=2)$param.inla)[c("M0","M1","M2")]
-    if(Version%in%c("simple_v1")) Data = list( "c_i"=c_i, "s_i"=s_i-1, "X_sj"=Zy, "spde"=spde)
[…]

### 2016-06-24 pref_sampling/inst/run_simple_spatial.R

@@ -1,235 +1,0 @@
-#test basic spatial regression models in TMB
-
-#  1) ICAR: Y ~ Pois( exp(X*beta + Eta) )
-#  2) Thin plate spline formulation of above
-#  3) Geostatistical model
-
-#funtions and libraries for data generating model
-source("./pref_sampling/R/sim_data_abundance.R")
-source("./pref_sampling/R/util_funcs.R")
-source("./pref_sampling/R/spat_pred.R")
-library(spsurvey)
-library(rgeos)
-library(ggplot2)
-library(RColorBrewer)
-
-#stuff for estimation model (TMB)
-library( RandomFields )
-library( TMB )
-library( INLA )
-
-COMPILE=FALSE
-SIM=T
-prop.sampled=0.1
-n.transects=90
-
-
-if(COMPILE){
-  Tmb.dir="./TMB_version/inst/executables/"
-  TmbFile = paste0(Tmb.dir,"spatial_only.cpp")  #includes shared spatial model specification
-  compile(TmbFile )
-  TmbFile2 = paste0(Tmb.dir,"spatial_only_skaug.cpp")  #includes shared spatial model specification
-  compile(TmbFile2 )
-}
-
-set.seed(223456)
-n.sims=100 #number of simulations at each design point
-S=900
-Adj=rect_adj(sqrt(S),sqrt(S))
-Q=-Adj
-diag(Q)=apply(Adj,2,'sum')+0.01  #the +1 is just for debugging to make it proper
-Q=Matrix(Q)
-
-#initialize Results data frame
-Result=matrix(0,n.sims,11)  #number of simulations * number of generating models * number of estimation models
-colnames(Result)=c("Gen.mod","Samp.intens","Covs","Est.mod","Sim.no","N.est","N.true","SE","MARE","NA.fixed","N.est2")
-
-
-if(SIM){   #true abundance is independent of data generating model
-  for(isim in 1:n.sims){
-    #simulate covariates, abundance
-    Grid=sim_data_generic(S=S,n.covs=3,tau.epsilon=10000)
-    Grid$cov1.quad=Grid$cov1^2
-    Grid$cov2.quad=Grid$cov2^2
-    Grid$cov3.quad=Grid$cov3^2
[…]

### 2016-06-24 pref_sampling/inst/summarize_TMB_sims.R

@@ -1,22 +1,0 @@
-#Summarize TMB pref sampling sim results
-
-load('Pref_sims_result.Rda')
-Dat=Result
-load('Pref_sims_result2.Rda')
-Dat=rbind(Dat,Result)
-load('Pref_sims_result3.Rda')
-Dat=rbind(Dat,Result)
-load('Pref_sims_result4.Rda')
-Dat=rbind(Dat,Result)
-
-#remove 24 results with NAs occurring in fixed effect REs
-#Dat=Dat[-which(Dat[,"NA.fixed"]>0),]
-#remove missinig sim records (original Results rows > # sims conducted)
-Dat=Dat[-which(Dat[,1]==0),]
-Dat=data.frame(Dat)
-for(i in 1:4)Dat[,i]=as.factor(as.integer(Dat[,i]))
-
-Dat$Bias=(Dat[,"N.est"]-Dat[,"N.true"])/Dat[,"N.true"]
-
-library(doBy)
-summaryBy(Bias~Gen.mod+Samp.intens+Covs+Est.mod,data=Dat,FUN='mean')

### 2016-06-24 pref_sampling/src/PrefSampling.cpp

@@ -1,0 +1,261 @@
+// Space time 
+#include <TMB.hpp>
+//#include <atomic_math.hpp>  //for D_lgamma
+
+
+/** Precision matrix for the anisotropic case, eqn (20) in Lindgren et al. (2011) */    
+namespace R_inla_generalized {
+using namespace Eigen;
+using namespace tmbutils;
+using namespace R_inla;
+
+template<class Type>
+  SparseMatrix<Type> Q_spde_generalized(spde_t<Type> spde, Type kappa, int alpha=2){
+  Type kappa_pow2 = kappa*kappa;
+  Type kappa_pow4 = kappa_pow2*kappa_pow2;
+  	
+  if( alpha==1 ) return kappa_pow2*spde.M0 + spde.M1;
+  if( alpha==2 ) return kappa_pow4*spde.M0 + Type(2.0)*kappa_pow2*spde.M1 + spde.M2;
+}
+
+template<class Type>
+  SparseMatrix<Type> Q_spde_generalized(spde_aniso_t<Type> spde, Type kappa, matrix<Type> H, int alpha=2){
+
+  int i;
+  Type kappa_pow2 = kappa*kappa;
+  Type kappa_pow4 = kappa_pow2*kappa_pow2;
+  
+  int n_s = spde.n_s;
+  int n_tri = spde.n_tri;
+  vector<Type> Tri_Area = spde.Tri_Area;
+  matrix<Type> E0 = spde.E0;
+  matrix<Type> E1 = spde.E1;
+  matrix<Type> E2 = spde.E2;
+  matrix<int> TV = spde.TV;
+  SparseMatrix<Type> G0 = spde.G0;
+  SparseMatrix<Type> G0_inv = spde.G0_inv;
+	  	  
+  //Type H_trace = H(0,0)+H(1,1);
+  //Type H_det = H(0,0)*H(1,1)-H(0,1)*H(1,0);
+  SparseMatrix<Type> G1_aniso(n_s,n_s); 
+  SparseMatrix<Type> G2_aniso(n_s,n_s); 
+  // Calculate adjugate of H
+  matrix<Type> adj_H(2,2);
+  adj_H(0,0) = H(1,1);
+  adj_H(0,1) = -1 * H(0,1);
+  adj_H(1,0) = -1 * H(1,0);
+  adj_H(1,1) = H(0,0);
+  // Calculate new SPDE matrices
+
+  // Calculate G1 - pt. 1
[…]

### 2016-06-24 pref_sampling/src/RcppExports.cpp

@@ -1,78 +1,0 @@
-// This file was generated by Rcpp::compileAttributes
-// Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
-
-#include <RcppArmadillo.h>
-#include <Rcpp.h>
-
-using namespace Rcpp;
-
-// armaNorm
-arma::vec armaNorm(int n);
-RcppExport SEXP PCspatreg_armaNorm(SEXP nSEXP) {
-BEGIN_RCPP
-    SEXP __sexp_result;
-    {
-        Rcpp::RNGScope __rngScope;
-        Rcpp::traits::input_parameter< int >::type n(nSEXP );
-        arma::vec __result = armaNorm(n);
-        PROTECT(__sexp_result = Rcpp::wrap(__result));
-    }
-    UNPROTECT(1);
-    return __sexp_result;
-END_RCPP
-}
-// GCN
-arma::vec GCN(const arma::mat& V, const arma::vec& v);
-RcppExport SEXP PCspatreg_GCN(SEXP VSEXP, SEXP vSEXP) {
-BEGIN_RCPP
-    SEXP __sexp_result;
-    {
-        Rcpp::RNGScope __rngScope;
-        Rcpp::traits::input_parameter< const arma::mat& >::type V(VSEXP );
-        Rcpp::traits::input_parameter< const arma::vec& >::type v(vSEXP );
-        arma::vec __result = GCN(V, v);
-        PROTECT(__sexp_result = Rcpp::wrap(__result));
-    }
-    UNPROTECT(1);
-    return __sexp_result;
-END_RCPP
-}
-// GCN2
-arma::vec GCN2(const arma::mat& V, const arma::vec& v);
-RcppExport SEXP PCspatreg_GCN2(SEXP VSEXP, SEXP vSEXP) {
-BEGIN_RCPP
-    SEXP __sexp_result;
-    {
-        Rcpp::RNGScope __rngScope;
-        Rcpp::traits::input_parameter< const arma::mat& >::type V(VSEXP );
-        Rcpp::traits::input_parameter< const arma::vec& >::type v(vSEXP );
-        arma::vec __result = GCN2(V, v);
-        PROTECT(__sexp_result = Rcpp::wrap(__result));
-    }
-    UNPROTECT(1);
-    return __sexp_result;
-END_RCPP
-}
-// PCspatregMCMC
[…]

### 2016-06-24 pref_sampling/src/rawSourceCode.cpp

@@ -1,103 +1,0 @@
-#include "RcppArmadillo.h"
-// [[Rcpp::depends(RcppArmadillo)]]
-using namespace Rcpp;
-using namespace arma;
-
-// [[Rcpp::export]]
-arma::vec armaNorm(int n){
-  NumericVector x = rnorm(n,0,1);
-  arma::vec out(x.begin(), x.size(), false);
-  return out;
-}
-
-// [[Rcpp::export]]
-arma::vec GCN(const arma::mat& V, const arma::vec& v){
-  arma::mat out = solve(V,v) + solve(chol(V), armaNorm(v.n_elem));
-  return out;
-}
-
-// [[Rcpp::export]]
-arma::vec GCN2(const arma::mat& V, const arma::vec& v){
-  arma::mat out = solve(V,v) + solve(chol(V), arma::randn<arma::vec>(v.n_elem));
-  return out;
-}
-
-
-// [[Rcpp::export]]
-List PCspatregMCMC(const arma::vec& y, const arma::mat& X, const arma::mat& K, const arma::uvec& sampleIndex,
-const arma::vec& betaMean, const arma::mat& betaSig, const double& phiScale, const int& burn, const int& iter) {
-  int na = K.n_cols;
-  int nsite = X.n_rows;
-  int nb = X.n_cols;
-  int ns = y.n_elem;
-  arma::uvec idx = sampleIndex-1;
-  arma::mat Xsmp = X.rows(idx);
-  arma::mat Ksmp = K.rows(idx);
-  
-  arma::mat betaStor(iter, nb, fill::zeros);
-  arma::mat alphaStor(iter, na, fill::zeros);
-  arma::vec phiStor(iter, fill::zeros);
-  arma::vec sigStor(iter, fill::zeros);
-  arma::mat predStor(iter, nsite);
-  
-  // Beta items
-  arma::mat betaPrec = betaSig.i();
-  arma::mat V_beta_inv(nb, nb);
-  arma::vec v_beta(nb);
-  // Alpha items
-  arma::mat I(na,na, fill::eye);
-  arma::mat V_alpha_inv(na, na);
-  arma::vec v_alpha(na);
-  // phi items
-  arma::vec Kbar(ns);
-  double V_phi_inv;
-  double v_phi;
-  
-  // Initial values
[…]

### 2016-11-15 manuscript/ReviewerResponses.tex
Commit during ms. revisions.  Changed link function of Bernoulli sampling model to be logit and now include an intercept.  Simulation loop fails (crashes R) in TMB_version/examples/simple_example_v6.R

\documentclass[12pt,fleqn]{article}
%\documentclass[12pt,a4paper]{article}
\usepackage{amsmath,epsfig,epsf,psfrag,color,ecology,natbib}
%\usepackage{a4wide,amsmath,epsfig,epsf,psfrag}
\usepackage[nolists]{endfloat}
\bibliographystyle{ecology}

%  put your commands here

\def\be{{\ensuremath\mathbf{e}}}
\def\bx{{\ensuremath\mathbf{x}}}
\def\bftheta{{\ensuremath\boldsymbol{\theta}}}
\def\bfeta{{\ensuremath\boldsymbol{\eta}}}
\def\bfdelta{{\ensuremath\boldsymbol{\delta}}}
\def\bfSigma{{\ensuremath\boldsymbol{\Sigma}}}
\def\bfB{\textbf{B}}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}

\begin{document}

\bf \hspace{-.32in} Conn et al. responses to reviewer comments

%  make sure that the document has 25 lines per page (it is 12 pt)

\setlength{\textheight}{575pt} \setlength{\baselineskip}{23pt}

%  IMPORTANT -- place the \title, \author, etc, \begin{summary}...
%  \end{summary} \keywords statements in the order demonstrated below
\rm
Here, we provide line by line responses to associate editor and reviewer comments, suggestions, and edits.  For completeness, we include all reviewer comments in addition to our responses (our responses are in blue).  We have also included responses to additional comments we received from a quantitative ecologist (Mark K\'{e}ry) that we hope will improve presentation to non-statisticians.  In addition to these responses, we provide a marked up PDF version of our manuscript showing all changes made.

\section{Associate editor}

[From decision letter]
In view of the comments received, I must decline your manuscript for publication in the Methods in Ecology and Evolution at this time.  However, we would be willing to consider a new manuscript which takes into consideration these comments. In addition to the other comments, I would ask you to try to make the presentation more accessible to an ecological audience - as you put it in the context of SDMs, I think it should really be understandable by ecologists working with SDMs. I appreciate that this can be difficult.

\textcolor{blue}{
We thank the associate editor for the invitation to resubmit, and for their constructive criticism.  In our revision, we have made an effort to make the manuscript more accessible to quantitative ecologists and to those working in the SDM modeling field.  It is unfortunate that there seems to be a different set of traditions (and jargon) between those working in the scientific survey world (e.g., planned surveys of occupancy and abundance), and those working with presence-only data.  Admittedly, there are different challenges in each, but it can make cross-communication somewhat difficult.  In the attached we've tried to emphasize relationships between the two fields and to locate or work within the broader sphere of SDM modeling.}

To give one example, you wrote (p4, l78-80):
``For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high."

which could be re-written as

``For instance, PS can occur when the investigator tends to sample more in places where a species has already been observed, or is known to be abundant."

\textcolor{blue} {This formulation is much more to the point; we have accepted this change.}

[From the comments to author section]
Your paper definitely focuses on an important issue, and provides ways to address the problem. Besides the issues raised by the referee, I think you should relate your paper a bit more to the ecological literature, in particular on SDMs (see Edwards et al. 2006). I think in particular that in addition to the types of sampling you discuss, one could also consider sampling designs that are optimal to estimate response curve (eg Albert et al. 2010 - sorry for the self-citation...). Ecological studies may have different objectives, and I think you should try to make your paper relevant to as many readers as possible.

Albert, C. H., N. G. Yoccoz, T. C. Edwards, C. H. Graham, N. E. Zimmermann, and W. Thuiller. 2010. Sampling in ecology and evolution – bridging the gap between theory and practice. Ecography 33:1028-1037.
Edwards, T. C., D. R. Cutler, N. E. Zimmermann, L. Geiser, and G. G. Moisen. 2006. Effects of sample survey design on the accuracy of classification tree models in species distribution models. Ecological Modelling 199:132-141.

\textcolor{blue}{We thank the editor for pointing us to these articles - our knowledge of the species distribution modeling literature is admittedly a little lacking.  These papers are certainly germane to the present study, and we have included additional material in the introduction (2nd paragraph) that talks about (1) implications of design vs. model-based inference, and (2) the importance of the particular objective function (e.g., minimizing variance, estimating species-habitat relationships) for optimal design.}

\section{Reviewer 1}

{\bf A generalized preferential sampling model}

Your random effects model doesn’t appear similar to \citet{RoyleBerliner1999} to me at all, and in fact this suggestion led to some confusion on my part about how these random effects were specified. The joint
specification of $\bfeta$ and $\bfdelta$, following \citet{RoyleBerliner1999} and assuming 0 mean, is:
\begin{equation}
  \left(
    \begin{array}{c}
    \bfeta \\
    \bfdelta
    \end{array}
   \right)
   \sim \textrm{Gaussian}
   \left(
     \left(
       \begin{array}{c}
       0 \\
       0
       \end{array}
     \right),
     \left(
       \begin{array}{cc}
         \bfSigma_\eta + \bfB \bfSigma_\delta \bfB^\prime & \bfB \bfSigma_\delta \\
         (\bfB \bfSigma_\delta)^\prime & \bfSigma_\delta
       \end{array}
     \right)
   \right),
\end{equation}
which has a very different covariance structure than the random effects model presented in Appendix S1.

\textcolor{blue}{There are several points to make here.  First, we never claimed that we were fitting the exact model proposed by \citet{RoyleBerliner1999} - we're not.  However, the covariance structure implied by the two model \emph{is} actually very similar.  For instance, if $\bfeta \sim \text{Gaussian}({\bf 0},\bfSigma_\eta)$ and $\bfdelta \sim \text{Gaussian}({\bf 0},\bfSigma_\delta)$, then
\begin{equation}
  \text{Cov}\left(
    \begin{array}{c}
    \bfeta + \bfB \bfdelta \\
    \bfdelta
    \end{array}
   \right)
   =
   \left(
   \begin{array}{cc}
      \bfSigma_\eta + \bfB \bfSigma_\delta \bfB^\prime & \bfB \bfSigma_\delta \\
         (\bfB \bfSigma_\delta)^\prime & \bfSigma_\delta
    \end{array}
   \right),
\end{equation}
as desired.  Second, it seems less important that the model be exactly the same as it does to provide a motivation (and citation) for how the $\bfB$ matrix can be specified.  We think both of these points were adequately accounted for in our initial submission, as we (a) described the \citet{RoyleBerliner1999} model as being ``similar" to the one we proposed here, and (b) used their paper as a reference for how the $\bfB$ matrix can be formulated and interpreted.
}

In your application, line 199, you report difficulties fitting a model with even modest structure on $\bfB$. Might a
joint model for $\bfeta$ and $\bfdelta$, as specified by \citet{RoyleBerliner1999}, allow you to specify a more complicated
model for $\bfB$?

\textcolor{blue}{
  As suggested above, we believe our model to be very similar to that proposed by \citet{RoyleBerliner1999}.  We're not seeing how drawing the random effects jointly would necessarily improve performance (we're also having trouble seeing how we could implement this in our chosen software).  Also, after further investigation of the \citet{RoyleBerliner1999} paper, we're not sure they ever actually fit a model with a more complicated relationship in $\bfB$.  In fact, their one example on ozone-temperature analysis uses the same simple diagonal parameterization we used here.  So we think the question remains whether more complicated models for $\bfB$ are ever identifiable, and if so, whether they are reasonable to fit with data sets typical of ecological applications.  We've adjusted text in several places to reflect this. For instance, we state that ``However, to our knowledge the identifiability of higher order models (such as trend surfaces) has not been investigated" when talking about higher order preferential sampling models.  In the discussion, we indicate that ``We attempted to fit models to bearded seal data where the degree of PS changed over the landscape, as \citet{RoyleBerliner1999} suggested might be possible (but never demonstrated) for multivariate spatial models."  We also suggest that identifiability of such models be further investigated in future research.
}

On line 139, you suggest “we often need to fix $\beta_0^* = 0.0$ . . . to permit parameter identification”. The word
``often" suggests you can sometimes include the intercept parameter when estimating random effects. Can
you provide more guidance about when you would need to fix the intercept at 0, and how a practitioner
would know they need to fit the intercept at 0?

\textcolor{blue}{
  This wording was too ambiguous, and was likely a result of initially fitting different types of spatial models to the data.  For instance, ICAR type models have a known parameter redundancy built in, such that the intercept is confounded with the spatial process \citep{RueHeld2005}.  The type of Gaussian process model we settled on and used in analysis does not suffer from this deficiency.  Further confusion resulted from incorrectly specifying a multinomial logit link function in our initial analyses.  We have rectified this mistake in the present version, and all models for $R_i$ are now Bernoulli with a logit link, so that the intercept term is now identifiable.  We now report results for this corrected analysis.
}

{\bf Notation}

I have difficulty following your notation at several points throughout the manuscript, which are outlined
below.

On line 41, you state $Z_i$ is the state variable of interest (e.g., occupancy or abundance), which I take will
be the notation you use throughout given the section header. Your specication for a count model on eqn
4 therefore becomes confusing. In particular, if $Z_i$ represents abundance, for reasonable values of $A_i$ and $p_i$ you could have random realizations $Y_i$ from a Poisson distribution that are greater than abundance, $Z_i$. If $Z_i$ does not represent abundance, what state process is represented here?

\textcolor{blue}{
  We have replaced the Poisson observation model with a binomial one in eqn 4 to make this clearer, and have indicated dependence on $Z_i$ through conditioning.  The reason that we had the Poisson in there originally is that the models we fit to data use the fact that the marginal distribution for $Y_i$ given the Poisson-Binomial mixture is itself a Poisson distribution.  However, we think it will be clearer to present the model hierarchically at this stage and then to describe our marginalization (something many ecologists may not care about) when describing software.  We have created a new ``Software" methods subsection which describes this in greater detail. Importantly, we describe how predictions are made; when detection probability is equal to 1.0, predictions are made using the formulation $\hat{N}_i = Y_i + (1-A_i)\exp(\hat{\mu}_i)$.  This framework incorporates a finite population correction and is similar to that employed by \citet{JohnsonEtAl2010} in the context of point process likelihood analysis of distance sampling data.  It also prevents predicted abundance in a cell from being less than the observed count.
}

\section{Additional reviewer: Marc K\'{e}ry}

I read your ms and found it very stimulating and also well-written. Just a couple of minor comments which I noted during my reading.
-          P.3, eq. 4: why do you put a Poisson observation model here ? In particular, since the Poisson has no “ceiling”, p cannot be detection probability but must be a detection rate, i.e., the combination of false-positive and false-negative errors. I note that in the Appendix 1 you have the usual Binomial observation model

\textcolor{blue}{This was also noted by Reviewer 1. There are two alternatives here.  In the first, one can use some sort of simulation-based inference (e.g. MCMC) to model both components of the hierarchy (i.e. $[{\bf Z}|\bftheta]$ and $[{\bf Y}|{\bf Z}]$), in which case one could specify $[{\bf Z}|\bftheta]$ to be Poisson and $[{\bf Y}|{\bf Z},p]$ to be binomial.  Keeping track of the latent $Z_i$s, one then has an entirely coherent model for abundance.  However, from a computational standpoint, it is much easier to deal with the unconditional distribution $[{\bf Y}|\bftheta,p]$.  If $\bftheta$ are the expectations of ${\bf Z}$, then unconditionally $[{\bf Y}|\bftheta,p] = \text{Poisson}(\bftheta p)$ (apologies for a little sloppiness in notation here).   From a computational standpoint, the Poisson distribution is much easier to deal with, as one can bring the TMB machinery into the analysis (in particular, since it works with derivatives, it can't handle discrete random effects). The inference about the parameters $\bftheta$ should be completely coherent. The disadvantage is that one must then predict the latent ${\bf Z}$ parameters: e.g. once we have conducted inference relative to $\bftheta$, one must then predict $Z_i \sim \text{Poisson}(\theta_i)$. This isn't perfect, as we could potentially have the case that the predicted $Z_i$ is actually less that the count $Y_i$.  However, it should be a reasonable approximation, especially when focus is on absolute abundance (i.e. $\sum Y_i$) and will actually be a good approximation at the individual cell level whenever $Z_i$ is large and $p$ is small (e.g., when only a small fraction of each survey unit is sampled, as in the bearded seal example).  We note that a) this strategy of inference has been used in several recent applications with good results (e.g. \cite{ThorsonEtAl}, and b) the simulation results suggest this procedure is unbiased (see Fig. 4, Independent estimation model, $b=0$ case).  Note that in simulations, we did generate data as $[{\bf Z}] = \text{Poisson}$ and $[{\bf Y} | {\bf Z}] = \text{Binomial}$.  We now discuss this tradeoff more explicitly, both in Appendix S1 and in the Discussion.
}

Line 68: What are platforms of opportunity ? perhaps sea mammal jargon ? Hard to understand for non-sea-mammal guys like myself

\textcolor{blue}{This is a good point - it is actually marine mammal jargon (for example, U.S. military ships providing locations where whale observations are made opportunistically).  We deleted this term, replacing it with ``opportunistic sampling."}

-          Line 139: doesn’t setting beta-star-0 to zero imply that the probability of a sample unit to be surveyed is fixed at 1 for the average environmental conditions (if covariates are centred) ?
-          168: lest
-          199: I would say “and estimates are not reported here”
-          209 and 211: fig. instead of Fig
-          317: typo
-          Fig. 1: main title left: Coarse scale
-          Fig. 4, legend: since you’re doing ML you may briefly comment on why you have a posterior mode here ? Presumably some application of Bayes rule for empirical Bayes estimates ?

\bibliography{master_bib}

\end{document}

### 2016-11-15 manuscript/pref_sampling.tex
Commit during ms. revisions.  Changed link function of Bernoulli sampling model to be logit and now include an intercept.  Simulation loop fails (crashes R) in TMB_version/examples/simple_example_v6.R

@@
 
-\def\volumeyear{2015}
+\def\volumeyear{2016}
 \def\VOC{VO}
@@
 \item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g.
-$40\%$) bias in estimates of animal density, and that our modeling approach can considerably reduce bias.
+$40\%$) bias in estimates of animal density, and that our modeling approach can considerably reduce this bias.
 
@@
 
-\item When wildlife surveys are conducted without a well defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling.
+\item When wildlife surveys are conducted without a well-defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling.
 Joint models, such as those described in this paper, can be used to test and correct for such biases.
@@
 Designing animal surveys is relatively straightforward in such applications, and unbiased point and variance estimators are available.
-Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture and Dail-Madsen models for repeated point counts \citep{Royle2004a,DailMadsen2011}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,MillerEtAl2013,JohnsonEtAl2010}.
+Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture and Dail-Madsen models for repeated point counts \citep{Royle2004a,DailMadsen2011}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,JohnsonEtAl2010,MillerEtAl2013}.
 In such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g.
 via trend surfaces or spatial random effects) to predict density or distributions across the landscape.
-We shall refer to the amalgam of model-based approaches for making spatially explicit inference about animal populations as ``Species distribution models" \citep[SDMs; {\it sensu}][]{ElithLeathwick2009}, even though this term is more often used to refer to animal occurrence than it is to density or abundance.
+We shall refer to the amalgam of model-based approaches for making spatially explicit inference about animal populations as ``species distribution models" \citep[SDMs; {\it sensu}][]{ElithLeathwick2009}, even though this term is more often used to refer to animal occurrence than it is to density or abundance.
 
@@
 in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.
-In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance, regardless of predictive covariates.
+In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance.
 For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.
@@
 For instance, when allocating line transect survey effort, it may be tempting to place the transect in a manner that targets habitat or landscape features that maximize the number of animals that will be encountered.
-Depending upon the interpretation of occupancy, this may or may not be reasonable.
-However, if trying to estimate density or abundance, this strategy will clearly lead to positive bias.
+However, this strategy will clearly lead to positive bias when estimating density or abundance.
 
@@
 Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.
-Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling.
+Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurrence) and the locations chosen for sampling.
 For the abundance process model, we start with eq.
@@
 We used maximum likelihood to conduct statistical inference.
-In particular, we used Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}, interfaced with the R programming environment, to conduct maximization.
+In particular, we used Template Model Builder \citep[TMB; ][]{KristensenEtAl2016}, interfaced with the R programming environment, to conduct maximization.
 The TMB software uses a Laplace approximation to integrate out random effects ($\bfeta$ and $\bfdelta$), and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
@@
 
-In this study, we showed that course-grained preferential sampling (Fig.
+In this study, we showed that coarse-grained preferential sampling (Fig.
 \ref{fig:pref}) can have a profound impact on the quality of estimates (e.g.
@@
 abundance or occupancy).
-However, such models can be considerably less precise and have greater instability than models without a preferential sampling parameter and it would be preferable not to have to go to such elaborations.
+However, such models can be considerably less precise and have greater instability than models without a preferential sampling parameter.
 Where possible, we suggest that survey planners incorporate design-based elements (e.g.

### 2016-11-15 TMB_version/examples/simple_example_v6.R

@@ -1,0 +1,267 @@
+
+# Z -- true abundance
+# Y -- sampled abundance
+# R -- predicted probability of sampling
+# delta -- affects density
+# eta -- affects sampling probability
+
+library( RandomFields )
+library( TMB )
+library( INLA )
+library(PrefSampling)
+
+#library( ThorsonUtilities )
+
+Version = "simple_v6"
+# v2 -- added option for changing distribution for random effects, and implemented ICAR distribution
+# v3 -- added Bernoulli model for sampling locations, and removed the ICAR functionality (by commenting it out, to save speed during building the object)
+# v4 -- added simpler less-smooth SPDE option (may be faster!), and option to turn off delta or eta
+# v5 -- Implements Devin's finite population correction approach to posterior prediction; proportion area sampled additional input vector  
+# v6 -- Fixes link function for R_i; now logit() instead of multinomial-logit()
+
+# Compile
+#TmbFile = "C:/Users/paul.conn/git/pref_sampling/pref_sampling/src/PrefSampling"
+#compile(paste0(TmbFile,".cpp")) 
+
+
+# Settings
+grid_dim = c("x"=25, "y"=25)
+n_samp = 50
+n_cells = grid_dim[1]*grid_dim[2]
+prop_sampled=0.5
+Prop_sampled=rep(prop_sampled,n_samp)
+SpatialScale = sqrt(prod(grid_dim))/5  # Range ~ 2*Scale
+SD_eta = SD_x = SD_delta = 1
+beta0 = 2
+Use_REML = FALSE   # 
+Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
+Spatial_model = c("SPDE_GMRF", "ICAR")[1]
+Alpha = 1  # Smoothness for GMRF, 1 or 2 (1 is faster)
+RandomSeed = 123456
+n_sim = 500
+
+# Configurations
+b_set = c(0,1,5)            # Impact of delta on sampling intensity
+EM_set = c("fix_b","est_b")
+
+# Configurations
+
+# Source ICAR precision matrix function
[…]

### 2016-11-15 pref_sampling/inst/run_bearded_models_ice0.R

@@ -67,7 +67,7 @@ for(icov in 1:2){
     if(icov==1)Data$X_sj=Data$X_sk=matrix(1,nrow=n_cells)
     if(icov>1){
       Data$X_sj=cbind(1,x_s_z)
-      Data$X_sk=cbind(x_s_z)
+      Data$X_sk=cbind(1,x_s_z)
     }   
     if(ib<3)Data$X_sb=matrix(1,nrow=n_cells)
     if(ib==3)Data$X_sb=cbind(x_s_b)
@@ -135,8 +135,8 @@ for(icov in 1:2){
       }
       Data$Options_vec[Which+2] = 0
       # Re-run
-      if( length(Which)!=2 ) Obj = MakeADFun( data=Data, parameters=Params, random=Random, map=Map, silent=TRUE)
-      if( length(Which)==2 ) Obj = MakeADFun( data=Data, parameters=Params, random=NULL, map=Map, silent=TRUE)
+      if( length(Which)!=2 ) Obj = MakeADFun( data=Data, parameters=Params, random=Random, map=Map, silent=TRUE,DLL="PrefSampling")
+      if( length(Which)==2 ) Obj = MakeADFun( data=Data, parameters=Params, random=NULL, map=Map, silent=TRUE,DLL="PrefSampling")
       Opt = nlminb( start=Obj$par, objective=Obj$fn, gradient=Obj$gr, lower=Lower, upper=Upper, control=list(trace=1, maxit=1000))         #
       Opt[["diagnostics"]] = data.frame( "Param"=names(Obj$par), "Lower"=-Inf, "Est"=Opt$par, "Upper"=Inf, "gradient"=Obj$gr(Opt$par) )
     }
@@ -144,7 +144,7 @@ for(icov in 1:2){
     # SD
     #Report = Obj$report()
     SD=sdreport(Obj,bias.correct=TRUE)
-    Out=list(SD=SD,Report=Report)
+    Out=list(Opt=Opt,SD=SD,Report=Report)
     fname=paste0("./OutBearded_cov",icov,"_b",ib,"ice0.RData")
     save(Out,file=fname)
   } 
@@ -152,14 +152,14 @@ for(icov in 1:2){
 
 #plot map of estimates
 str1='./OutBearded_cov'
-load(paste0(str1,2,"_b",1,"ice0.Rdata"))
[…]

### 2016-11-15 pref_sampling/inst/run_sim_example.R

@@ -24,11 +24,6 @@ EM_set = c("fix_b","est_b")
 
 # Configurations
 
-# Source ICAR precision matrix function
-Q = rect_adj( x=grid_dim['x'], y=grid_dim['y'] )
-Q = -1*Q
-diag(Q) = -1 * colSums(Q)
-
 set.seed(12345)
 
 Counts=Est=vector("list",3)
@@ -157,8 +152,8 @@ for(SimI in 1:length(b_set)){
       }
       Data$Options_vec[Which+2] = 0
       # Re-run
-      if( length(Which)!=2 ) Obj = MakeADFun( data=Data, parameters=Params, random=Random, map=Map, silent=TRUE)
-      if( length(Which)==2 ) Obj = MakeADFun( data=Data, parameters=Params, random=NULL, map=Map, silent=TRUE)
+      if( length(Which)!=2 ) Obj = MakeADFun( data=Data, parameters=Params, random=Random, map=Map, silent=TRUE,DLL="PrefSampling")
+      if( length(Which)==2 ) Obj = MakeADFun( data=Data, parameters=Params, random=NULL, map=Map, silent=TRUE,DLL="PrefSampling")
       Opt = nlminb( start=Obj$par, objective=Obj$fn, gradient=Obj$gr, lower=Lower, upper=Upper, control=list(trace=1, maxit=1000))         #
       Opt[["diagnostics"]] = data.frame( "Param"=names(Obj$par), "Lower"=-Inf, "Est"=Opt$par, "Upper"=Inf, "gradient"=Obj$gr(Opt$par) )
     }

### 2016-11-15 pref_sampling/src/PrefSampling.cpp

@@ -221,9 +221,9 @@ Type objective_function<Type>::operator() ()
   vector<Type> linpredR_s = X_sk * beta_k;
   vector<Type> b_s = X_sb * beta_b;
   for(int s=0; s<n_s; s++){
-    R_s(s) = exp( linpredR_s(s) + b_s(s)*delta_s(s) + eta_s(s) );
+    R_s(s) = 1 / (1 + exp(-linpredR_s(s) - b_s(s)*delta_s(s) - eta_s(s) ));
   }
-  R_s = R_s / R_s.sum();
+
   
   // Probability of sampling locations
   for(int s=0; s<n_s; s++){

### 2016-11-18 manuscript/ReviewerResponses.tex
Commit during ms. revisions and change to link function; simulations now running with some error trapping and bounds set on parameter space.  Still getting some negative estimates using sdreport with bias.correct=TRUE when b is estimated.

@@
 \def\bfdelta{{\ensuremath\boldsymbol{\delta}}}
+\def\bfnu{{\ensuremath\boldsymbol{\nu}}}
+\def\bfmu{{\ensuremath\boldsymbol{\mu}}}
 \def\bfSigma{{\ensuremath\boldsymbol{\Sigma}}}
@@
 For completeness, we include all reviewer comments in addition to our responses (our responses are in blue).
-We have also included responses to additional comments we received from a quantitative ecologist (Mark K\'{e}ry) that we hope will improve presentation to non-statisticians.
-In addition to these responses, we provide a marked up PDF version of our manuscript showing all changes made.
+In addition, our revision includes copy editing corrections from our editorial office and some clarifications in response to additional comments we received from an interested reader.
+We provide a marked up PDF version of our manuscript showing all changes made.
 
@@
 In our revision, we have made an effort to make the manuscript more accessible to quantitative ecologists and to those working in the SDM modeling field.
-It is unfortunate that there seems to be a different set of traditions (and jargon) between those working in the scientific survey world (e.g., planned surveys of occupancy and abundance), and those working with presence-only data.
-Admittedly, there are different challenges in each, but it can make cross-communication somewhat difficult.
-In the attached we've tried to emphasize relationships between the two fields and to locate or work within the broader sphere of SDM modeling.}
+We provide numerous examples below.}
 
@@
 model-based inference, and (2) the importance of the particular objective function (e.g., minimizing variance, estimating species-habitat relationships) for optimal design.}
+
+Additionally, we have made the following changes to try to make the manuscript accessible to a wider audience:
+\begin{itemize}
+  \item 
+\end{itemize}
 
@@
 
-\section{Additional reviewer: Marc K\'{e}ry}
+On line 61, you indicate $\bfmu$ and $\bfnu$ represent log-scale abundance (associated with {\bf Z}) and the logit of the probability of sampling (associated with {\bf R}).
+For equation 5, however, you use $\bfmu_i$ to represent the mean value of a Gaussian process, which created some confusion.
+You might consider more generally allowing $\bfmu$ to be a function of covariates influencing the state variable of interest, and $\bfnu$ to be a function of covariates
+influencing placement of sampling locations.
+You could then replace $\psi_i$ in equation 6 with $\nu_i$, which would make it immediately clear you were discussing factors influencing the probability of site placement.
 
-I read your ms and found it very stimulating and also well-written. Just a couple of minor comments which I noted during my reading.
--          P.3, eq.
-4: why do you put a Poisson observation model here ?
-In particular, since the Poisson has no “ceiling”, p cannot be detection probability but must be a detection rate, i.e., the combination of false-positive and false-negative errors.
-I note that in the Appendix 1 you have the usual Binomial observation model
-
-\textcolor{blue}{This was also noted by Reviewer 1.
-There are two alternatives here.
-In the first, one can use some sort of simulation-based inference (e.g.
-MCMC) to model both components of the hierarchy (i.e.
-$[{\bf Z}|\bftheta]$ and $[{\bf Y}|{\bf Z}]$), in which case one could specify $[{\bf Z}|\bftheta]$ to be Poisson and $[{\bf Y}|{\bf Z},p]$ to be binomial.
-Keeping track of the latent $Z_i$s, one then has an entirely coherent model for abundance.
-However, from a computational standpoint, it is much easier to deal with the unconditional distribution $[{\bf Y}|\bftheta,p]$.
-If $\bftheta$ are the expectations of ${\bf Z}$, then unconditionally $[{\bf Y}|\bftheta,p] = \text{Poisson}(\bftheta p)$ (apologies for a little sloppiness in notation here).
-From a computational standpoint, the Poisson distribution is much easier to deal with, as one can bring the TMB machinery into the analysis (in particular, since it works with derivatives, it can't handle discrete random effects).
-The inference about the parameters $\bftheta$ should be completely coherent.
-The disadvantage is that one must then predict the latent ${\bf Z}$ parameters: e.g.
-once we have conducted inference relative to $\bftheta$, one must then predict $Z_i \sim \text{Poisson}(\theta_i)$.
-This isn't perfect, as we could potentially have the case that the predicted $Z_i$ is actually less that the count $Y_i$.
-However, it should be a reasonable approximation, especially when focus is on absolute abundance (i.e.
-$\sum Y_i$) and will actually be a good approximation at the individual cell level whenever $Z_i$ is large and $p$ is small (e.g., when only a small fraction of each survey unit is sampled, as in the bearded seal example).
-We note that a) this strategy of inference has been used in several recent applications with good results (e.g.
-\cite{ThorsonEtAl}, and b) the simulation results suggest this procedure is unbiased (see Fig.
-4, Independent estimation model, $b=0$ case).
-Note that in simulations, we did generate data as $[{\bf Z}] = \text{Poisson}$ and $[{\bf Y} | {\bf Z}] = \text{Binomial}$.
-We now discuss this tradeoff more explicitly, both in Appendix S1 and in the Discussion.
+\textcolor{blue}{
+  We appreciate the reviewer's attempt to help us streamline notation and to make things clearer for the reader.
+We have implemented the basic premise of this idea, although some of the particulars differ from the reviewer's suggestion.
+Specifically we now describe $\bfmu$ and $\bfnu$ as indicating ``the link-scale process of interest and the logit of the probability of sampling, respectively.
+As with $\bfmu$, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects."  Then in the next section we use $\mu_i$ in eq.
+5 and $\nu_i$ in eq.
+6, although we retain $xi_i$, since in the next section we also wish to write $\nu_i$ in terms of shared spatial effects.
+This is all a little tricky, in that the \citet{DiggleEtAl2010} paper considers a continuous support model (whereas ours is discrete), and the structure is slightly different.
+Admittedly, maintaining recognizable notation while keeping precise definitions is a challenge.
 }
 
-Line 68: What are platforms of opportunity ? perhaps sea mammal jargon ? Hard to understand for non-sea-mammal guys like myself
+I understand the model you are specifying under the ``simulation study" subheading, but the notation is
+different from that presented in the rest of the paper (i.e., use of ``=" vs. ``$\sim$"). I have no preference for
+the notation used, but it seems like consistent notation would be more reader friendly. The same comment
+holds true for the model presented under the subheading ``bearded seal count surveys".
 
-\textcolor{blue}{This is a good point - it is actually marine mammal jargon (for example, U.S.
-military ships providing locations where whale observations are made opportunistically).
-We deleted this term, replacing it with ``opportunistic sampling."}
+\textcolor{blue}{
+  We agree.  We've changed all distributions to brackets and have eliminated the ``$\sim$" notation.
+}
 
--          Line 139: doesn’t setting beta-star-0 to zero imply that the probability of a sample unit to be surveyed is fixed at 1 for the average environmental conditions (if covariates are centred) ?
--          168: lest
--          199: I would say “and estimates are not reported here”
--          209 and 211: fig. instead of Fig
--          317: typo
--          Fig. 1: main title left: Coarse scale
--          Fig. 4, legend: since you’re doing ML you may briefly comment on why you have a posterior mode here ? Presumably some application of Bayes rule for empirical Bayes estimates ?
+Finally, your definition of $Z_i$ in the ``bearded seal count surveys" section seems inconsistent with the general
+definition of $Z_i$ at the beginning of your manuscript. I think of state variables as presence / absence,
+abundance, survival, etc. This definition for $Z_i$ seems more closely related to the expected value of a state
+process.
+
+\textcolor{blue}{
+  We've rewritten the model here in terms of the Poisson-binomial hierarchical mixture.
+In the following ``Software" section, we then indicate that we marginalize over $\bfZ$ when performing ML inference.
+}
+
+{\bf Other Comments}
+
+Line 65: This sentence seems to discount the value of good study design, which is still important for model-
+based estimation.
+
+\textcolor{blue}{
+We've altered text to be a little less cavalier, now stating that 
+``Design-based estimation requires that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.
+Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement."
+}
+
+Line 77: It's unclear to me why the distinction between fine and coarse-scale preferential sampling is important.
+They both seem to represent the process of dependence between sampling locations and the process
+of interest.
+
+\textcolor{blue}{
+As the size of the sampling unit decreases (with the limit being a point process model in continuous space) there is indeed little difference.
+But practically, it seems likely that one could use a reasonable design to select sampling units, but still open themselves to bias if they target non-representative habitat.
+However, this point is well taken, and we don't think the distinction necessarily needs to prominence that we've given it.
+We now address this topic for the first time in the discussion.
+}
+
+Line 86: I do not see how the definition offered here differs from your initial description of preferential
+sampling, unless the environmental covariate associated with the probability a location is selected is itself
+independent from the state variable of interest, which seems unlikely in general.
+
+\textcolor{blue}{
+This is an extremely important point that we may have failed to get across.
+Selecting sampling units based on a covariate related to abundance is just fine, \textit{provided} that the covariate is also used in the model for animal density/occurrence.
+In essence, we really only have preferential sampling if there the probability of sampling is related to the residuals (say $\hat{N_i} - N$) from a fitted SDM.
+We've produced an additional figure (now fig.
+1) that attempts to provide intuition about the distinction between sampling that is covariate-dependent and sampling that is preferential.
+}
+
+Line 186: I understand why you fit a relatively simple abundance model in this case. However, given the
+difficulties you encountered in incorporating even modestly complex structure on B, how well do you suppose
+your model will perform when you attempt to correct for nuisance processes?
+
+\textcolor{blue}{
+We don't know the answer to this question, although it seems like an important one.  In the discussion, we now suggest this as a topic for future research.
+}
+
+Line 195: You only describe 4 models below, not 6.
+
+\textcolor{blue}{Changed}
+
+Simulation study results: can you report bias for other parameters (e.g. slope parameters) in your model,
+even if this is relegated to an appendix?
+
+\textcolor{blue}{Yes, we have created an additional appendix table (S3) that summarizes bias in $b$ (the preferential sampling parameter), as well as bias in the species-habitat parameters.
+We don't look at bias in the sampling probability parameters because data were simulated by sampling without replacement (with a fixed number of sampled survey units), whereas the fitted model was Bernoulli (which assumes the number of sampled units is random).
+Thus parameters would not be directly comparable.
+We also provide convergence statistics in this appendix.
+%We note that the bias in the slope parameter of the species-habitat relationship %is likely because of the well-known confounding that occurs between regression %coefficients and spatially autocorrelated random effects in spatial models %\citep[e.g.][]{ReichEtAl2006,HodgesReich2010}.
+}
+
+Paragraph starting on line 219: I find this paragraph very confusing. In particular, it is unclear to me exactly
+what you mean when you state you ``...refitted $M_{cov=1;b=0}$ to produce an estimate of apparent abundance
+without this feature. Specifically, we refitted the model with 20 pseudo-absences in this portion of the study
+area to better inform abundance-covariate relationships". Does Fig. 5 reflect this re-fitted model? Why not
+fit all models in this manner for consistency?
+
+\textcolor{blue}{We now include these known zero values in all models.}
+
+Line 225: a model-averaged estimate of abundance?
+
+\textcolor{blue}{After correcting the link function in present model runs, it no longer makes sense to present a model-averaged estimate.}
+
+Table 1: I find these tables are easier to read if models are listed in decreasing order of $\Delta$AIC.
+
+\textcolor{blue}{We now present models in order of $\Delta$AIC.}
+
+Fig. 3: In my version, the blue and green dashed lines do not show up in the legend.
+
+\textcolor{blue}{Strange, this is not the case in our version.
+Guessing it is an issue with PDF compilation on the MEE submission website.
+We'll work with copy editors to ensure it looks right as it heads into production.}
 

### 2016-11-18 manuscript/pref_sampling_revision.tex
Commit during ms. revisions and change to link function; simulations now running with some error trapping and bounds set on parameter space.  Still getting some negative estimates using sdreport with bias.correct=TRUE when b is estimated.

--- 2016-11-15 manuscript/pref_sampling.tex
@@
 
-\address{\affilnum{1}National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; \affilnum{2} Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, National Marine Fisheries Service (NMFS),
-NOAA, 2725 Montlake Boulevard E, Seattle, WA 98112, USA}
+\address{\affilnum{1}Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; \affilnum{2} Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, NOAA, National Marine Fisheries Service,
+2725 Montlake Boulevard E, Seattle, WA 98112, USA}
 
@@
 \item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models fitted to count data.
-The approach works by jointly modeling wildlife state variables and the locations selected for sampling, and specifying a dependent correlation structure between the two models.
+The approach works by jointly modeling wildlife state variables and the locations selected for sampling and specifying a dependent correlation structure between the two models.
 
-\item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g.
-$40\%$) bias in estimates of animal density, and that our modeling approach can considerably reduce this bias.
+\item Using simulation, we show that even moderate levels of preferential sampling can lead to large (e.g.
+$40\%$) bias in estimates of animal density and that our modeling approach can considerably reduce this bias.
 
@@
 \def\bfeta{\boldsymbol{\eta}}
+\def\bfnu{\boldsymbol{\nu}}
+\def\bfmu{\boldsymbol{\mu}}
 
@@
 
-One of the main advantages of using SDMs is that one is no longer beholden to predetermined sampling frames, and can potentially use data gathered from non-randomized designs or platforms of opportunity to make inferences about animal populations \citep{JohnsonEtAl2010}.
-However,
-in a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.
+One of the main advantages of using SDMs is that one is less beholden to predetermined sampling frames and can potentially use data from non-randomized designs or opportunistic sampling to make inferences about animal populations \citep{JohnsonEtAl2010}.
+Several authors have shown that non-randomized designs can have poorer predictive ability than randomized designs when used to estimate species distributions \citep[e.g.][]{EdwardsEtAl2006,AlbertEtAl2010}, but that the optimal design may depend on the question of interest.
+For instance, the optimal design may differ depending on whether the inferential goal is to maximize predictive ability or to estimate the relationship between species presence and an important habitat covariate \citep{AlbertEtAl2010}.
+
+Less precise predictions may be a sacrifice one is willing to make if it is the only cost incurred when using data at non-random locations to make inferences about species distributions.
+After all, such data are often all that are available for certain taxa and locations (e.g.
+from citizen science programs), and are certainly less expensive to acquire than data from randomized, scientific surveys.
+However, recent research suggests that one may be opening oneself up to bias as well, which is potentially much more concerning than loss of precision.
+In a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.
 In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance.
 For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.
-Bias might also arise if surveying effort is higher near bases of operations, and if animal abundance is higher (or lower) near bases of operations than elsewhere in the landscape.
+Bias might also arise if survey effort is higher near bases of operations or features such as roads, particularly if animal abundance is higher (or lower) near these locations than elsewhere in the landscape.
+In our view, this phenomenon, termed ``preferential sampling," is potentially much more problematic, because estimates of animal abundance and occurrence from SDMs could be severely overstated.
 
-In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS), and describe several model-based approaches for detecting and correcting for such biases.
+In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS) and describe several model-based approaches for detecting and correcting for such biases.
 We start by describing a common currency for notation and basic model structures considered in this paper.
@@
   \label{eq:process}
-  Z_i & \sim & f(g^{-1}(\mu_i)).
+  [Z_i] & = & f(g^{-1}(\mu_i)).
 \end{eqnarray}
@@
 occupancy or abundance), $g()$ is a link function (e.g.
-probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value.
+probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, or spatially autocorrelated random effects.
+We use the bracket notation (e.g.
+$[Z_i]$) to denote a probability mass or probability density function (in the case of discrete or continuous distribution, respectively).
 In applications described in this paper, we write the intensity as
@@
 \begin{eqnarray}
-  Y_i & \sim & \text{Bernoulli}(Z_i p_i).
+  [Y_i|Z_i] & = & \text{Bernoulli}(Z_i p_i).
 \end{eqnarray}
@@
 \begin{eqnarray}
-  \label{eq:obs_pois}
-  Y_i & \sim & \text{Poisson}(Z_i A_i p_i),
+  \label{eq:obs_binomial}
+  [Y_i|Z_i] & = & \text{Binomial}(Z_i, A_i p_i),
 \end{eqnarray}
@@
 For the remainder of this treatment, we use bold symbols to denote vector-valued quantities or matrices.
-We also use standard bracket notation to denote probability mass and density functions.
-For instance $[{\bf Z}]$ denotes the marginal probability mass function for ${\bf Z}$, and  $[{\bf Z}|{\bf Y}]$ represents the conditional distribution of ${\bf Z}$ given ${\bf Y}$.
-We use $\boldsymbol{\mu}$ and $\boldsymbol{\nu}$ to denote log-scale abundance and the logit of the probability of sampling, so that $Z_i \sim f(\mu_i)$, and $R_i \sim f(\nu_i)$.
+We also use ``$|$" to specify that a distribution is conditional; for instance, if $[{\bf Y}]$ denotes the marginal probability mass function for ${\bf Y}$, then  $[{\bf Y}|{\bf Z}]$ represents the conditional distribution of ${\bf Y}$ given ${\bf Z}$.
+We use $\bfmu$ and $\bfnu$ to indicate the link-scale process of interest and the logit of the probability of sampling, respectively.
+As with $\bfmu$, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.
 We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.
@@
 
-One of the appealing aspects of model-based estimation is that there is no requirement that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame.
-For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location.
+Design-based estimation requires that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.
+Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement.
+For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location and still generate model-based estimates of abundance or occupancy.
 This can be a crucial advantage in surveys covering large areas with frequent inclement weather.
-It also opens the door for using platforms of opportunity, presence only, and citizen science data for estimation.
+It also opens the door for using presence only, citizen science, or opportunistically collected data for estimation.
 
-However, the manner in which effort is ultimately allocated can potentially have profound influence on SDM estimator performance.
-With respect to nonrandom sampling, two possible problems seem particularly likely in discrete spatial domains: coarse scale preferential sampling (CSPS), and fine scale preferential sampling (FSPS) (Fig.
-\ref{fig:pref}).
-FSPS arises when the observations taken at a  particular sampling unit are non-random with respect to the density of animals within that sampling unit.
-For instance, when allocating line transect survey effort, it may be tempting to place the transect in a manner that targets habitat or landscape features that maximize the number of animals that will be encountered.
-However, this strategy will clearly lead to positive bias when estimating density or abundance.
-
-By contrast, CSPS (hereafter, PS), the primary focus of this article, arises when the locations being sampled and the process of interest (e.g.
+However, the manner in which effort is allocated can potentially have profound influence on SDM estimator performance.
+Preferential sampling (PS) arises when the locations being sampled and the process of interest (e.g.
 density, occupancy) are conditionally dependent given modeled covariates \citep{DiggleEtAl2010}.
-For instance, PS can occur when the investigator uses a priori knowledge or observations of the state variable obtained during sampling to allocate survey effort in places where abundance or occurrence is known to be high.
+For instance, PS can occur when the investigator tends to sample more in places where a species has already been observed, or is known to be abundant.
 \citet{DiggleEtAl2010} showed that this type of PS can lead to bias when this extra information is not included in models for the state variable of interest.
-Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise.
+Specifically, PS arises when we consider the set of sampled locations as stochastic and when $[{\bf R},{\bf Z} | {\bf x}] \neq [{\bf R}|{\bf x}][{\bf Z}|{\bf x}]$, where \textbf{R} is an indicator vector whose elements $R_i$ are 1.0 if sampling unit $i$ is sampled and are zero otherwise (see e.g.
+fig.
+\ref{fig:PS_example}).
 We use this definition of PS throughout the rest of the manuscript, noting that it is somewhat different than has sometimes been used in the SDM literature.
@@
 
-As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS.
+As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS (e.g.
+fig.
+\ref{fig:PS_example}).
 However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.
@@
 \end{eqnarray}
-the spatially continuous relative intensity ($\psi_i$) of sampling locations at $i$ could be written as
+the log-scale relative intensity of sampling locations at $i$ ($\nu_i$) could be written as
 \begin{eqnarray}
-   \psi_i & \propto & \exp(\xi_i + b \mu_i).
+   \nu_i & \propto & \xi_i + b \mu_i,
    \label{eq:p_cont}
 \end{eqnarray}
+where $\mu_i$ and $\xi_i$ includes explanatory covariates.
 Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g.
@@
 is an association between the spatial surface of interest and the sampling density that cannot be
-explained by the shared spatial covariates."  \citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in Eq.
-\ref{eq:Z_cont}, finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.
+explained by the shared spatial covariates" (see e.g.
+fig.
+\ref{fig:PS_example}).
+\citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn.
+(\ref{eq:Z_cont}), finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.
 
@@
 Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurrence) and the locations chosen for sampling.
-For the abundance process model, we start with eq.
-\ref{eq:process} as a general formulation for non-Gaussian data, writing the link-scale expectation as in eq.
-\ref{eq:mu}.
+For the abundance process model, we start with eqn.
+(\ref{eq:process}) as a general formulation for non-Gaussian data, writing the link-scale expectation as in eqn.
+(\ref{eq:mu}).
 Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
@@
  \label{eq:R}
-  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
+  [R_i] & = & \text{Bernoulli}(h^{-1}(\nu_i)),
 \end{eqnarray}
@@
  In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).
-The predictive covariates ${\bf x}_i$ from Eq.
-\ref{eq:mu} and ${\bf x}_i^*$ from Eq.
-\ref{eq:nu} may or may not be the same.
-Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both Eqs.
-\ref{eq:mu} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.
+The predictive covariates ${\bf x}_i$ from eqn.
+(\ref{eq:mu}) and ${\bf x}_i^*$ from eqn.
+(\ref{eq:nu}) may or may not be the same.
+Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both eqns.
+(\ref{eq:mu}) and (\ref{eq:nu}), allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.
 The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.
-In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in Eq.
-\ref{eq:nu} are estimated to permit parameter identification.
+In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in eqn.
+(\ref{eq:nu}) are estimated to permit parameter identification.
 
- The formulation in Eq.
-\ref{eq:nu} is similar to the one previously proposed for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.
+ The formulation in eqn.
+(\ref{eq:nu}) is similar to the one previously proposed for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.
 There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the PS process \citep{RoyleBerliner1999}.
@@
 Alternatively, we could allow the degree of PS to vary across the landscape.
-For instance, one can contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
-Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surface or spline formulation \citep{RoyleBerliner1999}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.
+For instance, one could contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
+However, to our knowledge the identifiability of higher order models (such as trend surfaces) has not previously been investigated.
 
@@
 \end{eqnarray*}
-where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to Eq.
-\ref{eq:mu}.
+where $i$ indexes survey unit $i$, and $\mu_i$ is determined according to eqn.
+(\ref{eq:mu}).
 Abundance was generated as a function of a single spatially autocorrelated landscape covariate, as well as residual spatial autocorrelation ($\delta_i$) and overdispersion (fig.
@@
 In the first, we set ${\bf B}=0$, so that surveyed locations were selected independently of the abundance generating process.
-For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=5$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.
+For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=3$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.
 We refer to these scenarios as moderate and pathological PS, respectively (see fig.
@@
 We fitted two different models to each count dataset, both of which were provided the habitat covariate used (in part) to generate the data for which a log-linear coefficient $\beta$ was estimated.
-In the first model, the elements of \textbf{B} in eq.
-\ref{eq:nu} were all set to zero.
-In this case, the abundance and sampling process submodels were independent, as is the case canonical SDMs (at lest when fitted to presence-absence or count data).
+In the first model, the elements of \textbf{B} in eqn.
+(\ref{eq:nu}) were all set to zero.
+In this case, the abundance and sampling process submodels were independent, as is the case canonical SDMs (at least when fitted to presence-absence or count data).
 In the second model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix.
-
-We used maximum likelihood to conduct statistical inference.
-In particular, we used Template Model Builder \citep[TMB; ][]{KristensenEtAl2016}, interfaced with the R programming environment, to conduct maximization.
-The TMB software uses a Laplace approximation to integrate out random effects ($\bfeta$ and $\bfdelta$), and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
-This approach resulted in a facile implementation and speedy computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.
-Further detail on statistical methods are provided in Appendix S1; requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.
+For ease of exposition, we set $p_i = 1.0$ for all simulations (i.e.
+all animals present were assumed to be detected).
 
@@
 
-We applied our modeling technique to counts of bearded seals obtained on aerial transects flown over the eastern Bering Sea from 10-16 April, 2012 (Fig.
+We applied our modeling technique to counts of bearded seals obtained on aerial transects flown over the eastern Bering Sea from 10-16 April 2012 (fig.
 \ref{fig:bearded_ests}).
 These counts were gathered as part of a larger survey designed to estimate abundance of four species of ice-associated seals; the survey is described in greater detail elsewhere \citep{ConnEtAl2014,ConnEtAl2015}.
-The survey area considered here consists of 25 by 25 km grid cells bordered to the north by the Bering strait, to the west by the international date line, to the south by maximal April ice extent, and to the east by the Alaska, USA mainland.
+The survey area considered here consists of 25 x 25 km grid cells bordered to the north by the Bering Strait, to the west by the International Date Line, to the south by maximal April ice extent, and to the east by Alaska.
 Here, we limit counts to those gathered within a one week period so that relative abundance will remain relatively constant throughout the study area.
@@
 
-To test for such an effect, we modeled bearded seal counts using the formulation
+To test for such an effect, we conceptualized bearded seal counts as arising according to the formulation
 \begin{eqnarray*}
-  [Y_i|Z_i] & = & \textrm{Poisson}(P_i Z_i), \text{ where}\\
-  Z_i & = & A_i \exp(\mu_i),
+  [Z_i | \mu_i] & = & \textrm{Poisson}(a_i \exp(\mu_i)) \\
+  [Y_i|Z_i] & = & \textrm{Binomial}(Z_i, A_i), \text{ where}
 \end{eqnarray*}
-where $P_i$ defines the proportion of grid cell $i$ that is sampled, $A_i$ gives the proportion of grid cell $i$ that is composed of salt water habitat, and $\mu_i$ is defined in Eq. \ref{eq:mu}.
-We modeled the grid cells that were chosen for sampling using Eqs. \ref{eq:R}-\ref{eq:nu}.
+$A_i$ defines the proportion of grid cell $i$ that is sampled, $a_i$ gives the proportion of grid cell $i$ that is composed of salt water habitat, and $\mu_i$ is defined in eqn. (\ref{eq:mu}).
+We modeled the grid cells that were chosen for sampling using eqns. (\ref{eq:R}-\ref{eq:nu}).
 
-We fitted a total of six models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1}$) to bearded seal count data using the same estimation framework as in the simulation study.
+We fitted a total of four models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1}$) to bearded seal count data using the same estimation framework as in the simulation study.
 Models varied by (i) whether or not habitat and landscape variables were used as predictors of bearded seal density ($cov=1$ and $cov=0$, respectively), and (ii) the form of PS ($b=0$ indicates no PS; $b=1$ indicates ${\bf B}=b{\bf I}$, where $b$ is an estimated parameter).
-We also attempted to fit models where the PS ${\bf B}$ matrix varied over the landscape using a trend surface specification, but parameter identification was suspect in these models and are not reported here (see Appendix S2 for more information).
-When habitat and landscape variables were included, we used three log-linear predictors: linear and quadratic functions of sea ice concentration, and distance from the southern ice edge.
-Remotely sensed sea ice data were obtained at a $25 \times 25$ km resolution from the National Snow and Ice Data Center, Boulder, CO, USA, as described by \citet{ConnEtAl2014}.
+We also attempted to fit models where the PS ${\bf B}$ matrix varied over the landscape using a trend surface specification, but parameter identification was suspect in these models so estimates are not reported here (see Appendix S2 for more information).
+When habitat and landscape variables were included, we used three log-linear predictors: linear and quadratic functions of sea-ice concentration, and distance from the southern ice edge.
+Remotely sensed sea-ice data were obtained at a $25 \times 25$ km resolution from the National Snow and Ice Data Center, Boulder, CO, USA, as described by \citet{ConnEtAl2014}.
 Models for $\mu_i$ and $\nu_i$ both utilized spatially autocorrelated random effects with a Mat\'{e}rn covariance function between grid cell centroids (see Appendix S1 for further details).
@@
 for $\mu_i$ and $\nu_i$).
+
+\subsection{Computing}
+
+We used maximum (marginal) likelihood to conduct statistical inference.
+For increased numerical tractability, we used the result that the marginal distribution of a Poisson-binomial mixture is itself a Poisson distribution to formulate the likelihood.
+In particular, we replaced the binomial observation model (eqn.
+\ref{eq:obs_binomial}) with the following unconditional specification:
+\begin{eqnarray}
+\label{eq:obs_pois}
+   [Y_i] & = & \text{Poisson}(\exp(\mu_i) A_i p_i).
+\end{eqnarray}
+Predictions of abundance in locations $i$ and attendant standard errors were then estimated with respect to $\hat{N}_i = \exp(\mu_i)$.
+
+We used Template Model Builder \citep[TMB; ][]{KristensenEtAl2016}, interfaced with the R programming environment, to conduct maximization.
+The TMB software uses a Laplace approximation to integrate out random effects ($\bfeta$ and $\bfdelta$), and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensen2016} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
+This approach resulted in a facile implementation and speedy computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.
+Further detail on statistical methods are provided in Appendix S1; requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.
 
@@
 
+A substantial proportion of simulation replicates either did not converge or resulted in divergent estimates of $b$ (i.e.
+the parameter was estimated on a boundary when a flexible constraint was imposed).
+
 Estimates of cumulative animal abundance across simulated landscapes were median unbiased for both estimation methods when the sites selected for sampling were independent of animal density, though when $b$ was estimated, abundance estimates were more right skewed and had higher variance (fig.
@@
 Under moderate PS ($b=1$), estimation of the PS parameter $b$ led to a median bias of 5\%, while the canonical SDM model ignoring preferential sampling had a median bias of 40\%.
-Under pathological PS ($b=5$), both estimation methods were extremely biased, but was even more severe for the naive model ignoring PS (fig.
+Under pathological PS ($b=3$), the canonical SDM model was extremely biased while the joint model remained unbiased (fig.
 \ref{fig:sim_bias}).
@@
 
-In this study, we showed that coarse-grained preferential sampling (Fig.
-\ref{fig:pref}) can have a profound impact on the quality of estimates (e.g.
+In this study, we showed that preferential sampling can have a profound impact on the quality of estimates (e.g.
 animal abundance) when sampling is non-randomized.
@@
 When PS was present, we were able to substantially reduce bias by conducting estimation under a framework where the state variable of interest and the sites chosen for sampling were jointly modeled under a dependent covariance structure.
-In absence of PS, simulations suggest that this structure results in lower precision than a model without a PS effect; thus the need to account for PS reduces the quality of inference.
+In absence of PS, simulations suggest that this structure results in lower precision than a model without a PS effect; thus, the need to account for PS reduces the quality of inference.
 
-Bias attributed to PS may seem counterintuitive, especially given the maxim in survey sampling to allocate more effort to strata for which animal density is high.
-For instance, in large scale line transect surveys under stratified sampling, the optimal amount of effort that should be allocated to stratum $s$ is $A_s D_s^{0.5}$, where $A_s$ is the area of $s$ and $D_s$ is the anticipated density \citep[][eqn 7.7]{BucklandEtAl2001}.
-Thus, there are theoretical reasons to sample more in high density areas than in low density areas.
-The obvious solution in this instance is to account for variation in sampling intensity with explanatory covariates or post hoc stratification.
-However, it is not always clear how to perform post hoc stratification when effort is allocated in a subjective manner.
+Bias attributed to PS may seem counterintuitive, especially given the maxim in survey sampling to allocate more effort to strata for which animal density is high (at least when the goal is to maximize precision of a population estimate).
+For instance, in large-scale line transect surveys under stratified sampling, the optimal amount of effort that should be allocated to stratum $s$ is $A_s D_s^{0.5}$, where $A_s$ is the area of $s$ and $D_s$ is the anticipated density \citep[][eqn 7.7]{BucklandEtAl2001}.
+Thus, there are theoretical reasons to sample more in high density areas than in low-density areas.
+One solution is to sample more in locations where covariates suggest abundance will be higher; in this case, sampling will not be preferential as long as the covariate used to select sampling locations is also used in the species distribution model fitted to the data (fig.
+\ref{fig:PS_example}).
+Another approach is to account for variation in sampling intensity with explanatory covariates or \textit{post hoc} stratification.
+However, it is not always clear how to perform \textit{post hoc} stratification when effort is allocated in a subjective manner.
 
@@
 
-We attempted to fit models to bearded seal data where the degree of PS changed over the landscape, in a similar manner to multivariate spatial models \citep{RoyleBerliner1999}.
+We attempted to fit models to bearded seal data where the degree of PS changed over the landscape, as \citet{RoyleBerliner1999} suggested might be possible (but never demonstrated) for multivariate spatial models.
 However, such models led to difficulties with parameter identification in our bearded seal application (see Appendix S2).
-Evidently, such models may require greater spatial balance or richer data sets.
-At this time, we suggest limiting initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
-Although more complex models are clearly identifiable in some situations \citep{RoyleBerliner1999}, further research on the viability of such models as a function of data quality appears warranted.
+At this time, we are not sure whether these models can be fit reliably, and suggest limiting initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
+Further research into the viability of including spatial variation in the ${\bf B}$ matrix certainly appears warranted.
 It would also be worthwhile to investigate whether parameter identification varies as a function of the support of the state variable being modeled (e.g.
 binary vs.
-count data).
+count data), and whether parameters remain identifiable when nuisances processes such as incomplete detection are modeled.
 
@@
 If, instead, a fixed number of locations are sampled, the Bernoulli model is somewhat misspecified.
-Our simulations suggest some robustness to this misspecification, as the Bernoulli model performed reasonably well when sampling was without replacement for a fixed sample size (Fig.
+Our simulations suggest some robustness to this misspecification, as the Bernoulli model performed reasonably well when sampling was conducted with a fixed sample size (fig.
 \ref{fig:sim_bias}).
-Still, a more precise treatment would need to rely on an extended hypergeometric distribution with variable inclusion probabilities when formulating the sampling model; this extension is nontrivial.
+A more precise treatment would require extending the hypergeometric distribution to have variable inclusion probabilities when formulating the sampling model; this extension is nontrivial.
+
+We have focused here on selection of survey units at a rather coarse scale.
+If survey units are large relative to the area that is sampled, ecologists also need to pay attention to how sampling is conducted within survey units selected for sampling.
+In particular, the areas visited within selected sample units should be representative (or random) with respect to available habitat within the survey unit or additional bias can result (e.g.
+fig.
+\ref{fig:pref})
 
@@
   \hline
-  Model & Log likelihood & Params & $\Delta \textrm{AIC}$ & Wgt & $\hat{N}$(SE) \\
+  Model & Log likelihood & Params. & $\Delta \textrm{AIC}$ & Wgt & $\hat{N}$(SE) \\
   \hline
-  $M_{cov=0,b=0}$ & -2667.1 & 3 & 21.7 & 0.00 & 68556 (7408)\\
-  $M_{cov=0,b=1}$ & -2665.3 & 4 & 20.1 & 0.00 & 45857 (5114) \\
-  $M_{cov=1,b=0}$ & -2650.3 & 9 &  0.0 & 0.53 & 59312$^\dag$ (5231)  \\
-  $M_{cov=1,b=1}$ & -2649.4 & 10 & 0.3 & 0.47 & 49826 (10369) \\
+ $M_{cov=1,b=1}$ & -2808.3 & 10 & 0.0 & 0.65 & 45216 (4534) \\
+ $M_{cov=1,b=0}$ & -2809.9 & 9 &  1.2 & 0.35 & 59313 (5231)  \\
+ $M_{cov=0,b=1}$ & -2822.6 & 4 & 16.7 & 0.00 & 39397 (2931) \\
+ $M_{cov=0,b=0}$ & -2829.1 & 3 & 27.5 & 0.00 & 54001 (4243)\\
  \hline
@@
 \\
-$\dag$ Refitted model; see \textit{Results}.
 \end{table}
@@
 \pagebreak
-%\begin{figure*}
-%\begin{center}
-%\includegraphics[width=170mm]{../figures/state_diagram.eps}
-%\caption{A depiction of potential inferential obstacles when data from marked animals are %used to address mixed migration strategies.
-The migratory strategy of an animal will often %be unknown at the time of marking, but observational data may be used to help infer the %proportion of animals subscribing to each strategy, as well as the probability of %migrating for conditional migrants ($\theta_t$).
-%behavior} \label{fig:state_diagram}
-%\end{center}
-%\end{figure*}
 
-\begin{figure*} %[ht]
+\begin{figure*}
 \begin{center}
-\includegraphics[width=170mm]{Pref_sampling_diag.pdf}
-\caption{A depiction of two types of preferential sampling.
-In (A), an investigator preferentially places point transects (red squares) within regions of high known animal density (blue polygons).
-This can cause bias in abundance or occupancy estimators unless this a priori knowledge about density is explicitly modeled.
-In (B), a fine scale version of preferential sampling occurs when a line transect (red line) is intentionally placed across a region of high quality habitat.
-If a landscape is discretized into homogeneous survey units for analysis (as in a grid), it is essential that the habitat surveyed within each survey unit be randomly determined when estimating abundance.
-If not, bias (usually positive) can be expected.}
-\label{fig:pref}
+\includegraphics[width=170mm]{pref_sampling_ex.pdf}
+\caption{A depiction of preferential sampling in a simplified setting.
+In this example, we generate abundance as a log-linear function (blue lines) of a hypothetical covariate subject to lognormal error (black and orange points).
+In (A.
+Nonpreferential), we sample 20 abundance values (orange points), where the probability of selection is proportional to the covariate value.
+In (B.
+Preferential), the probability of selection is proportional to the abundance value.
+Orange lines give bias-corrected log-linear models fitted to sampled points (assuming abundances are observed perfectly).
+In both cases, the mean sampled abundance is greater than the population mean.
+However, the sample in (A) is non-preferential if the covariate is also used to estimate the species-covariate relationship.
+In contrast, the sample in (B) is preferential and can lead to biased inference.}
+\label{fig:PS_example}
 \end{center}
@@
 \caption{Aerial survey counts and estimated apparent abundance of bearded seals in the eastern Bering Sea, April 10-16, 2012.
-Counts and estimates are shown relative to a survey grid that extends south from the Bering Strait and borders the Alaska, USA mainland to the east.
+Counts and estimates are shown relative to a survey grid that extends south from the Bering Strait and borders Alaskato the east.
 In (A), tan shading denotes land, unsurveyed grid cells appear in dark gray, and counts appear in a white-blue spectrum.
@@
 
+\begin{figure*} %[ht]
+\begin{center}
+\includegraphics[width=170mm]{Pref_sampling_diag.pdf}
+\caption{A depiction of two types of preferential sampling.
+In (A), an investigator preferentially places point transects (red squares) within regions of high known animal density (blue polygons).
+This can cause bias in abundance or occupancy estimators unless this \textit{a priori} knowledge about density is explicitly modeled.
+In (B), a fine-scale version of preferential sampling occurs when a line transect (red line) is intentionally placed across a region of high quality habitat.
+If a landscape is discretized into homogeneous survey units for analysis (as in a grid), it is essential that the habitat surveyed within each survey unit is randomized (or representative of available habitat) when estimating abundance.
+If not, bias (usually positive) can be expected.}
+\label{fig:pref}
+\end{center}
+\end{figure*}
+
 \end{document}

### 2016-11-18 TMB_version/examples/simple_example_v6.R

@@ -41,7 +41,7 @@ RandomSeed = 123456
 n_sim = 500
 
 # Configurations
-b_set = c(0,1,5)            # Impact of delta on sampling intensity
+b_set = c(0,1,3)            # Impact of delta on sampling intensity
 EM_set = c("fix_b","est_b")
 
 # Configurations
@@ -52,216 +52,245 @@ EM_set = c("fix_b","est_b")
 #diag(Q) = -1 * colSums(Q)
 
 # Results
-Results = array(NA, dim=c(length(b_set),length(EM_set),n_sim,3), dimnames=list(paste0("b=",b_set),EM_set,NULL,c("b","sum_Ztrue","sum_Zpred")) )
+Results = data.frame(matrix(NA,length(b_set)*length(EM_set)*n_sim,12))
+colnames(Results)=c("sim","b","EM","b_est","SEb","N_true","N_est","N_SE","Convergence","LogLik","BiasBeta11","BiasBeta12")
 
 # Loop through replicates
+counter=1
 for(i in 1:n_sim){
-for(SimI in 1:length(b_set)){
-for(EstI in 1:length(EM_set)){
-  cat(paste("Simulation ",i," SimI ",SimI," EstI ",EstI,"\n"))
-  # Settings
-  b = b_set[SimI]
-  EM = EM_set[EstI]
-  set.seed( RandomSeed + i )
+  for(SimI in 1:length(b_set)){
+    for(EstI in 1:length(EM_set)){
+      cat(paste("Simulation ",i," SimI ",SimI," EstI ",EstI,"\n"))
+      # Settings
+      b = b_set[SimI]
+      EM = EM_set[EstI]
+      set.seed( RandomSeed + i )
+      
+      #note: betax set to zero in last commit by Jim (v4)
+      betax = betax_prob = runif(1,-.5,.5)       # Impact of x on density
+      # did this so places that have covariate values with higher abundance are more likely to be sampled
+      
+      # Spatial model
+      loc_s = expand.grid( "x"=1:grid_dim['x'], "y"=1:grid_dim['y'])
+      model_delta <- RMgauss(var=SD_delta^2, scale=SpatialScale)
+      model_x <- RMgauss(var=SD_x^2, scale=SpatialScale)
+      model_eta <- RMgauss(var=SD_eta^2, scale=SpatialScale)
+      
[…]

### 2016-11-18 pref_sampling/inst/run_bearded_models.R

@@ -1,12 +1,8 @@
-### run simple preferential sampling example.  Doing in this in two steps since there seems to be
-# a conflict between TMB loop and rgeos ('gCentroid') functionality
-# need to declare an ADREPORT for Z_s in TMB for this
+### run simple preferential sampling models on bearded seal data
 library(PrefSampling)
-
-data(AlaskaBeringData)
+data(Data_for_ST_plot) #includes spatial objects associated w landscape
 data(Bearded_effort)
-data(Data_for_ST_plot)
-
+data(AlaskaBeringData)  #from bass package on github/nmml
 Effort=Bearded_effort
 
 #create single Grid SpatialPolygonsDataFrame, averaging over covariate values
@@ -20,6 +16,7 @@ Grid.new@data=Grid.new@data/length(Days)
 Which.incl = which(Effort$Mapping[,2]<8)
 Mapping=Effort$Mapping[Which.incl,1]
 Prop_sampled=Effort$Area.trans[Which.incl]
+Prop_sampled[which(is.na(Prop_sampled))]=1.0
 Count.data=Effort$Count.data[Which.incl,]
 
 # Settings
@@ -41,11 +38,13 @@ set.seed(12345)
 # massage into paper notation
 s_i = Mapping
 y_s = ifelse(1:n_cells %in% s_i, 1, 0)
-x_s_z = cbind(Grid.new@data[,"ice_conc"],Grid.new@data[,"ice_conc"]^2,Grid.new@data[,"dist_edge"])  
-x_s_y = x_s_z
+x_s_z = cbind(Grid.new@data[,"ice_conc"],Grid.new@data[,"ice_conc"]^2,Grid.new@data[,"dist_edge"],Grid.new@data[,"depth"],Grid.new@data[,"depth"]^2,Grid.new@data[,"dist_land"])  
+x_s_y = cbind(Grid.new@data[,"dist_land"])  
 x_s_b = matrix(1,nrow=n_cells,ncol=3)  #design matrix for preferential sampling effect
 Centroids=gCentroid(Grid.new,byid=TRUE)
 loc_s=cbind(Centroids$x,Centroids$y)/25067.53 #standardize by making distance between neighboring cells = 1.0
+loc_s[,1]=loc_s[,1]/mean(loc_s[,2])
[…]

### 2016-11-18 pref_sampling/inst/run_bearded_models_ice0.R

@@ -1,199 +1,0 @@
-### run simple preferential sampling example.  Doing in this in two steps since there seems to be
-# a conflict between TMB loop and rgeos ('gCentroid') functionality
-# need to declare an ADREPORT for Z_s in TMB for this
-library(PrefSampling)
-data('Data_for_ST_plot') #includes spatial objects associated w landscape
-data('Bearded_effort_ice0')  #formatted with BOSSst "format_Bering2012_all.R" ; includes several extra absences in southern ice-free area to prevent large estimates there
-data("AlaskaBeringData")  #from bass package on github/nmml
-Effort=Bearded_effort
-
-#create single Grid SpatialPolygonsDataFrame, averaging over covariate values
-Grid.new=Data$Grid$y2012[[1]]
-Days = c(6:12) #Apr 10-16
-Grid.new@data=0*Grid.new@data
-for(iday in Days)Grid.new@data=Grid.new@data+Data$Grid$y2012[[iday]]@data
-Grid.new@data=Grid.new@data/length(Days)
-
-#limit effort data to April 10-16
-Which.incl = which(Effort$Mapping[,2]<8)
-Mapping=Effort$Mapping[Which.incl,1]
-Prop_sampled=Effort$Area.trans[Which.incl]
-Prop_sampled[which(is.na(Prop_sampled))]=1.0
-Count.data=Effort$Count.data[Which.incl,]
-
-# Settings
-n_cells=length(Grid.new)
-n_samp = length(Mapping)
-#SpatialScale = sqrt(n_cells)/5  # Range ~ 2*Scale
-#SD_eta = SD_x = SD_delta = 1
-#beta0 = 2
-Use_REML = FALSE   # 
-#Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
-Spatial_model = c("SPDE_GMRF", "ICAR")[1]
-Alpha = 2  # Smoothness for GMRF, 1 or 2 (1 is faster)
-
-# Model configurations
-EM_set = c("fix_b","est_b")
-
-set.seed(12345)
-
-# massage into paper notation
-s_i = Mapping
-y_s = ifelse(1:n_cells %in% s_i, 1, 0)
[…]

### 2016-11-18 pref_sampling/src/PrefSampling.cpp

@@ -254,6 +254,7 @@ Type objective_function<Type>::operator() ()
   ADREPORT( beta_j );
   ADREPORT( total_abundance);
   if(Options_vec(4)==1){
+    ADREPORT(beta_k);
     ADREPORT( beta_b ); 
     ADREPORT(Z_s);
   }

### 2017-04-24 manuscript/ReviewerResponses.tex
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

@@
 \def\bfB{\textbf{B}}
+\def\bfZ{\textbf{Z}}
 \newcommand{\VS}{V\&S}
@@
 We provide a marked up PDF version of our manuscript showing all changes made.
+
+In the process of conducting revisions we corrected an error in our estimation code (somewhere in the model development and testing process, the link function for $R_i$ was misspecified as multinomial-logit instead of logit).
+Correcting this error has led to several qualitative changes to the paper.
+Conclusions from the simulation study were strengthened; however, bearded seal conclusions were made more murky.
+In particular, both of the model types incorporating preferential sampling (linear and trend surface) appeared to produce biologically unrealistic estimates in the bearded seal example.
+This was particularly troubling because AIC favored these models.
+We decided to conduct an additional cross validation analysis to quantify predictive ability, and found that non-preferential models produced more accurate predictions.
+We now argue against using preferential sampling models for the case study.
 
@@
 
+\textcolor{blue}{
 Additionally, we have made the following changes to try to make the manuscript accessible to a wider audience:
 \begin{itemize}
-  \item 
-\end{itemize}
+  \item An additional figure (fig. 1) to more clearly depict preferential sampling for visual learners.
+  \item  In the methods section, we motivate inclusion of spatially autocorrelated random effects by indicating ``Spatially autocorrelated random effects are often included to allow animal density to vary smoothly in space and to account for residual patchiness not explained by predictive covariates."
+  \item We changed ``it is usually impossible to view system state perfectly even in locations where sampling is conducted" to ``it is usually impossible to conduct a complete census of all animals present in a surveyed location"
+  \item We provide examples of survey and observer specific covariates. (visibility, observer experience)
+  \item We changed the ``underlying spatial field of interest" and ``spatially-referenced random field values" to ``the process of interest"
+  \item We changed ``surveyed locations were selected independently of the abundance generating process" to ``surveyed locations were randomly selected (i.e. independent of abundance)."
+  \item We changed ``In this case, the abundance process and sampling intensity submodels were independent, as is the with case canonical SDMs (at least when fitted to presence-absence or count data)" to ``In this case, sampling intensity is assumed to be independent of underlying abundance, as implicitly assumed by most SDMs in the ecological literature."
+  \item Numerous other small changes to wording are documented in the track changes document.
+\end{itemize}}
 
@@
 In fact, their one example on ozone-temperature analysis uses the same simple diagonal parameterization we used here.
-So we think the question remains whether more complicated models for $\bfB$ are ever identifiable, and if so, whether they are reasonable to fit with data sets typical of ecological applications.
 We've adjusted text in several places to reflect this.
 For instance, we state that ``However, to our knowledge the identifiability of higher order models (such as trend surfaces) has not been investigated" when talking about higher order preferential sampling models.
-In the discussion, we indicate that ``We attempted to fit models to bearded seal data where the degree of PS changed over the landscape, as \citet{RoyleBerliner1999} suggested might be possible (but never demonstrated) for multivariate spatial models."  We also suggest that identifiability of such models be further investigated in future research.
+}
+
+\textcolor{blue}{
+  After correcting the error in our estimation model link function (see below), qualitative results have changed somewhat in the bearded seal example, with both the simple single parameter $b$ model and the trend surface model producing similar, vexing results.
+So, the question of identifiability may actually be one of estimability - i.e., how often do these methods produced reasonable estimates for real world datasets?
+We know the single parameter $b$ model works $\approx 80\%$ of the time with simulated data; were we simply unlucky with the bearded seal data set?
+Or do other factors (e.g., overdispersion, line transects vs.
+point counts) affect the quality of inference?
+In the discussion, we now suggest that identifiability and reliability of inference of such models be investigated in future research.
 }
@@
 \textcolor{blue}{
-  This wording was too ambiguous, and was likely a result of initially fitting different types of spatial models to the data.
-For instance, ICAR type models have a known parameter redundancy built in, such that the intercept is confounded with the spatial process \citep{RueHeld2005}.
-The type of Gaussian process model we settled on and used in analysis does not suffer from this deficiency.
-Further confusion resulted from incorrectly specifying a multinomial logit link function in our initial analyses.
+  Initially we were trying to fit ICAR type models which have a known parameter redundancy built in, such that the intercept is confounded with the spatial process \citep{RueHeld2005}.
+However, the type of Gaussian process model we settled on and used in analysis does not suffer from this deficiency, but confusion resulted because somewhere along the way we mistakenly built in a multinomial logit link function into the estimation code for the $R_i$ (whose intercept is also not identifiable).
 We have rectified this mistake in the present version, and all models for $R_i$ are now Bernoulli with a logit link, so that the intercept term is now identifiable.
-We now report results for this corrected analysis.
+We have rerun all analyses with this corrected configuration.
 }
@@
 5 and $\nu_i$ in eq.
-6, although we retain $xi_i$, since in the next section we also wish to write $\nu_i$ in terms of shared spatial effects.
+6, although we retain $\xi_i$, since in the next section we also wish to write $\nu_i$ in terms of shared spatial effects.
 This is all a little tricky, in that the \citet{DiggleEtAl2010} paper considers a continuous support model (whereas ours is discrete), and the structure is slightly different.
@@
 As the size of the sampling unit decreases (with the limit being a point process model in continuous space) there is indeed little difference.
-But practically, it seems likely that one could use a reasonable design to select sampling units, but still open themselves to bias if they target non-representative habitat.
-However, this point is well taken, and we don't think the distinction necessarily needs to prominence that we've given it.
+But practically, it seems likely that one could use a reasonable design to select sampling units, but still open themselves to bias if they target non-representative habitat within individual sampling units.
+However, this point is well taken, and we don't think the distinction necessarily needs the prominence we've given it.
 We now address this topic for the first time in the discussion.
@@
 
-\textcolor{blue}{Changed}
+\textcolor{blue}{We now report results form all 6 models.}
 
@@
 \textcolor{blue}{Yes, we have created an additional appendix table (S3) that summarizes bias in $b$ (the preferential sampling parameter), as well as bias in the species-habitat parameters.
-We don't look at bias in the sampling probability parameters because data were simulated by sampling without replacement (with a fixed number of sampled survey units), whereas the fitted model was Bernoulli (which assumes the number of sampled units is random).
-Thus parameters would not be directly comparable.
 We also provide convergence statistics in this appendix.
+Thanks for this suggestion; in particular, it helps show that preferential sampling is not necessarily a big deal if one is only interested in drawing inference on parameters informing species-habitat relationships.
 %We note that the bias in the slope parameter of the species-habitat relationship %is likely because of the well-known confounding that occurs between regression %coefficients and spatially autocorrelated random effects in spatial models %\citep[e.g.][]{ReichEtAl2006,HodgesReich2010}.
@@
 
-\textcolor{blue}{We now include these known zero values in all models.}
+\textcolor{blue}{Our initial thought was that inserting ``pseudo-absences" could possibly impact inference about the preferential sampling parameter because those locations were not actually sampled and shouldn't affect the $R_i$ model.
+So, we conducted model selection without pseudo-absences, but then inserted them when making predictions.
+In the revision, we've circumvented this issue by introducing depth as an additional covariate (with linear and quadratic effects).
+This seems to have remedied the issue with positive estimates of abundance in offshore areas.
+So, no need to mess around with pseudo-absences anymore.}
 
@@
 
-\textcolor{blue}{We now present models in order of $\Delta$AIC.}
+\textcolor{blue}{We now present models in order of decreasing predictive ability (as measured with cross-validated mean square prediction error)}
 
@@
 Guessing it is an issue with PDF compilation on the MEE submission website.
-We'll work with copy editors to ensure it looks right as it heads into production.}
+We'll work with copy editors to ensure it looks right if it heads into production.}
 

### 2017-04-24 manuscript/AppendixS1_revision.tex
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

--- 2016-05-26 manuscript/AppendixS1.tex
@@
 
+\rm
+
 \vspace{0.7cm}
@@
 
-$^1$ National Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; $^2$ Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, National Marine Fisheries Service (NMFS),
-NOAA, 2725 Montlake Boulevard E, Seattle, WA 98112, USA
+$^1$ Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA, National Marine Fisheries Service, 7600 Sand Point Way NE, Seattle, WA 98115 USA; $^2$ Fisheries Resource Assessment and Monitoring Division (FRAM), Northwest Fisheries Science Center, NOAAA, National Marine Fisheries Service, 2725 Montlake Boulevard E, Seattle, WA 98112, USA
 
@@
 In this appendix, we provide further details on model development, implementation, and simulation testing for the preferential sampling model outlined in the main article text.
-Our modeling approach works by constructing (a) a species distribution model (SDM) describing abundance or presence/absence that includes spatially autocorrelated random effects, and (b) a model for the probability of site selection.
-Both submodels can be expressed in terms of habitat or landscape covariates.
-However, site selection probability can also be written as a function of the same spatially autocorrelated random effects in (a).
+Our modeling approach works by constructing (a) a species distribution model (SDM) that explains variation in survey counts, and (b) a model for the probability of site selection.
+Both submodels can be expressed in terms of habitat or landscape covariates and spatially autocorrelated random effects.
 Including the same set of random effects in both submodels is the mechanism for allowing dependence betwen the two models, and allows us to estimate and control for the potentially biasing effects of preferential sampling.
@@
 In some cases, observations may be equivalent to the state variable in sampled cells (i.e.
-when detection probability is 1.0).
+with complete area coverage and detection probability equal to 1.0).
 In others, an observation model $[Y_i | Z_i, \boldsymbol{\theta}]$ relating $Y_i$ to the underlying process of interest will be required, where $\boldsymbol{\theta}$ are nuisance parameters describing detection.
@@
 \end{eqnarray*}
-where $A_i$ gives the proportion of survey unit $i$ that is sampled and $p_i$ is a detection probability ($p_i$ is set to 1.0 for the simulation study and an informative prior distribution is specified in the bearded seal example).
+where $A_i$ gives the proportion of survey unit $i$ that is sampled and $p_i$ is a detection probability.
+Note that for ease of exposition we simply set $p_i = 1.0$ for analyses in this paper, but this constraint should be relaxed in the future if such estimates are to be used in conservation or management.
 
@@
   \label{eq:process}
-  Z_i & \sim & f(g^{-1}(\mu_i)),
+  [Z_i|\mu_i] & = & f(g^{-1}(\mu_i)),
 \end{eqnarray}
@@
  \label{eq:R}
-  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
+  [R_i|\nu_i] & = & \text{Bernoulli}(h^{-1}(\nu_i)),
 \end{eqnarray}
@@
 Here, $\beta_0^*$ is an intercept, ${\bf x}_i^*$ is a vector of covariates (in practice, these can be the same or different from ${\bf x}_i$), and $\boldsymbol{\eta}=\{ \eta_1,\eta_2,\hdots,\eta_S \}$ are spatially autocorrelated random effects.
-The connection between the two models (eqs.
+The connection between the two models (eqns.
 \ref{eq:mu} and \ref{eq:nu}) is the reliance of both on $\delta_i$.
-In eq.
+In eqn.
 \ref{eq:nu}, the matrix ${\bf B}$ describes this relationship.
 Previous preferential sampling models in the statistical literature have set $\textbf{B} = b \textbf{I}$, where $\textbf{I}$ is an identity matrix and $b$ is a parameter to be estimated.
-This is the approach we take in simulation study and in several of the models fit to real data.
-Another approach is to construct $\textbf{B}$ so it changes smoothly in space, as in \citep{RoyleBerliner1999}.
+This is the approach we take in our simulation study and in several of the models fit to real data.
+Another approach is to construct $\textbf{B}$ so it changes smoothly in space, as suggested (but apparently not implemented) by \citep{RoyleBerliner1999}.
 In this manner, the degree of preferential sampling could vary across the landscape; presumably, one would need fairly rich data to be able to model a spatially varying relationship.
@@
 \begin{eqnarray*}
-  \boldsymbol{\lambda} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\delta}), \text{ and} \\
+  \boldsymbol{\delta} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\delta}), \text{ and} \\
   \boldsymbol{\eta} & \sim & MVN(\textbf{0},\boldsymbol{\Sigma}_{\eta}),
@@
 To improve computational efficiency, we use a Gaussian Markov random field (GMRF) representation of the underlying Gaussian field \citep{LindgrenEtAl2011}, as described by \citet{ThorsonEtAl2015} (Appendix B).
+For increased stability we estimated only one $\kappa$ value (i.e.
+the same $\kappa$ parameter was used for $[{\bf Z}]$ and $[{\bf R}]$.
 
-\hspace{0.5in} To conduct statistical analysis of the preceding models, we use maximum marginal likelihood as implemented in Template Model Builder \citep[TMB; ][]{KristensenEtAl2015}.
-Interfaced with the R programming environment, the TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
+\hspace{0.5in} To conduct statistical analysis of the preceding models, we use maximum marginal likelihood as implemented in Template Model Builder \citep[TMB; ][]{KristensenEtAl2016}.
+Interfaced with the R programming environment, the TMB software uses a Laplace approximation to integrate out random effects, and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensen2016} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
 Requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.
@@
       \begin{eqnarray}
-        \textbf{X} & \sim & \textrm{MVN} (\textbf{0},\boldsymbol{\Sigma}).
+        [\textbf{X}] & = & \textrm{MVN} (\textbf{0},\boldsymbol{\Sigma}).
         \label{eq:cov}
@@
   \item Generate true abundance in each survey unit (grid cell), $N_i$, as a function of the covariate from step 1, together with additional spatially autocorrelated error.
-We generate a vector of such abundances as
+We generate a vector of such abundances from the distributions
       \begin{eqnarray}
-        N_i & \sim & \textrm{Poisson}(\exp(\beta_0 + X_i \beta_1 + \delta_i)),
+        [N_i] & = & \textrm{Poisson}(\exp(\beta_0 + X_i \beta_1 + \delta_i)),
       \end{eqnarray}
-      where the $\boldsymbol{\delta}$ are drawn in an identical manner to $\textbf{X}$ in step 1.  We set $\beta_0 = 2.0$ and drew $\beta_1 \sim \textrm{Uniform}(-0.5,0.5)$ for each simulation.
+      where the $\boldsymbol{\delta}$ are drawn in an identical manner to $\textbf{X}$ in step 1.
+We set $\beta_0 = 2.0$ and drew $\beta_1$ from a $\textrm{Uniform}(-0.5,0.5)$ distribution for each simulation.
   \item For each simulation, we selected 50 grid cells for sampling (i.e., set $R_i = 1.0$) according to 3 alternative survey configurations.
@@
       Here, $\eta_i$ is a spatially autocorrelated random effect drawn in an identical manner to $\textbf{X}$ in step 1, and $b$ describes the level of preferential sampling.
-Three levels are used: $b=0$, which makes the sampling model conditionally independent from the abundance process (i.e., no preferential sampling); $b=1$, which induces some dependence between the abundance and sampling process (i.e., moderate preferential sampling); $b=5$ which makes cells of high abundance much more likely to be sampled than low abundance cells (i.e., pathological preferential sampling).
-We drew $\beta_1^* \sim \textrm{Uniform}(-0.5,0.5)$ for each simulation replicate.
+Three levels are used: $b=0$, which makes the sampling model conditionally independent from the abundance process (i.e., no preferential sampling); $b=1$, which induces some dependence between the abundance and sampling process (i.e., moderate preferential sampling); $b=3$ which makes cells of high abundance much more likely to be sampled than low abundance cells (i.e., pathological preferential sampling).
+We drew $\beta_1^*$ from a  $\textrm{Uniform}(-0.5,0.5)$ distribution for each simulation replicate.
 An illustration of the consequences of different survey configurations is provided in Fig.
@@
   \item For each of the 3 sampling configurations, we generated a count dataset.
-In each case, we generated a count $Y_i$ for each cell selected for sampling (i.e., for cells where $R_i=1.0$) according to
+In each case, we generated a count $Y_i$ for each cell selected for sampling (i.e., for cells where $R_i=1.0$) from the distribution
       \begin{eqnarray*}
-        Y_i & \sim & \textrm{Binomial}(N_i,A).
+        [Y_i|N_i] & = & \textrm{Binomial}(N_i,A).
       \end{eqnarray*}
@@
   \item For each count data set, we fitted 2 different estimation models in TMB (see \textit{Preferential sampling model development}).
-In the first, the preferential sampling parameter $b$ was fixed to 0.0 (thus in effect assuming no preferential sampling).
-In the second, $b$ was estimated.
-  \item For each combination of simulation replicate, survey configuration, and estimation model, we recorded relative bias with respect to total estimated abundance, $\hat{N} = \sum \hat{N_i}$, where a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensenInPress} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
-This bias correction is also used to obtain abundance estimates in the bearded seal example
+In the first, the the models for $[{\bf R}]$ and $[{\bf N}]$ were configured to be independent.
+In the second, a single parameter $b$ was estimated, where ${\bf B} = b {\bf I}$ in eqn.
+(\ref{eq:nu}).
+  \item For each combination of simulation replicate, survey configuration, and estimation model, we recorded relative bias with respect to total estimated abundance, $\hat{N} = \sum \hat{N_i}$, where a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensen2016} is used to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
+This bias correction is also used to obtain abundance estimates in the bearded seal example.
+Estimates of $N$ are made relative to $N_i = Y_i + (1-A) \exp(\beta_0 + X_i \beta_1 + \delta_i)$, which includes an implicit finite population correction \citep{JohnsonEtAl2010}.
 \end{enumerate}

### 2017-04-24 manuscript/AppendixS2.tex
Restructured R files for package compilation and public consumption (including putting TMB .cpp code in the src directory for automatic compilation when package is loaded). Final changes to ms. before submission.
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

@@
 
+\rm
 \vspace{0.7cm}
@@
 
-In addition to the models reported in the paper, we attempted to fit two different models with a trend surface specification for the preferential sampling process.
-The models differed by whether or not predictive covariates were also modeled (the ice covariates described in the main article text), but otherwise the preferential sampling process was modeled similarly.
+We attempted to fit four models that estimated preferential sampling parameter(s) to bearded seal data.
+The models differed by whether or not habitat and landscape covariates were modeled, and how the preferential sampling effect was modeled.
 Specifically, when the probability $R_i$ that a location $i$ is selected for sampling is modeled as
@@
  \label{eq:R}
-  R_i & \sim & \text{Bernoulli}(\textrm{expit}(\nu_i))),
+  R_i & \sim & \text{Bernoulli}(\textrm{logit}^{-1}(\nu_i)),
 \end{eqnarray}
@@
 \end{eqnarray}
-we implemented a trend surface model by writing $\bf{B}$ as
+we considered two formulations for \textbf{B}.
+In the first, it was simply set to $b \textbf{I}$, where $\textbf{I}$ is an $(S \times S)$ identity matrix and $b$ is an estimated parameter.
+In the second, we implemented a trend surface model by writing $\bf{B}$ as
 \begin{eqnarray}
   \label{eq:B}
-  \textbf{B} & = & {\bf C}{\bf b}.
+  \textbf{B} & = & {\bf C}{\bf b} \textbf{I}.
 \end{eqnarray}
@@
 
-\hspace{0.5in} On the surface, there did not appear to be any numerical errors when fitting these trend surface models to bearded seal data.
-The Hessian was invertible, and incorporation of the trend surface model resulted in a much lower AIC score than the other models (AIC scores were 19 and 26 points better than the highest ranked AIC model reported in the paper for the trend surface models omitting and including covariate effects, respectively).
-However, when examining estimates graphically, it was clear that the trend surface models suffered from some parameter identification and/or aliasing issues (Fig.
+\hspace{0.5in} On the surface, there did not appear to be any numerical errors when fitting preferential sampling models to bearded seal data.
+The Hessian was invertible, and incorporation of the preferential sampling effects resulted in much lower AIC scores than the other models (AIC scores were close to 100 points better than the highest ranked AIC model omitting preferential sampling effects).
+However, when examining estimates graphically, it was clear that the preferential sampling models suffered from some aliasing issues (Fig.
 \ref{fig:bearded_RE_maps}).

### 2017-04-24 manuscript/pref_sampling_revision.tex
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

@@
 \item Species distribution models fitted to wildlife survey data often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.
-However, this assumption is likely violated in many cases when survey effort is non-randomized, leading to preferential sampling.
+However, this assumption may be violated when survey effort is non-randomized, leading to preferential sampling.
 
@@
 
-\item Using simulation, we show that even moderate levels of preferential sampling can lead to large (e.g.
+\item Using simulation, we show that moderate levels of preferential sampling can lead to large (e.g.
 $40\%$) bias in estimates of animal density and that our modeling approach can considerably reduce this bias.
+In contrast, preferential sampling did not appear to bias inferences about parameters informing species-habitat relationships (i.e.
+slope parameters).
 
 \item We apply our approach to aerial survey counts of bearded seals (\textit{Erignathus barbatus}) in the eastern Bering Sea.
-Models that included a preferential sampling effect led to lower estimates of abundance than models without, but the effect size of the preferential sampling parameter decreased in models that included explanatory environmental covariates.
+As expected, models with a preferential sampling effect led to lower abundance than those without.
+However, several lines of reasoning (better predictive performance, higher biological realism) led us to prefer models without a preferential sampling effect for this data set.
 
-\item When wildlife surveys are conducted without a well-defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling.
+\item When wildlife surveys are conducted without a well-defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling when estimating animal density or occurrence.
 Joint models, such as those described in this paper, can be used to test and correct for such biases.
-Predictive covariates are also useful for bias reduction, but ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
+However, such models can be unstable; ultimately the best way to avoid preferential sampling bias is to incorporate design-based principles such as randomization and/or systematic sampling into survey design.
 \end{enumerate}
@@
 Several authors have shown that non-randomized designs can have poorer predictive ability than randomized designs when used to estimate species distributions \citep[e.g.][]{EdwardsEtAl2006,AlbertEtAl2010}, but that the optimal design may depend on the question of interest.
-For instance, the optimal design may differ depending on whether the inferential goal is to maximize predictive ability or to estimate the relationship between species presence and an important habitat covariate \citep{AlbertEtAl2010}.
+For instance, the optimal design may differ depending on whether the inferential goal is to maximize predictive ability or to estimate the relationship between species presence and important habitat covariates \citep{AlbertEtAl2010}.
 
@@
   \label{eq:process}
-  [Z_i] & = & f(g^{-1}(\mu_i)).
+  [Z_i] & = & f(g^{-1}(\mu_i)),
 \end{eqnarray}
+where we use the bracket notation ($[\cdot]$) to denote a probability distribution (density) function.
 In this example, $Z_i$ denotes the state variable of interest (e.g.
 occupancy or abundance), $g()$ is a link function (e.g.
-probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, or spatially autocorrelated random effects.
-We use the bracket notation (e.g.
-$[Z_i]$) to denote a probability mass or probability density function (in the case of discrete or continuous distribution, respectively).
+probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, and spatially autocorrelated random effects.
+Spatially autocorrelated random effects are often included to allow animal density to vary smoothly in space and to account for residual patchiness not explained by predictive covariates \citep{Legendre1993,LichsteinEtAl2002}.
 In applications described in this paper, we write the intensity as
@@
 For occupancy, $f()$ would typically be Bernoulli, while the Poisson or negative binomial are typically choices for analysis of count data; common forms for $\delta_i$ include geostatistical specifications \citep{Cressie1993,DiggleEtAl1998}, Gaussian Markov random fields \citep[e.g.
-conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{Reich2006,Hughes2013}.
+conditionally autoregressive models;][]{RueHeld2005}, or low rank alternatives such as predictive process \citep{BanerjeeEtAl2008,LatimerEtAl2009} or restricted spatial regression models \citep{ReichEtAl2006,Hughes2013}.
 
 The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.
-However, it is usually impossible to observe the system perfectly even in locations where sampling occurs, so it is customary to include an observation model describing incomplete detection.
-For occupancy studies, the response variable $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
+However, it is usually impossible to conduct a complete census of all animals present in a surveyed location, so it is customary to include an observation model describing incomplete detection.
+For occupancy studies, the response variable is $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
 \begin{eqnarray}
@@
 \end{eqnarray}
- Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates.
+ Here, the detection probability $p_i$ is possibly a function of survey and observer specific covariates (e.g.
+visibility, observer experience).
 Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.
@@
 We use $\bfmu$ and $\bfnu$ to indicate the link-scale process of interest and the logit of the probability of sampling, respectively.
-As with $\bfmu$, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.
-We use the notation $Z_i$ when describing the state process in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.
+For animal abundance, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.
+We use the notation $Z_i$ when describing the process model in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.
 
@@
 
-Design-based estimation requires that surveys rely on a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.
+Design-based estimation requires that surveys adhere to a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.
 Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement.
@@
 This can be a crucial advantage in surveys covering large areas with frequent inclement weather.
-It also opens the door for using presence only, citizen science, or opportunistically collected data for estimation.
+It also opens the door for using presence-only, citizen science, or opportunistically collected data for estimation.
 
@@
 However, in many ecological applications, predictive covariates explain only a small portion of variation present in the data.
-If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally) , bias may still occur.
+If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally), bias may still occur.
 Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.
@@
 However, in the case of PS, this inclusion probability also depends on the value of the response associated with the sampling unit.
-Evidently, any approach to account for PS should also account for the dependence between the state variable of interest and the locations chosen for sampling.
+Evidently, any approach to account for PS should also account for the dependence between the process of interest and the locations chosen for sampling.
 
 Several authors have attempted model-based corrections for PS in the statistical literature.
-For Gaussian models in a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the locations that are chosen for sampling and the underlying random field of interest.
-In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity depended linearly on spatially-referenced random field values.
-For instance, writing observations of the spatial random field at a location $i$ as
+For Gaussian models on a continuous spatial domain, \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} jointly modeled the process of interest and the locations chosen for sampling.
+In particular, they expressed sampled locations as an inhomogeneous Poisson point process where the underlying log-scale intensity at a given location depended linearly on the process of interest at that location.
+For instance, writing observations of the spatial process (random field) at a location $i$ as
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $\mu_i$ and $\xi_i$ includes explanatory covariates.
+where $\mu_i$ and $\xi_i$ can both be written as a function of predictive covariates.
 Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g.
@@
 \ref{fig:PS_example}).
-\citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, where the log of a nonparameteric estimate of sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn.
+\citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, log sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn.
 (\ref{eq:Z_cont}), finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.
@@
 However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.
-Also, they require the linear predictor of the PS model to be written as a simple linear function of the the spatial process model for density.
+Also, they require the linear predictor of the PS model to be written as a simple linear function of the spatial process.
 In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.
@@
 \end{eqnarray}
- In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).
+  As with the process model, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$) and spatially autocorrelated random effects ($\boldsymbol{\eta}$ and $\boldsymbol{\delta}$).
 The predictive covariates ${\bf x}_i$ from eqn.
@@
 Note also that the  spatially autocorrelated random effects $\boldsymbol{\delta}$ are included in both eqns.
-(\ref{eq:mu}) and (\ref{eq:nu}), allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.
+(\ref{eq:mu}) and (\ref{eq:nu}), allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between sampling intensity and the process of interest.
 The spatially autocorrelated random effects $\bfeta$ are assumed independent of the $\bfdelta$.
-In practice, we find we often need to fix $\beta_0^*=0.0$ when random effects in eqn.
-(\ref{eq:nu}) are estimated to permit parameter identification.
 
  The formulation in eqn.
-(\ref{eq:nu}) is similar to the one previously proposed for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.
+(\ref{eq:nu}) is similar (but not identical) to one proposed by \citet{RoyleBerliner1999} for hierarchical multivariate models with spatial dependence.
 There are multiple ways of structuring ${\bf B}$ depending on the complexity of spatial dependence desired for the PS process \citep{RoyleBerliner1999}.
-For instance, setting ${\bf B} = {\bf 0}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no PS).
+For instance, setting ${\bf B} = {\bf 0}_{S \times S}$ (i.e.
+to an all-zero matrix) corresponds to an absence of spatial dependence (and thus no PS).
 Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear PS model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.
 Alternatively, we could allow the degree of PS to vary across the landscape.
-For instance, one could contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
-However, to our knowledge the identifiability of higher order models (such as trend surfaces) has not previously been investigated.
+For instance, one could contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by
+ \begin{eqnarray*}
+   \label{eq:ts}
+   b_{ii} & = & b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i,
+ \end{eqnarray*}
+ where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
+To our knowledge, the identifiability of higher order models (such as trend surfaces) has not been investigated.
 
 A comparison of the performance of models with different sets of constraints on ${\bf B}$ can serve as a test of PS.
-In particular, if one can demonstrate that models with ${\bf B}={\bf 0}$ perform similarly or better than models with ${\bf B} \ne {\bf 0}$, then PS is likely not worth modeling and inference can proceed using standard SDMs (i.e.
-not modeling sampling intensity).
+In particular, if one can demonstrate that models with ${\bf B}={\bf 0}$ perform similarly or better than models with ${\bf B} \ne {\bf 0}$, then PS is likely not worth modeling and inference can proceed using standard SDMs.
 
@@
 \ref{eq:R} and \ref{eq:nu}.
-Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, no covariate or spatially autocorrelated random effects), but differed in how the matrix ${\bf B}$ was parameterized.
-In the first, we set ${\bf B}=0$, so that surveyed locations were selected independently of the abundance generating process.
-For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=3$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the latent abundance in that unit.
+Each survey had $\boldsymbol{\beta}^*=\eta_i=0$ (that is, no covariate or spatially autocorrelated random effects on the probability of sampling), but differed in how the matrix ${\bf B}$ was parameterized.
+In the first, we set ${\bf B}=0$, so that surveyed locations were randomly selected (i.e.
+independent of abundance).
+For the second and third, we set ${\bf B}$ to be a diagonal matrix with entries $b=1$ and $b=3$, respectively, so that the probability of sampling a given survey unit (grid cell) was explicitly dependent on the spatial abundance process in that unit.
 We refer to these scenarios as moderate and pathological PS, respectively (see fig.
 \ref{fig:sim_R}).
-Simulations were configured so that $n=50$ of the 625 survey units were sampled; each survey was set to cover half of the target cell.
+Simulations were configured so that $n=50$ of the 625 survey units were sampled; each survey covered half of each target cell.
 
-We fitted two different models to each count dataset, both of which were provided the habitat covariate used (in part) to generate the data for which a log-linear coefficient $\beta$ was estimated.
+We fitted two different models to each count dataset.
 In the first model, the elements of \textbf{B} in eqn.
 (\ref{eq:nu}) were all set to zero.
-In this case, the abundance and sampling process submodels were independent, as is the case canonical SDMs (at least when fitted to presence-absence or count data).
+In this case, sampling intensity is assumed to be independent of underlying abundance, as implicitly assumed by most SDMs in the ecological literature.
 In the second model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix.
@@
 all animals present were assumed to be detected).
+Each model included the same habitat covariate that was used to generate the data as a log-linear predictor of abundance.
 
@@
   [Z_i | \mu_i] & = & \textrm{Poisson}(a_i \exp(\mu_i)) \\
-  [Y_i|Z_i] & = & \textrm{Binomial}(Z_i, A_i), \text{ where}
+{}  [Y_i|Z_i] & = & \textrm{Binomial}(Z_i, A_i), \text{ where}
 \end{eqnarray*}
@@
 
-We fitted a total of four models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1}$) to bearded seal count data using the same estimation framework as in the simulation study.
-Models varied by (i) whether or not habitat and landscape variables were used as predictors of bearded seal density ($cov=1$ and $cov=0$, respectively), and (ii) the form of PS ($b=0$ indicates no PS; $b=1$ indicates ${\bf B}=b{\bf I}$, where $b$ is an estimated parameter).
-We also attempted to fit models where the PS ${\bf B}$ matrix varied over the landscape using a trend surface specification, but parameter identification was suspect in these models so estimates are not reported here (see Appendix S2 for more information).
-When habitat and landscape variables were included, we used three log-linear predictors: linear and quadratic functions of sea-ice concentration, and distance from the southern ice edge.
+We fitted a total of six models ($M_{cov=0,b=0},M_{cov=0,b=1},M_{cov=1,b=0},M_{cov=1,b=1},M_{cov=1,b=ts},M_{cov=1,b=ts}$) to bearded seal count data using the same estimation framework as in the simulation study.
+Models varied by (i) whether or not habitat and landscape variables were used as predictors of bearded seal density ($cov=1$ and $cov=0$, respectively), and (ii) the form of PS ($b=0$ indicates no PS; $b=1$ indicates ${\bf B}=b{\bf I}$; $b=ts$ indicates a trend surface specification for PS, see eqn.
+\ref{eq:ts}).
+When habitat and landscape variables were included, we used the following log-linear predictors for the abundance process: linear and quadratic functions of sea-ice concentration, linear and quadratic functions of depth, distance from land, and distance from the southern ice edge.
+In contrast, distance from the Alaska mainland was the only predictor used for sampling intensity ($\bfnu$).
 Remotely sensed sea-ice data were obtained at a $25 \times 25$ km resolution from the National Snow and Ice Data Center, Boulder, CO, USA, as described by \citet{ConnEtAl2014}.
-Models for $\mu_i$ and $\nu_i$ both utilized spatially autocorrelated random effects with a Mat\'{e}rn covariance function between grid cell centroids (see Appendix S1 for further details).
-When covariates were included, they were included in both models (i.e.
-for $\mu_i$ and $\nu_i$).
+All measurements were made relative to grid cell centroids.
+Models for $\mu_i$ and $\nu_i$ both utilized spatially autocorrelated random effects with a Mat\'{e}rn covariance function (see Appendix S1 for further details).
+
+To compare and contrast models, we calculated marginal AIC and computed mean square prediction error.
+Prediction error was calculated using a Monte Carlo cross validation procedure.
+For a total of 100 replicates, we randomly selected 40/394 counts to be withheld for testing.
+Estimation proceeded using the remaining 354 counts for each of the six estimation models; mean squared error of observed and predicted counts was used as the scoring function for calculation of prediction error with each test data set \citep{HootenHobbs2015}.
+Note that in each cross validation analysis that we used $R_i$ values from the full data set (i.e.
+count predictions were made assuming these cells had been sampled).
 
@@
 \end{eqnarray}
-Predictions of abundance in locations $i$ and attendant standard errors were then estimated with respect to $\hat{N}_i = \exp(\mu_i)$.
+Predictions of abundance in locations $i$ and attendant standard errors were then estimated with respect to $\hat{N}_i = a_i \exp(\mu_i)$.
 
@@
 
-A substantial proportion of simulation replicates either did not converge or resulted in divergent estimates of $b$ (i.e.
-the parameter was estimated on a boundary when a flexible constraint was imposed).
+A substantial proportion of simulation replicates included models that were unstable (i.e.
+either did not converge or had $b$ estimated on a boundary; Appendix S2).
+In particular, 00.6\% of spatial models fit with $b$ fixed to zero were unstable, compared to  20.5\% when $b$ was estimated and the true value of $b$ was either zero (no preferential sampling) or 1.0 (moderate preferential sampling).
+Under pathological preferential sampling ($b=3.0$), 78.8\% of simulations failed to produce reliable results when $b$ was estimated.
+Here, we only present results for cases where models produced stable estimates.
 
-Estimates of cumulative animal abundance across simulated landscapes were median unbiased for both estimation methods when the sites selected for sampling were independent of animal density, though when $b$ was estimated, abundance estimates were more right skewed and had higher variance (fig.
+Estimates of cumulative animal abundance across simulated landscapes were relatively unbiased (-4\% for the canonical SDM and 5\% for the joint model) in the scenario where survey locations were selected randomly (i.e.
+independent of abundance; fig.
 \ref{fig:sim_bias}).
-Under moderate PS ($b=1$), estimation of the PS parameter $b$ led to a median bias of 5\%, while the canonical SDM model ignoring preferential sampling had a median bias of 40\%.
-Under pathological PS ($b=3$), the canonical SDM model was extremely biased while the joint model remained unbiased (fig.
+Under this scenario, abundance estimates had higher variance when $b$ was estimated.
+Under moderate PS ($b=1$), estimation of the PS parameter $b$ remained unbiased (-1\%), while the canonical SDM model ignoring preferential sampling had a mean bias of 48\%.
+Under pathological PS ($b=3$), the canonical SDM model was extremely biased (mean bias of 159\%) while the joint model had a relatively small positive mean bias of 6\% (fig.
 \ref{fig:sim_bias}).
+Monte Carlo error on bias of abundance estimates was 1-2\% for each scenario.
+In contrast to abundance, the slope of the regression parameter relating abundance to the simulated habitat covariate was unbiased for all scenarios (Appendix S2).
 
@@
 
-Marginal AIC strongly favored models with covariate effects, but for such models the presence of PS was equivocal (Table \ref{tab:aic}).
-Further intuition can be gained by examining estimates of the PS parameter, $b$.
-For the PS model without predictive covariates,  $\hat{b}=0.27$ (SE 0.11), and for the model with predictive covariates, $\hat{b}=0.19$ (SE 0.13).
-Thus, it appeared that including predictive covariates decreased the PS effect size, as suggested by \citet{PatiEtAl2011}.
-Estimates of abundance were substantially higher for models without a PS effect, with the non-PS model having a 49\% higher estimate when covariates were not modeled, and a 19\% higher estimate when covariate effects were included.
+In models where one was estimated, the preferential sampling effect was large and positive (e.g.
+$\hat{b}=9.7$ for the $M_{cov=1,b=1}$ model).
+An estimate of this magnitude implies that we effectively sampled all locations where bearded seals were present, which seems biologically implausible given that survey effort was allocated to maximize spatial coverage rather than to target specific pockets of high bearded seal density.
+Further, these point estimates well exceeded estimates from the simulation study where moderate or even pathological levels of preferential sampling were employed.
+Thus, despite better marginal AIC scores for models with a preferential sampling effect than without (Table \ref{tab:aic}), we regard results of the joint preferential sampling model as unreliable in this example.
+This view is further supported by examination of cross-validated mean square prediction error, which suggests better predictive performance of models without a preferential sampling effect (Table \ref{tab:aic}).
 
-Note that unlike the other models, $M_{cov=1,b=0}$ predicted anomalously high bearded seal abundance in the extreme southern portion of the study area where sea ice was absent (where there was no habitat for seals).
-Thus, while we present original likelihood and AIC values to permit direct comparison with other models, we refitted $M_{cov=1,b=0}$ to produce an estimate of apparent abundance without this feature.
-Specifically, we refitted the model with 20 pseudo-absences in this portion of the study area to better inform abundance-covariate relationships.
+Predictive performance was roughly similar for models with and without landscape and habitat covariates (Table \ref{tab:aic}).
+However, this does not necessarily indicate that such covariates are ecologically unimportant, as spatial random effects and habitat effects can be confounded in spatial regression models \citep{HodgesReich2010}.
 
-The fact that models with and without a PS effect garner approximately equal weight suggests a need to account for PS when producing abundance estimates from this data set.
-A model averaged estimate calculated using AIC machinery \citep{BurnhamAnderson2002} is 54854 (SE 9351), which is 7.5\% less than the estimate assuming no PS.
-Notably, the standard error of the model averaged estimate was 79\% higher than the model assuming no PS.
+Finally, we note that estimates of bearded seal abundance in individual cells can become negative when the preferential sampling effect is strongly positive using our estimation approach.
+This is because the bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensen2016} used to correct for the nonlinear transformation of log-scale abundance does not prevent estimates from becoming negative.
+Post hoc experimentation with different priors on the preferential sampling parameter $b$ indicated that the frequency and magnitude of negative estimates declined as estimates of $b$ decreased towards biologically plausible values.
+This phenomenon could be avoided altogether by conducting a fully Bayesian analysis.
 
@@
 animal abundance) when sampling is non-randomized.
-In simulations, estimators were increasingly positively biased as PS increased.
-When PS was present, we were able to substantially reduce bias by conducting estimation under a framework where the state variable of interest and the sites chosen for sampling were jointly modeled under a dependent covariance structure.
-In absence of PS, simulations suggest that this structure results in lower precision than a model without a PS effect; thus, the need to account for PS reduces the quality of inference.
+In simulations, naive estimators were increasingly positively biased as PS increased.
+When PS was present, we were able to substantially reduce bias by conducting estimation under a framework where the state variable of interest and the sites chosen for sampling were jointly modeled using a dependent covariance structure.
+In absence of PS, simulations indicated that models with a PS effect produced less precise estimates.
+Thus, there was a real cost for attempting to account for PS within a model-based framework.
+Fortunately, PS only seemed to be a problem if density or abundance is of interest.
+Estimates of slope parameters governing the relationship of relative abundance to habitat and landscape parameters did not appear to be affected (Appendix S2).
 
@@
 Thus, there are theoretical reasons to sample more in high density areas than in low-density areas.
-One solution is to sample more in locations where covariates suggest abundance will be higher; in this case, sampling will not be preferential as long as the covariate used to select sampling locations is also used in the species distribution model fitted to the data (fig.
+One solution is to sample more in locations where covariates suggest abundance will be higher; in this case, sampling will not be preferential as long as the covariate used to select sampling locations is also used as an explanatory covariate in the SDM (fig.
 \ref{fig:PS_example}).
@@
 
-When applied to bearded seal count data, approximately equal support was given to models with and without a PS effect.
-The PS effect size was estimated to be positive and to produce considerably lower abundance estimates than models without a PS effect.
-Differences between apparent abundance estimates decreased when covariates were added to model structure, supporting previous theoretical results \citep{DiggleEtAl2010,PatiEtAl2011} that covariates serve to decrease the conditional dependence between site selection and the state variable of interest.
-However, in our data set, adding covariates did not eliminate evidence of PS.
-Accounting for PS in a model averaging framework led to a moderate decrease in our apparent abundance estimate for bearded seals in this region, and markedly decreased precision.
-As in our simulations, the need to account for PS thus appeared to have a real cost in terms of variance inflation.
+When applied to bearded seal count data, models with a preferential sampling effect produced estimates that were considerably lower abundance estimates than models without a PS effect (Table \ref{tab:aic}).
+However, such estimates were biologically unreasonable despite being favored by AIC (Table \ref{tab:aic}).
+In particular, estimates of preferential sampling parameters were so extreme as to indicate we had visited all survey units with substantial numbers of bearded seal.
+This was clearly unreasonable given the manner in which survey effort had been allocated, which sought to maximize spatial coverage while ignoring the expected spatial distribution of animals at finer scales.
+Given this finding, we caution against using likelihood-based model selection diagnostics for choosing between models with and without preferential sampling effects.
+Metrics based on predictive ability (e.g.
+cross-validated mean square prediction error) seem more useful for this purpose; in our bearded seal example, models without a preferential sampling effect had better predictive ability (Table \ref{tab:aic}).
+We also caution ecologists to be aware of biologically implausible values for $\hat{b}$.
+In particular, ``high" values (e.g.
+$>3.0$) seem improbable, even when sampling is nonrandom (fig.
+\ref{fig:PS_example}).
+For scientific surveys that are nonrandom but avoid explicit targeting (e.g.
+in the bearded seal example), values greater than 1.0 seem unlikely.
+Finally, values less than zero indicate that survey effort is concentrated in places where abundance is anomalously low, which seems unlikely to occur in practice.
 
-We attempted to fit models to bearded seal data where the degree of PS changed over the landscape, as \citet{RoyleBerliner1999} suggested might be possible (but never demonstrated) for multivariate spatial models.
-However, such models led to difficulties with parameter identification in our bearded seal application (see Appendix S2).
-At this time, we are not sure whether these models can be fit reliably, and suggest limiting initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
+We fitted models to bearded seal data where the degree of PS changed over the landscape, as \citet{RoyleBerliner1999} suggested might be possible (but never demonstrated) for multivariate spatial models.
+Such models led to similar inferences to models with a simple, linear effect for PS.
+However, given the potential for PS model instability, we currently suggest limit initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
 Further research into the viability of including spatial variation in the ${\bf B}$ matrix certainly appears warranted.
-It would also be worthwhile to investigate whether parameter identification varies as a function of the support of the state variable being modeled (e.g.
+It would also be worthwhile to investigate whether the reliability of inference varies as a function of data type (e.g.
 binary vs.
-count data), and whether parameters remain identifiable when nuisances processes such as incomplete detection are modeled.
+count data), survey type (e.g.
+line transects vs.
+point counts), and whether parameters remain identifiable when nuisances processes such as incomplete detection are modeled.
 
-The models we have developed here are specific to spatial models with discrete support, as when data are gathered at a plot level, or aggregated prior to analysis.
+The models we have developed here are specific to spatial models with discrete support, as when data are gathered at the plot level, or aggregated prior to analysis.
 However, it should be possible to extend our approach to continuous space.
@@
 Our conception of PS is related, but not equivalent to ``sample selection bias" \citep[e.g.][]{PhillipsEtAl2009} in presence-only models.
-In such models, absence of a species at a given site is never directly observed.
+In such models, species absences are never directly observed.
 To draw inference about space use, it is thus necessary to produce a background sample representing the range of locations and habitats that could have been sampled.
@@
 \begin{table}[ht]
-\caption{A summary of model selection results and estimated abundance for the four models fitted to bearded seal counts.
-The models include formulations with or without predictive covariates ($cov=1$ or 0, respectively) , and with or without the preferential sampling parameter $b$ estimated ($b=1$ or 0, respectively) .
-All models included spatially autocorrelated random effects on log-scale abundance intensity.
-Shown are the log integrated likelihood, the number of fixed effect parameters, $\Delta \textrm{AIC}$, AIC model weights, and estimated apparent abundance over the landscape ($\hat{N}$) together with a Hessian-based standard error estimate.
+\caption{A summary of model selection results and estimated abundance for the six models fitted to bearded seal counts.
+The models include formulations with or without predictive covariates ($cov=1$ or $cov=0$, respectively) , and with different specifications for preferential sampling ($b=0$: fixed to zero; $b=1$: a single $b$ parameter estimated; $b=ts$: a trend surface model for $b$).
+All models included spatially autocorrelated random effects on log-scale abundance intensity and on the probability of sampling.
+Shown are the log integrated likelihood, the number of fixed effect parameters, cross-validated mean square prediction error (``CV-MSPE"), $\Delta \textrm{AIC}$, and estimated apparent abundance over the landscape ($\hat{N}$) together with a Hessian-based standard error estimate.
+Estimates are ranked in order of decreasing predictive ability (as determined by increasing Median MSE).
 }
@@
   \hline
-  Model & Log likelihood & Params. & $\Delta \textrm{AIC}$ & Wgt & $\hat{N}$(SE) \\
+  Model & Log likelihood & Params. & CV-MSPE (SE)& $\Delta \textrm{AIC}$ & $\hat{N}$(SE) \\
   \hline
- $M_{cov=1,b=1}$ & -2808.3 & 10 & 0.0 & 0.65 & 45216 (4534) \\
- $M_{cov=1,b=0}$ & -2809.9 & 9 &  1.2 & 0.35 & 59313 (5231)  \\
- $M_{cov=0,b=1}$ & -2822.6 & 4 & 16.7 & 0.00 & 39397 (2931) \\
- $M_{cov=0,b=0}$ & -2829.1 & 3 & 27.5 & 0.00 & 54001 (4243)\\
+ $M_{cov=0,b=0}$ & -852.7 & 5 & 87.2 & 99.1 & 70738 (6988) \\
+ $M_{cov=1,b=0}$ & -847.6 & 12 & 88.3 & 103.0 & 66989 (20374) \\
+ $M_{cov=1,b=ts}$ & -794.1 & 15 & 99.5 & 1.8 & 40925 (4403) \\
+ $M_{cov=1,b=1}$ & -795.1 & 13 & 105.3 & 0.0 & 40656 (3664) \\
+ $M_{cov=0,b=ts}$ & -802.7 & 8 &  110.7 & 5.2 & 43520 (2770)  \\
+ $M_{cov=0,b=1}$ & -803.0 & 6 & 116.2 & 1.9 & 43232 (2778) \\
   \hline
@@
 However, the sample in (A) is non-preferential if the covariate is also used to estimate the species-covariate relationship.
-In contrast, the sample in (B) is preferential and can lead to biased inference.}
+In contrast, the sample in (B) is preferential and can lead to biased inference about the intercept of the covariate-abundance relationship (and hence to estimates of absolute abundance).}
 \label{fig:PS_example}
@@
 First, true abundance (C) is generated as a function of a spatially autocorrelated covariate (A) and a spatially autocorrelated random effect (B).
-Second, counts are generated for three different types of surveys, including a simple random sample ($b=0$; D) and surveys with moderate ($b=1$; E) or pathological ($b=5$; F) levels of preferential sampling.
+Second, counts are generated for three different types of surveys, including a simple random sample ($b=0$; D) and surveys with moderate ($b=1$; E) or pathological ($b=3$; F) levels of preferential sampling.
 Finally, spatially explicit estimates of abundance are generated using a traditional SDM (with $b$ set to 0.0) to each of the count datasets (G-I).
-In this particular simulation replicate, cumulative abundance was underestimated by 18\% when $b=0$, overestimated by 17\% when $b=1$, and overestimated by 293\% when $b=5$.
+In this particular simulation replicate, cumulative abundance was overestimated by 9\% when $b=0$, overestimated by 79\% when $b=1$, and overestimated by 308\% when $b=3$.
 For a summary of bias over 500 simulation replicates, see fig.
@@
 \caption{Expected relationship between the probability of a survey unit being selected for sampling and its abundance residual in the simulation study.
-The base case $b=0$ represents simple random sampling, while $b=1$ and $b=5$ represent moderate and pathological levels of preferential sampling, respectively.
+The base case $b=0$ represents simple random sampling, while $b=1$ and $b=3$ represent moderate and pathological levels of preferential sampling, respectively.
 Also shown are is the realized distribution (smoothed histogram) of abundance residuals among survey units in the simulation study, scaled to fit in the plot margins (solid black line).}
@@
 \includegraphics[width=170mm]{bias.pdf}
-\caption{Relative proportional error in abundance from the simulation experiment as computed with respect the posterior mode with a bias correction.
-Each boxplot summarizes the distribution of relative proportional error as a function of the type of sampling, including simple random sampling ($b=0$), moderate preferential sampling ($b=1$), and pathological preferential sampling ($b=5$).
+\caption{Relative proportional error in abundance from the simulation experiment computed with respect to bias-corrected marginal maximum likelihood estimates.
+Each boxplot summarizes the distribution of relative proportional error as a function of the type of sampling, including simple random sampling ($b=0$), moderate preferential sampling ($b=1$), and pathological preferential sampling ($b=3$).
 Results vary by the type of estimation model; in the ``independent" model, $b$ is set to 0.0; in the ``joint" model, $b$ is estimated.
@@
 In (A), tan shading denotes land, unsurveyed grid cells appear in dark gray, and counts appear in a white-blue spectrum.
-Apparent bearded seal abundance estimates (B) are presented from the model with the lowest integrated AIC score, which included covariate effects but no preferential sampling effect.
+Apparent bearded seal abundance estimates (B) are presented from the model with the best predictive ability (according to cross-validation), which was a simple spatial surface model without predictive covariates or a preferential sampling effect (model $M_{cov=0,b=0}$ from Table \ref{tab:aic}).
 Apparent abundance estimates are uncorrected for imperfect detection or species misclassification.}
@@
 \end{document}
-
-%http://www.plosone.org/article/info:doi/10.1371/journal.pone.0036527

### 2017-04-24 manuscript/pref_sampling_revision2.tex
This commit corresponds to code and manuscript status when the manuscript was accepted at MEE. It represents the first commit where cross validation was used in the sim example.

--- 2017-04-24 manuscript/pref_sampling_revision.tex
@@
 
-\def\volumeyear{2016}
+\def\volumeyear{2017}
 \def\VOC{VO}
@@
 \begin{document}
-\runningheads{P.~B.~Conn, J.~T.~Thorson, \& D.~S.~Johnson}{Preferential sampling in species distribution models}
+\runningheads{P.~B.~Conn, J.~T.~Thorson, \& D.~S.~Johnson}{Preferential sampling in spatial distribution models}
 
@@
 
-\title{Confronting preferential sampling in wildlife surveys: diagnosis and model-based triage \footnotemark[2]}
+\title{Confronting preferential sampling when analyzing population distributions: diagnosis and model-based triage \footnotemark[2]}
 
@@
 \begin{enumerate}
-\item  Wildlife surveys are often used to estimate the density, abundance, or distribution of animal populations.
+\item  Population surveys are often used to estimate the density, abundance, or distribution of natural populations.
 Recently, model-based approaches to analyzing survey data have become popular because one can more readily accommodate departures from pre-planned survey routes and construct more detailed maps than one can with design-based procedures.
 
-\item Species distribution models fitted to wildlife survey data often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.
+\item Spatial models for population distributions often make the implicit assumption that locations chosen for sampling and animal abundance at those locations are conditionally independent given modeled covariates.
 However, this assumption may be violated when survey effort is non-randomized, leading to preferential sampling.
 
-\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in species distribution models fitted to count data.
-The approach works by jointly modeling wildlife state variables and the locations selected for sampling and specifying a dependent correlation structure between the two models.
+\item We develop a hierarchical statistical modeling framework for detecting and alleviating the biasing effects of preferential sampling in spatial distribution models fitted to count data.
+The approach works by specifying a joint model for population density and the locations selected for sampling, and specifying a dependent correlation structure between the two processes.
 
@@
 
-\item When wildlife surveys are conducted without a well-defined sampling frame, ecologists should recognize the potentially biasing effects of preferential sampling when estimating animal density or occurrence.
+\item When population surveys break from traditional scientific survey design principles, ecologists should recognize the potentially biasing effects of preferential sampling when estimating population density or occurrence.
 Joint models, such as those described in this paper, can be used to test and correct for such biases.
@@
 
-\keywords{count data, preferential sampling, spatial autocorrelation, species distribution model}
+\keywords{count data, population survey, preferential sampling, spatial autocorrelation, spatial distribution model, species distribution model}
 
@@
 
-Surveys of unmarked animal populations are often used to estimate abundance and occurrence of animal populations and to predict species distributions, enterprises central to conservation, ecology, and management.
-For studies of abundance, researchers historically relied on design-based statistical inference \citep[e.g.][]{Cochran1977}, which requires adoption of a pre-defined sampling frame (e.g.
-using systematic random sampling, stratified random sampling, or some variant thereof).
-Designing animal surveys is relatively straightforward in such applications, and unbiased point and variance estimators are available.
-Recently, however, there has been a surge in research describing model-based procedures for estimating abundance, density, and occupancy from surveys of unmarked animals, including N-mixture and Dail-Madsen models for repeated point counts \citep{Royle2004a,DailMadsen2011}, occupancy models for presence-absence surveys \citep{MacKenzie2002,JohnsonEtAl2013}, and various model-based formulations for distance-sampling data \citep{HedleyBuckland2004,JohnsonEtAl2010,MillerEtAl2013}.
-In such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g.
+Ecologists often fit statistical models to data from surveys of natural populations (e.g., plants, animals) to explain variation in species distributions and to predict presence or abundance in unsampled areas.
+Such models may serve as the basis for testing ecological theory \citep{HilbornMangel1997,Austin2007}, for directed conservation and management \citep{WilliamsEtAl2002,NicholsWilliams2006}, or for monitoring biological diversity \citep{PereiraCooper2006,YoccozEtAl2001}.
+Depending on the application, data quality can vary widely, ranging from scientific survey data to opportunistic records of species presence reported by the public.
+
+Where possible, researchers often use sound survey design principles (e.g.
+randomization) to ensure unbiased inference when implementing scientific surveys.
+For instance, under protocols such as simple, systematic, or stratified random sampling, design-based estimators of average density or presence-absence over large survey regions are guaranteed to be unbiased and avoid making assumptions about the factors influencing density or distribution \citep{Cochran1977,BucklandEtAl2000}.
+However, spatial resolution is typically poor for such estimators, limiting the breadth of ecological and management questions that can be addressed.
+Further, for logistical reasons, many ecological surveys simply cannot afford to be this rigorous in their design.
+For example, even the North American Breeding Bird Survey uses roads and other non-random features that necessitate use of alternate estimation procedures \citep{SauerLink2011}.
+
+Over the last decade, there has thus been a huge surge in research describing model-based procedures for estimating abundance, density, or presence-absence from population survey data.
+In contrast to design-based procedures, model-based procedures make explicit assumptions about how populations vary over space, both in terms of explanatory variables and often in terms of distribution.
+For instance, in such applications, it is common to use habitat or environmental covariates together with spatial effects (e.g.
 via trend surfaces or spatial random effects) to predict density or distributions across the landscape.
-We shall refer to the amalgam of model-based approaches for making spatially explicit inference about animal populations as ``species distribution models" \citep[SDMs; {\it sensu}][]{ElithLeathwick2009}, even though this term is more often used to refer to animal occurrence than it is to density or abundance.
+This approach can be useful for fine-scale management, identifying range shifts, and comparing the fit (or lack thereof) of alternative models that express different ecological hypotheses \citep{Austin2007}.
+We shall refer to the amalgam of model-based approaches for making spatially explicit inference about natural populations as ``spatial models for population distributions" (SMPDs).
+Ultimately, the quality of inference from SMPDs depends on several factors, including (i) how well the fitted model approximates reality, and (ii) the quality of the data \citep{BucklandEtAl2000}.
 
-One of the main advantages of using SDMs is that one is less beholden to predetermined sampling frames and can potentially use data from non-randomized designs or opportunistic sampling to make inferences about animal populations \citep{JohnsonEtAl2010}.
+The radiation of model-based approaches has proceeded at different rates and pathways in different ecological modelling subfields, which are often linked to the data being modeled.
+For instance, species distribution models \citep[SDMs; ][]{ElithLeathwick2009} are often used to predict species presence as a function of habitat variables and encompass both statistical (e.g.
+generalized linear or generalized additive models) and machine learning algorithms.
+The term SDM also connotes analysis of presence-only data (as available from museum collection or citizen science programs for instance).
+In the field of population assessment, statistical models are almost exclusively used, often with observation models tailored to the sampling approach (e.g.
+presence-absence surveys, distance sampling, quadrat samples, point counts).
+In this paper, our perspective is one shaped more heavily by the population assessment literature, though we expect our general approach to be applicable to a wider variety of SMPDs.
+
+One of the main advantages of using SMPDs is that one can potentially use data from non-randomized designs or opportunistic sampling to make inferences about animal populations \citep{JohnsonEtAl2010}.
 Several authors have shown that non-randomized designs can have poorer predictive ability than randomized designs when used to estimate species distributions \citep[e.g.][]{EdwardsEtAl2006,AlbertEtAl2010}, but that the optimal design may depend on the question of interest.
@@
 In a recent paper, \citet{DiggleEtAl2010} emphasized that spatially explicit statistical models can easily provide biased estimates when sampling disproportionately targets locations where the response of interest is higher (or lower) than expected given a particular set of explanatory covariates.
-In the context of SDMs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance.
+In the context of SMPDs, this might occur if sampling disproportionately occurs in locations where animals are known to be present or of high abundance.
 For example, if volunteer inventory participants have access to multiple sites with similar covariate values, bias might arise if they consistently choose sites where species are thought or known to be present.
 Bias might also arise if survey effort is higher near bases of operations or features such as roads, particularly if animal abundance is higher (or lower) near these locations than elsewhere in the landscape.
-In our view, this phenomenon, termed ``preferential sampling," is potentially much more problematic, because estimates of animal abundance and occurrence from SDMs could be severely overstated.
+In our view, this phenomenon, termed ``preferential sampling," is potentially much more problematic, because estimates of animal abundance and occurrence from SMPDs could be severely overstated.
 
-In this article, we explore potential for bias in SDMs resulting from preferential sampling (hereafter, PS) and describe several model-based approaches for detecting and correcting for such biases.
+In this article, we explore potential for bias in SMPDs resulting from preferential sampling (hereafter, PS) and describe several model-based approaches for detecting and correcting for such biases.
 We start by describing a common currency for notation and basic model structures considered in this paper.
@@
 
-We focus here exclusively on discrete space (areal) models for animal encounter data as these seem to be the dominant form used in design and analysis of animal population surveys, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.
-We suppose that the investigator intending to fit a SDM to animal encounter data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are selected for sampling (call the set of sampled locations $\mathcal{S}$).
+We focus here exclusively on discrete space (areal) models for population survey data, although we note that PS is likely to affect analyses similarly regardless of the choice of spatial domain.
+We suppose that the investigator intending to fit a SMPD to survey data breaks their study area up into $S$ survey units (label these $U_1, U_2, \hdots, U_S$), of which $n$ are sampled (call the set of sampled locations $\mathcal{S}$).
 Each survey unit $i$ is assigned a vector of covariates, ${\bf x}_i$, and an indicator $R_i$ that takes on the value 1.0 if location $i$ is sampled (i.e.
 if $U_i \in \mathcal{S}$), and is 0 otherwise.
-To formulate a ``traditional" SDM, one could then write animal abundance or occurrence as a stochastic realization of a probability mass function $f()$:
+To formulate a ``traditional" SMPD, one could then write abundance or occurrence as a stochastic realization of a probability mass function $f()$:
 \begin{eqnarray}
@@
 probit or logit for occupancy, log for count data), and $\mu_i$ is a link-scale intensity value which can itself be written as a function of habitat covariates, regression coefficients, and spatially autocorrelated random effects.
-Spatially autocorrelated random effects are often included to allow animal density to vary smoothly in space and to account for residual patchiness not explained by predictive covariates \citep{Legendre1993,LichsteinEtAl2002}.
+Spatially autocorrelated random effects are often included to allow populations to vary smoothly in space and to account for residual patchiness not explained by predictive covariates \citep{Legendre1993,LichsteinEtAl2002}.
 In applications described in this paper, we write the intensity as
@@
 The model for $Z_i$ describes variation in the process of interest and is often described as the ``process'' model.
-However, it is usually impossible to conduct a complete census of all animals present in a surveyed location, so it is customary to include an observation model describing incomplete detection.
-For occupancy studies, the response variable is $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
+However, it is usually impossible to conduct a complete census of all individuals present in a surveyed location, so it is customary to include an observation model describing incomplete detection.
+For occupancy (presence/absence) studies, the response variable is $Y_i = 1$ if the species of interest is detected and is 0 otherwise, and is modeled with a Bernoulli distribution \citep{Royle2008}:
 \begin{eqnarray}
@@
 visibility, observer experience).
-Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$.
+Replicate surveys of the same sampling unit provide the necessary information to estimate $p_i$ \citep{MacKenzie2002}.
 For count surveys, a possible model is
@@
 We use $\bfmu$ and $\bfnu$ to indicate the link-scale process of interest and the logit of the probability of sampling, respectively.
-For animal abundance, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.
+For population abundance, this would mean $[Z_i] = f(\exp(\mu_i))$, whereas the distribution of sampling locations can be represented as $[R_i] = f(\text{logit}^{-1}(\nu_i))$, where $\mu_i$ and $\nu_i$ can be written as functions of regression coefficients, habitat covariates, and spatially autocorrelated random effects.
 We use the notation $Z_i$ when describing the process model in general terms, but often switch to the conventional notation $N_i$ when animal abundance is the explicit focus of interest.
@@
 
-Design-based estimation requires that surveys adhere to a pre-planned survey design selected probabilistically from an underlying sampling frame \citep{Cochran1977}.
-Although randomization can increase legitimacy, model-based estimation is not formally subject to this requirement.
-For instance, investigators can reallocate sampling effort if weather or logistics preclude surveying in a desired location and still generate model-based estimates of abundance or occupancy.
+Design-based estimation requires that surveys adhere to a pre-planned survey design selected probabilistically from an underlying sampling frame to ensure unbiasedness \citep{Cochran1977}.
+Although randomization can reduce potential for bias, researchers often view SMPDs in a different light, where the data gathering procedure is not important as long as the models being fitted are formulated well and produce reasonable predictions (see \textit{Discussion} for more on this topic).
+Under this view, investigators can reallocate effort if weather or logistics preclude surveying in a desired location and still generate model-based estimates of abundance or occupancy.
 This can be a crucial advantage in surveys covering large areas with frequent inclement weather.
@@
 
-However, the manner in which effort is allocated can potentially have profound influence on SDM estimator performance.
+However, the manner in which effort is allocated can potentially have profound influence on SMPD estimator performance.
 Preferential sampling (PS) arises when the locations being sampled and the process of interest (e.g.
@@
 
-As in the pollution example, having good explanatory covariates may also reduce bias when fitting SDMs to animal encounter data under PS (e.g.
+As in the pollution example, having good explanatory covariates may also reduce bias when fitting SMPDs to population survey data under PS (e.g.
 fig.
@@
 If the locations selected for sampling are a function of some unmodelled factor related to abundance (intentionally or unintentionally), bias may still occur.
-Despite the clear potential for bias in SDMs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SDMs.
+Despite the clear potential for bias in SMPDs, there are few examples where PS \citep[\textit{sensu}][]{DiggleEtAl2010} is discussed with regard to SMPDs.
 One exception is
-\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SDMs to data obtained using nonrandomized designs.
+\citet{ChakrabortyEtAl2010}, who acknowledged the likely presence of PS when fitting SMPDs to data obtained using nonrandomized designs.
 However, they did not attempt to account for PS in their models.
@@
 \end{eqnarray}
-the log-scale relative intensity of sampling locations at $i$ ($\nu_i$) could be written as
+the log-scale relative intensity of sampling locations at $i$ ($\kappa_i$) could be written as
 \begin{eqnarray}
-   \nu_i & \propto & \xi_i + b \mu_i,
+   \kappa_i & \propto & \xi_i + b \mu_i,
    \label{eq:p_cont}
@@
 Here, the parameter $b$ describes the level of PS; $b=0$ implies no PS, $b>0$ implies a greater level of sampling in locations where the spatial process (e.g.
-animal density) is high, and $b<0$ implies greater sampling where the spatial process is low.
-Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that ``.
+population density) is high, and $b<0$ implies greater sampling where the spatial process is low.
+Importantly, when explanatory covariates are used in models for $\mu_i$ and $\xi_i$, \citet{PatiEtAl2011} show that accounting for preferential sampling  ``.
 .
 .
-accounting for informative sampling is only necessary when there
+is only necessary when there
 is an association between the spatial surface of interest and the sampling density that cannot be
@@
 \ref{fig:PS_example}).
-\citet{PatiEtAl2011} also consider a simpler, plug-in based estimator, log sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn.
+\citet{PatiEtAl2011} also consider a simpler, plug-in based estimator in which log sampling density (specifically, a two dimensional kernel density estimate) is used as an additional fixed effect in eqn.
 (\ref{eq:Z_cont}), finding that this approach helped reduce bias associated with PS, but did not perform as well as the full joint model.
@@
 
-Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.
-Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurrence) and the locations chosen for sampling.
+Given these limitations, our present task is to generalize PS models to the types of data more typical of SMPDs, and to allow the degree of PS to vary across the landscape.
+Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (abundance or occurrence) and the locations chosen for sampling.
 For the abundance process model, we start with eqn.
@@
 (\ref{eq:mu}).
-Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is selected for sampling, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
+Next, recalling that $R_i$ is a binary indicator taking on the value 1.0 if survey unit $i$ is sampled, and is 0.0 otherwise, we model $R_i$ using a Bernoulli distribution:
 \begin{eqnarray}
@@
 Alternatively, we could allow the degree of PS to vary across the landscape.
-For instance, one could contemplate a trend surface model for PS by specifying a diagonal matrix for ${\bf B}$, with entries given by
+For instance, one could contemplate a trend surface model for PS by specifying a matrix for ${\bf B}$, with diagonal entries given by
  \begin{eqnarray*}
    \label{eq:ts}
-   b_{ii} & = & b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i,
+   B_{ii} & = & b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i
  \end{eqnarray*}
- where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
+ and $B_{ij}=0$ for $i \ne j$.
+Here, $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.
 To our knowledge, the identifiability of higher order models (such as trend surfaces) has not been investigated.
@@
 A comparison of the performance of models with different sets of constraints on ${\bf B}$ can serve as a test of PS.
-In particular, if one can demonstrate that models with ${\bf B}={\bf 0}$ perform similarly or better than models with ${\bf B} \ne {\bf 0}$, then PS is likely not worth modeling and inference can proceed using standard SDMs.
+In particular, if one can demonstrate that models with ${\bf B}={\bf 0}$ perform similarly or better than models with ${\bf B} \ne {\bf 0}$, then PS is likely not worth modeling and inference can proceed using standard SMPDs.
 
@@
 In this case, sampling intensity is assumed to be independent of underlying abundance, as implicitly assumed by most SDMs in the ecological literature.
-In the second model, we included an explicit connection between the distribution of animal abundance and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix.
+In the second model, we included an explicit connection between the abundance distribution and the sampling process by setting $\textbf{B}=b\textbf{I}$, where $b$ is an estimated parameter, and \textbf{I} is an identity matrix.
 For ease of exposition, we set $p_i = 1.0$ for all simulations (i.e.
@@
 This is also the location of one of the primary airports used to prosecute surveys (Nome, Alaska, USA).
-Higher survey coverage in areas of high bearded seal density could potentially lead to positive bias in apparent abundance owing to PS.
+Although we would not anticipate a large effect, higher survey coverage in areas of high bearded seal density could potentially lead to positive bias in apparent abundance owing to PS.
 
@@
 \end{eqnarray*}
-$A_i$ defines the proportion of grid cell $i$ that is sampled, $a_i$ gives the proportion of grid cell $i$ that is composed of salt water habitat, and $\mu_i$ is defined in eqn. (\ref{eq:mu}).
+$a_i$ gives the proportion of grid cell $i$ that is composed of salt water habitat, $A_i$ defines the proportion of salt water habitat in grid cell $i$ that is sampled, and $\mu_i$ is defined in eqn.
+(\ref{eq:mu}).
 We modeled the grid cells that were chosen for sampling using eqns. (\ref{eq:R}-\ref{eq:nu}).
@@
 
-To compare and contrast models, we calculated marginal AIC and computed mean square prediction error.
+To compare and contrast models, we calculated marginal AIC (having marginalized over spatial random effects; see below) and mean square prediction error.
 Prediction error was calculated using a Monte Carlo cross validation procedure.
@@
 
-We used maximum (marginal) likelihood to conduct statistical inference.
-For increased numerical tractability, we used the result that the marginal distribution of a Poisson-binomial mixture is itself a Poisson distribution to formulate the likelihood.
+We used maximum marginal likelihood to conduct statistical inference, where the random effects ($\bfeta$ and $\bfdelta$) are integrated out of the likelihood.
+For increased numerical tractability, we also used the result that the marginal distribution of a Poisson-binomial mixture is itself a Poisson distribution to formulate the likelihood.
 In particular, we replaced the binomial observation model (eqn.
@@
 \label{eq:obs_pois}
-   [Y_i] & = & \text{Poisson}(\exp(\mu_i) A_i p_i).
+   [Y_i] & = & \text{Poisson}(\exp(\mu_i) A_i a_i).
 \end{eqnarray}
@@
 The TMB software uses a Laplace approximation to integrate out random effects ($\bfeta$ and $\bfdelta$), and a bias correction algorithm \citep{TierneyEtAl1989,ThorsonKristensen2016} to obtain abundance estimates and standard errors that properly account for nonlinear transformations of random effects.
-This approach resulted in a facile implementation and speedy computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.
+This approach resulted in a facile implementation and fast computing times, allowing us to conduct simulation and model testing with greater efficiency than would have been possible with Bayesian simulation.
 Further detail on statistical methods are provided in Appendix S1; requisite R and TMB code will be published to a publicly accessible repository upon acceptance, and is also available at \url{https://github.com/NMML/pref_sampling/}.
@@
 
-Estimates of cumulative animal abundance across simulated landscapes were relatively unbiased (-4\% for the canonical SDM and 5\% for the joint model) in the scenario where survey locations were selected randomly (i.e.
+Estimates of cumulative animal abundance across simulated landscapes were relatively unbiased (-4\% for the canonical SMPD and 5\% for the joint model) in the scenario where survey locations were selected randomly (i.e.
 independent of abundance; fig.
@@
 Under this scenario, abundance estimates had higher variance when $b$ was estimated.
-Under moderate PS ($b=1$), estimation of the PS parameter $b$ remained unbiased (-1\%), while the canonical SDM model ignoring preferential sampling had a mean bias of 48\%.
-Under pathological PS ($b=3$), the canonical SDM model was extremely biased (mean bias of 159\%) while the joint model had a relatively small positive mean bias of 6\% (fig.
+Under moderate PS ($b=1$), estimation of the PS parameter $b$ continued to result in unbiased estimates of abundance (-1\%), while the canonical SMPD model ignoring preferential sampling had a mean bias of 48\%.
+Under pathological PS ($b=3$), the canonical SMPD model was extremely biased (mean bias of 159\%) while the joint model had a relatively small positive mean bias of 6\% (fig.
 \ref{fig:sim_bias}).
@@
 This view is further supported by examination of cross-validated mean square prediction error, which suggests better predictive performance of models without a preferential sampling effect (Table \ref{tab:aic}).
+Given these findings, we suggest that researchers employ multiple metrics to diagnose the existence of PS, with more weight given to cross validatory predictive performance than AIC (see \textit{Discussion} for further suggestions on diagnosing PS).
 
@@
 In this study, we showed that preferential sampling can have a profound impact on the quality of estimates (e.g.
-animal abundance) when sampling is non-randomized.
+population abundance) when sampling is non-randomized.
 In simulations, naive estimators were increasingly positively biased as PS increased.
@@
 
-Bias attributed to PS may seem counterintuitive, especially given the maxim in survey sampling to allocate more effort to strata for which animal density is high (at least when the goal is to maximize precision of a population estimate).
+Bias attributed to PS may seem counterintuitive, especially given the maxim in survey sampling to allocate more effort to strata for which population density is high (at least when the goal is to maximize precision of population estimates).
 For instance, in large-scale line transect surveys under stratified sampling, the optimal amount of effort that should be allocated to stratum $s$ is $A_s D_s^{0.5}$, where $A_s$ is the area of $s$ and $D_s$ is the anticipated density \citep[][eqn 7.7]{BucklandEtAl2001}.
 Thus, there are theoretical reasons to sample more in high density areas than in low-density areas.
-One solution is to sample more in locations where covariates suggest abundance will be higher; in this case, sampling will not be preferential as long as the covariate used to select sampling locations is also used as an explanatory covariate in the SDM (fig.
+One solution is to sample more in locations where covariates suggest abundance will be higher; in this case, sampling will not be preferential as long as the covariate used to select sampling locations is also used as an explanatory covariate in the SMPD (fig.
 \ref{fig:PS_example}).
@@
 
-When applied to bearded seal count data, models with a preferential sampling effect produced estimates that were considerably lower abundance estimates than models without a PS effect (Table \ref{tab:aic}).
+When applied to bearded seal count data, models with a preferential sampling effect produced considerably lower abundance estimates than models without a PS effect (Table \ref{tab:aic}).
 However, such estimates were biologically unreasonable despite being favored by AIC (Table \ref{tab:aic}).
@@
 Finally, values less than zero indicate that survey effort is concentrated in places where abundance is anomalously low, which seems unlikely to occur in practice.
+Unfortunately, placing informative priors on $b$ was not sufficient to remedy instability (a posthoc model with a Normal(0,1) prior on $b$ still produced implausible estimates).
 
@@
 Such models led to similar inferences to models with a simple, linear effect for PS.
-However, given the potential for PS model instability, we currently suggest limit initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
+However, given the potential for PS model instability, we currently suggest limiting initial consideration to models with a single, estimated $b$ parameter as a composite adjustment to abundance or occupancy.
 Further research into the viability of including spatial variation in the ${\bf B}$ matrix certainly appears warranted.
@@
 We (the authors) have noticed a tendency for analysts to assume that inclusion of spatial covariates or random effects into predictive models will make the underlying sampling design ignorable.
-We have shown in this paper that this is not the case, although our results do suggest that including predictive covariates can indeed decrease bias from preferential sampling.
+Although this may be reasonable for some goals (e.g.
+describing relative influence of different environmental and habitat predictors), we have shown that estimates of absolute quantities (abundance, density, or occupancy) may be biased under preferential sampling.
 We have also shown that it is possible to further diagnose and adjust for preferential sampling by jointly modeling dependence between the data collection mechanism and the process of interest (e.g.
@@
 Where possible, we suggest that survey planners incorporate design-based elements (e.g.
-random or systematic sampling) into their survey designs to reduce the need for model-based triage.
+random or systematic sampling) into their survey designs to increase robustness and reduce the need for model-based triage.
 
@@
 \includegraphics[width=170mm]{Pref_samp_sim_maps.pdf}
-\caption{An example of a single simulation replicate examining estimates of abundance from a naive species distribution model under preferential sampling.
+\caption{An example of a single simulation replicate examining estimates of abundance from a naive SMPD under preferential sampling.
 First, true abundance (C) is generated as a function of a spatially autocorrelated covariate (A) and a spatially autocorrelated random effect (B).
 Second, counts are generated for three different types of surveys, including a simple random sample ($b=0$; D) and surveys with moderate ($b=1$; E) or pathological ($b=3$; F) levels of preferential sampling.
-Finally, spatially explicit estimates of abundance are generated using a traditional SDM (with $b$ set to 0.0) to each of the count datasets (G-I).
+Finally, spatially explicit estimates of abundance are generated using a naive SMPD (with $b$ set to 0.0) to each of the count datasets (G-I).
 In this particular simulation replicate, cumulative abundance was overestimated by 9\% when $b=0$, overestimated by 79\% when $b=1$, and overestimated by 308\% when $b=3$.
@@
 \caption{Aerial survey counts and estimated apparent abundance of bearded seals in the eastern Bering Sea, April 10-16, 2012.
-Counts and estimates are shown relative to a survey grid that extends south from the Bering Strait and borders Alaskato the east.
+Counts and estimates are shown relative to a survey grid that extends south from the Bering Strait and borders Alaska to the east.
 In (A), tan shading denotes land, unsurveyed grid cells appear in dark gray, and counts appear in a white-blue spectrum.

### 2017-04-24 TMB_version/examples/simple_example_v6.R

@@ -255,9 +255,16 @@ Which.est = which(Results[,"EM"]>1)
 Results[,"Bias_b"]=NA
 Results[Which.est,"Bias_b"]=(Results[Which.est,"b_est"]-Results[Which.est,"b"])
 
-#For figures, limit to results for which abs(bias) < 20 for b
+#
+pdf("B_est.pdf")
+  Which_plot = which(Results[,"EM"]>1)
+  plot(Results[Which_plot,"b_est"],xlab="Simulation",ylab=expression(hat(b)))
+dev.off()
+
+
+#For figures, limit to results for which |b estimate| < 10 
 Results[,"OnBoundary"]=rep(0,nrow(Results))
-Results[which(abs(Results[,"Bias_b"])>20),"OnBoundary"]=1
+Results[which(abs(Results[,"b_est"])>10),"OnBoundary"]=1
 
 Results_plot=Results[Results[,"OnBoundary"]==0,]
 Results_plot[,"Est.model"]=rep("Independent",nrow(Results_plot))
@@ -294,3 +301,47 @@ bias.plot
 pdf("bias.pdf")
 bias.plot
 dev.off()
+
+#bias of b parameter
+BiasB.df = Bias.df[which(!is.na(Bias.df[,"Bias_b"])),]
+bias.plot = ggplot(BiasB.df,aes(factor(b),Bias_b))+geom_boxplot() #+facet_grid(~b) #,scales="free")
+bias.plot=bias.plot + theme(text=element_text(size=20)) #+ coord_cartesian(ylim = c(-1., 6.0))
+bias.plot=bias.plot + theme(axis.text.y=element_text(size=14))
+#bias.plot=bias.plot + geom_point(data=DF.trunc.1,aes(x=Est.mod,y=Bias),shape=2)
+#bias.plot=bias.plot + geom_point(data=DF.trunc.5,aes(x=Est.mod,y=Bias),shape=2)
+bias.plot=bias.plot + labs(x = "True b parameter", y="Absolute bias")
+bias.plot
+pdf("biasB.pdf")
+bias.plot
+dev.off()
+
+#produce plot of bias of species-habitat relationship parameters
+bias.plot = ggplot(Bias.df,aes(factor(Est.model),BiasBeta11))+geom_boxplot()+facet_grid(~b) #,scales="free")
+bias.plot=bias.plot + theme(text=element_text(size=20)) #+ coord_cartesian(ylim = c(-1., 6.0))
[…]

### 2017-04-24 pref_sampling/inst/plot_sim_example.R

@@ -1,56 +1,0 @@
-### plot preferential sampling plots using data from "run_sim_example.R"
-library(PrefSampling)
-
-#requires that the dataset sim_plot_data.RData be in the current working directory
-# this file can be created with "run_sim_example.R"
-load('./sim_plot_data.Rdata')
-
-N.true=matrix(Sim$N.true,ncol=1)
-N.1=matrix(Sim$N.1,ncol=1)
-N.2=matrix(Sim$N.2,ncol=1)
-N.3=matrix(Sim$N.3,ncol=1)
-C.1=matrix(Sim$C.1,ncol=1)
-C.2=matrix(Sim$C.2,ncol=1)
-C.3=matrix(Sim$C.3,ncol=1)
-Delta=matrix(Sim$Delta,ncol=1)
-Cov=matrix(Sim$Cov,ncol=1)
-
-greenPalette <- colorRampPalette(brewer.pal(9, "Greens"))
-bluePalette <- colorRampPalette(brewer.pal(9, "Blues"))
-YlOrBr<- colorRampPalette(brewer.pal(9, "YlOrBr"))
-txt.size=8
-
-
-#turn sim data into SpatialPolygonsDataFrame for use with plot_N_map
-Grid.topo=GridTopology(c(0,0),c(1,1),c(25,25))
-Grid.SpG=SpatialGrid(Grid.topo)
-Grid.SpP=as(as(Grid.SpG,"SpatialPixels"),"SpatialPolygons")
-laea_180_proj <- paste("+proj=laea +lat_0=90 +lon_0=180 +x_0=0 +y_0=0",
-                       "+datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
-proj4string(Grid.SpP)=CRS(laea_180_proj)   
-Grid=SpatialPolygonsDataFrame(Grid.SpP,data=data.frame(Cov=Cov),match.ID=FALSE)
-
-
-max.N=100
-Ntrue.plot=plot_N_map(1,N.true,Grid=Grid)+ggtitle(expression(paste("C. True abundance (",N[i],")")))+theme(plot.title = element_text(hjust = 0),legend.title = element_blank(),text=element_text(size=txt.size),plot.margin=unit(c(0,0,0,0),"lines"))+scale_fill_gradientn(colours=YlOrBr(100),limits=c(0,max.N))   
[…]

### 2017-04-24 pref_sampling/inst/run_bearded_models.R

@@ -39,7 +39,7 @@ set.seed(12345)
 s_i = Mapping
 y_s = ifelse(1:n_cells %in% s_i, 1, 0)
 x_s_z = cbind(Grid.new@data[,"ice_conc"],Grid.new@data[,"ice_conc"]^2,Grid.new@data[,"dist_edge"],Grid.new@data[,"depth"],Grid.new@data[,"depth"]^2,Grid.new@data[,"dist_land"])  
-x_s_y = cbind(Grid.new@data[,"dist_land"])  
+x_s_y = cbind(Grid.new@data[,"dist_mainland"])  
 x_s_b = matrix(1,nrow=n_cells,ncol=3)  #design matrix for preferential sampling effect
 Centroids=gCentroid(Grid.new,byid=TRUE)
 loc_s=cbind(Centroids$x,Centroids$y)/25067.53 #standardize by making distance between neighboring cells = 1.0
@@ -48,7 +48,6 @@ loc_s[,2]=loc_s[,2]/mean(loc_s[,2])
 colnames(loc_s)=c("x","y")
 rownames(loc_s)=c(1:n_cells)
 x_s_b[,2:3] = loc_s
-cov_y = FALSE  # include fixed effects for y ?
 c_i = Count.data[,"Count"]   #changed to binom for v5, 11/9
 Count=rep(NA,prod(n_cells))
 Count[s_i]=c_i
@@ -57,10 +56,10 @@ mesh = inla.mesh.create( loc_s )
 spde <- (inla.spde2.matern(mesh, alpha=1)$param.inla)[c("M0","M1","M2")]
 
 
-Options_vec = c( 'Prior'=switch(Spatial_model,"ICAR"=1,"SPDE_GMRF"=0), 'Alpha'=Alpha, 'IncludeDelta'=1, 'IncludeEta'=1, 'OutputSE'=1)
+Options_vec = c( 'Prior'=switch(Spatial_model,"ICAR"=1,"SPDE_GMRF"=0), 'Alpha'=Alpha, 'IncludeDelta'=1, 'IncludeEta'=1, 'OutputSE'=1,'bPrior'=0)
 
 
-for(icov in 1:3){
+for(icov in 1:2){
   for(ib in 1:3){
     # Data
[…]

### 2017-04-24 pref_sampling/inst/run_bearded_models_cross_val.R

@@ -1,0 +1,175 @@
+### run simple preferential sampling models on bearded seal data
+library(PrefSampling)
+data(Data_for_ST_plot) #includes spatial objects associated w landscape
+data(Bearded_effort)
+data(AlaskaBeringData)  #from bass package on github/nmml
+Effort=Bearded_effort
+
+#create single Grid SpatialPolygonsDataFrame, averaging over covariate values
+Grid.new=Data$Grid$y2012[[1]]
+Days = c(6:12) #Apr 10-16
+Grid.new@data=0*Grid.new@data
+for(iday in Days)Grid.new@data=Grid.new@data+Data$Grid$y2012[[iday]]@data
+Grid.new@data=Grid.new@data/length(Days)
+
+#limit effort data to April 10-16
+Which.incl = which(Effort$Mapping[,2]<8)
+Mapping=Effort$Mapping[Which.incl,1]
+Prop_sampled=Effort$Area.trans[Which.incl]
+Prop_sampled[which(is.na(Prop_sampled))]=1.0
+Count.data=Effort$Count.data[Which.incl,]
+
+# Settings
+n_test_obs = 40 # sample size for repeated subsample cross-validation training set
+n_folds = 100 # number of training sets to test
+n_cells=length(Grid.new)
+n_samp = length(Mapping)
+#SpatialScale = sqrt(n_cells)/5  # Range ~ 2*Scale
+#SD_eta = SD_x = SD_delta = 1
+#beta0 = 2
+Use_REML = FALSE   # 
+#Spatial_sim_model = c("GP_gaussian", "ICAR")[1]
+Spatial_model = c("SPDE_GMRF", "ICAR")[1]
+Alpha = 2  # Smoothness for GMRF, 1 or 2 (1 is faster)
+
+# Model configurations
+EM_set = c("fix_b","est_b")
+
+set.seed(12345)
+
+# massage into paper notation
+s_i = Mapping
+y_s = ifelse(1:n_cells %in% s_i, 1, 0)
+x_s_z = cbind(Grid.new@data[,"ice_conc"],Grid.new@data[,"ice_conc"]^2,Grid.new@data[,"dist_edge"],Grid.new@data[,"depth"],Grid.new@data[,"depth"]^2,Grid.new@data[,"dist_land"])  
+x_s_y = cbind(Grid.new@data[,"dist_mainland"])  
[…]

### 2017-04-24 pref_sampling/inst/run_sim_example.R

@@ -19,12 +19,12 @@ Alpha = 1  # Smoothness for GMRF, 1 or 2 (1 is faster)
 RandomSeed = ceiling(runif(1,min=1,max=1e6))
 
 # Configurations
-b_set = c(0,1,5)            # Impact of delta on sampling intensity
+b_set = c(0,1,3)            # Impact of delta on sampling intensity
 EM_set = c("fix_b","est_b")
 
 # Configurations
 
-set.seed(12345)
+set.seed(2222222)
 
 Counts=Est=vector("list",3)
 
@@ -127,8 +127,8 @@ for(SimI in 1:length(b_set)){
     Obj$fn( Obj$par )
     
     # Run
-    Lower = -Inf
-    Upper = Inf
+    Lower = -50
+    Upper = 50
     Opt = nlminb( start=Obj$par, objective=Obj$fn, gradient=Obj$gr, lower=Lower, upper=Upper, control=list(trace=1, maxit=1000))         #
     Opt[["diagnostics"]] = data.frame( "Param"=names(Obj$par), "Lower"=-Inf, "Est"=Opt$par, "Upper"=Inf, "gradient"=Obj$gr(Opt$par) )
     Report = Obj$report()
@@ -161,7 +161,7 @@ for(SimI in 1:length(b_set)){
     # SD
     #Report = Obj$report()
     SD=sdreport(Obj,bias.correct=TRUE)
-    Est[[SimI]]=SD$unbiased$value[5:629]
+    Est[[SimI]]=SD$unbiased$value[6:630]
     #if( all(c("etainput_s","deltainput_s")%in%names(Map)) ){
     #  Est[[EstI]]=Report$Z_s
     #}else{

### 2017-04-24 pref_sampling/src/PrefSampling.cpp

@@ -110,7 +110,7 @@ Type objective_function<Type>::operator() ()
   // Slot 2: Include hyperdistribution for delta
   // Slot 3: Include hyperdistribution for eta
   // Slot 4: Output Z_s in ADREPORT (0=no, 1=yes); needed for unbiased map estimates using bias.correct
-  
+  // Slot 5: Indicator for whether to put a prior on preferential sampling parameters
   // Data
   DATA_VECTOR( c_i );       	// Response (count) for each observation i
   DATA_VECTOR( P_i );      // Proportion of survey unit that is surveyed for each observation i
@@ -228,6 +228,11 @@ Type objective_function<Type>::operator() ()
   // Probability of sampling locations
   for(int s=0; s<n_s; s++){
     if( !isNA(y_s(s)) ) jnll_comp(1) -= dbern( y_s(s), R_s(s), true );
+  }
+  if(Options_vec(5)==1){
+    for(int ib=0;ib<n_b;ib++){
+      jnll_comp(1)+=0.5*beta_b(ib)*beta_b(ib);  // N(0,1) prior
+    }
   }
 
   // Total objective
